### 2025-04-04 draft/draft.tex
progress

%% ****** Start of file apstemplate.tex ****** %
%%
%%
%%   This file is part of the APS files in the REVTeX 4.2 distribution.
%%   Version 4.2a of REVTeX, January, 2015
%%
%%
%%   Copyright (c) 2015 The American Physical Society.
%%
%%   See the REVTeX 4 README file for restrictions and more information.
%%
%
% This is a template for producing manuscripts for use with REVTEX 4.2
% Copy this file to another name and then work on that file.
% That way, you always have this original template file to use.
%
% Group addresses by affiliation; use superscriptaddress for long
% author lists, or if there are many overlapping affiliations.
% For Phys. Rev. appearance, change preprint to twocolumn.
% Choose pra, prb, prc, prd, pre, prl, prstab, prstper, or rmp for journal
%  Add 'draft' option to mark overfull boxes with black boxes
%  Add 'showkeys' option to make keywords appear
\documentclass[aps,prl,twocolumn,groupedaddress]{revtex4-2}

% You should use BibTeX and apsrev.bst for references
% Choosing a journal automatically selects the correct APS
% BibTeX style file (bst file), so only uncomment the line
% below if necessary.
%\bibliographystyle{apsrev4-2}

\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{color}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{mathrsfs}
\usepackage{wasysym}
\usepackage{tikz-cd}
\usepackage{adjustbox}
\usepackage{booktabs}
\usepackage{orcidlink}

% https://tex.stackexchange.com/questions/269553/counter-too-large-error-with-item
\usepackage{alphalph}
%\renewcommand\thesubsectiondis{\AlphAlph{\value{subsection}}.}% for the headings in the text
\renewcommand%\thesubsection{\mbox{\thesection-\AlphAlph{\value{subsection}}}}% for the ToC, for example   \renewcommand
\thesubsection{\mbox{\AlphAlph{\value{subsection}}}}% for the ToC, for example

% so sections, subsections, etc. become numerated.
\setcounter{secnumdepth}{3}

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\avrg}[1]{\left\langle #1 \right\rangle}
\newcommand{\nelta}{\bar{\delta}}
\newcommand{\krn}{\mathrm{krn}}
\newcommand{\rng}{\mathrm{rng}}
\newcommand{\rnk}{\mathrm{rnk}}
\newcommand{\nll}{\mathrm{nll}}
\newcommand{\grad}{\mathrm{grad}}
\newcommand{\curl}{\mathrm{curl}}
\newcommand{\dive}{\mathrm{div}}
\newcommand{\bra}[1]{\left\langle #1\right|}
\newcommand{\ket}[1]{\left| #1 \right\rangle}
\newcommand{\sbra}[1]{\langle #1|}
\newcommand{\sket}[1]{| #1 \rangle}
\newcommand{\bek}[3]{\left\langle #1 \right| #2 \left| #3 \right\rangle}
\newcommand{\sbek}[3]{\langle #1 | #2 | #3 \rangle}
\newcommand{\braket}[2]{\left\langle #1 \middle| #2 \right\rangle}
\newcommand{\ketbra}[2]{\left| #1 \middle\rangle \middle\langle #2  \right|}
\newcommand{\sbraket}[2]{\langle #1 | #2 \rangle}
\newcommand{\sketbra}[2]{| #1 \rangle  \langle #2 |}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\snorm}[1]{\lVert#1\rVert}
\newcommand{\bvec}[1]{\boldsymbol{\mathsf{#1}}}
\newcommand{\bcov}[1]{\boldsymbol{#1}}
\newcommand{\bdua}[1]{\boldsymbol{\check{#1}}}
\newcommand{\bdov}[1]{\breve{#1}}
\newcommand{\bten}[1]{\boldsymbol{\mathfrak{#1}}}
\newcommand{\forany}{\tilde{\forall}}
\newcommand{\qed}{$\overset{\circ}{.}\;$}
\newcommand{\bigo}{\mathcal{O}}
\newcommand\scalemath[2]{\scalebox{#1}{\mbox{\ensuremath{\displaystyle #2}}}}

\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\KK}{\mathbb{K}}
\newcommand{\FF}{\mathbb{F}}

\newcommand{\smallo}{o}

\begin{document}

% Use the \preprint command to place your local institutional report
% number in the upper righthand corner of the title page in preprint mode.
% Multiple \preprint commands are allowed.
% Use the 'preprintnumbers' class option to override journal defaults
% to display numbers if necessary
%\preprint{}

%Title of paper
\title{Analysis of the inference of ratings and rankings on Higher Order Networks with complex topologies}

% repeat the \author .. \affiliation  etc. as needed
% \email, \thanks, \homepage, \altaffiliation all apply to the current
% author. Explanatory text should go in the []'s, actual e-mail
% address or url should go in the {}'s for \email and \homepage.
% Please use the appropriate macro foreach each type of information

\author{Juan I. Perotti\,\orcidlink{https://orcid.org/0000-0001-7424-9552}}
\email[]{juan.perotti@unc.edu.ar}
\affiliation{Instituto de F\'isica Enrique Gaviola (IFEG-CONICET),\\
Facultad de Matem\'atica, Astronom\'ia, F\'isica y Computaci\'on,\\
Universidad Nacional de C\'ordoba, Ciudad Universitaria, 5000 C\'ordoba, Argentina}

%Collaboration name if desired (requires use of superscriptaddress
%option in \documentclass). \noaffiliation is required (may also be
%used with the \author command).
%\collaboration can be followed by \email, \homepage, \thanks as well.
%\collaboration{}
%\noaffiliation

\date{\today}

\begin{abstract}
The inference of rankings plays a central role in the theory of social choice, which seeks to establish preferences from collectively generated data, such as pairwise comparisons.
Examples include political elections, ranking athletes based on competition results, ordering web pages in search engines using hyperlink networks, and generating recommendations in online stores based on user behavior. 
Various methods have been developed to infer rankings from incomplete or conflicting data. One such method, HodgeRank, introduced by Jiang et al.~\cite{jiang2011statistical}, utilizes Hodge decomposition of cochains in Higher Order Networks to disentangle gradient and cyclical components contributing to rating scores, enabling a parsimonious inference of ratings and rankings for lists of items.
This paper presents a systematic study of HodgeRank's performance under the influence of quenched disorder and across networks with complex topologies generated by four different network models. The results reveal a transition from a regime of perfect trieval of true rankings to one of imperfect trieval as the strength of the quenched disorder increases. A range of observables are analyzed, and their scaling behavior with respect to the network model parameters is characterized. This work advances the understanding of social choice theory and the inference of ratings and rankings within complex network structures.
\end{abstract}

% insert suggested keywords - APS authors don't need to do this
%\keywords{}

%\maketitle must follow title, authors, abstract, and keywords
\maketitle

\section{Introduction}

The rating and ranking candidates, agents, items or options based on collectively expressed preferences is deeply ingrained modern human civilization, making the study of inference methods for ratings and rankings under diverse conditions highly significant.
Ratings and rankings are often derived from pairwise comparisons, which naturally form a network structure. 
Understanding how the topology of these networks influences the accuracy of rating and ranking inference is of critical importance.

The modern theory of ratings and rankings has its origins in the field of social choice, with early foundations laid in the late 13th century by Ramon Llull in his {\em Ars Electionis}~\cite{colomer2013ramon}. 
Llull's work introduced the concept of pairwise comparisons, the majority principle, and fairness in voting systems. 
Unfortunately, his contributions were largely forgotten over time. 
A mathematical revival of these ideas emerged in the late 18th century with the works of Borda and Condorcet~\cite{borda1781memoire,condorcet1785essai}, who laid the groundwork for modern voting and ranking theory. 
Later, in 1929, Zermelo proposed a model for ranking inference using ratings as latent variables within a probabilistic framework~\cite{zermelo1929berechnung}, but his work was also overlooked~\cite{newman2023efficient}. 
In the 1950s, Bradley and Terry reintroduced Zermelo’s model in a broader context~\cite{bradley1952rank}. 
In the 1960s, a model based on similar principles was introduced and promoted by \'El\H{o}, eventually becoming the default rating system used by the World Chess Federation (FIDE)~\cite{elo1978rating}.
More recently, the advent of the World Wide Web and search engines highlighted the significance of network topology in ranking systems, exemplified by the success of PageRank, the original algorithm behind Google and introduced Brin and Page at the end of the 90's~\cite{brin1998anatomy}. 
The study of rating and ranking systems remains an active area of research~\cite{jiang2011statistical,csato2017ranking,debacco2018physical,newman2022ranking,jerdee2023luck}, as collective decision-making plays a vital role in modern civilization, particularly in the information age.

In 1736, Euler solved the Seven Bridges of K\"onigsberg problem, establishing the foundations of graph theory and foreshadowing the development of topology. 
Graphs, or networks, can be viewed as a specific type of topological space. 
Within the theory of Higher Order Networks~\cite{bianconi2021higher,battiston2021physics,majhi2022dynamics,bick2023what,ferraz2024contagion,muolo2024global,bianconi2024nature,bianconi2024theory}, hypergraphs extend the concept of graphs by incorporating hyperlinks, that may be used to represent many-body interactions or multiary-relations.
However, unlike graphs, hypergraphs are not necessarily topological spaces, as they may not satisfy the condition of topological closure. 
For example, a hypergraph may include a hyperlink $\{i,j,k\}$ without containing the link $\{i,j\}$. 
When hypergraphs satisfy the closure condition, they form a special class of topological spaces known as abstract simplicial complexes, or simply, simplicial complexes.

Although terminology and definitions may vary by context, many concepts from differential geometry on smooth manifolds can be adapted to the discrete or combinatorial setting of simplicial complexes~\cite{bobenko2008discrete,grady2010discrete,aygun2011spectral,tuncer2015spectral,rovelli2019natural,bianconi2021higher}.
Namely, discrete or combinatorial analogous of forms, gradients, divergences and Laplacians can be defined over simplicial complexes.
This is the reason for which simplicial complexes are often preferred over general hypergraphs.
In particular, and thanks to the property of topological closure, Hodge theory can be applied to decompose cochains over simplicial complexes into exact, coexact and harmonic components, much like the analogous decomposition of tensor fields~\cite{grady2010discrete,lim2020hodge}.
This is the approach followed by Jian et al.~\cite{jiang2011statistical} who developed HodgeRank, a method for the inference of ratings that leverages Hodge theory over simplicial complexes to disentangle different contributions to the inferred ratings.
The present article takes advantage of HodgeRank to study the inference of ratings and rankings in complex topologies~\cite{newman2018networks,perotti2009emergent}.
Section~\ref{sec:theory} provides an introduction to the theory of rating and ranking inference, and focuses on HodgeRank. Section~\ref{sec:results} presents the results of numerical experiments using HodgeRank applied to networks with complex topologies. 
Section~\ref{sec:discussion} discuss the results.
Section~\ref{sec:conclusions} summarizes the findings and outlines directions for future research.

\section{
Theory
\label{sec:theory}
}

Several methods exist for inferring ratings and rankings from pairwise comparison data~\cite{jiang2011statistical}. This section focuses on a particular approach inspired by the foundational works of Zermelo~\cite{zermelo1929berechnung} and Bradley-Terry~\cite{bradley1952rank}, to illustrate how pairwise comparisons can be associated with probabilities that, in turn, relate to rating scores. These concepts are then used to introduce the HodgeRank method, followed by a discussion of the numerical challenges that arise within this framework.

Consider a set of $N=n+1$ items indexed by $i \in \{0, 1, \dots, n\}$. 
These items or agents could represent movies subject to pairwise comparison or even chess players competing in chess tournaments~\cite{perotti2013innovation}. 
For any two items $i$ and $j$, there exists a (possibly empty) sequence $s_{ij} = (s_{ij1}, s_{ij2}, \dots, s_{ijt}, \dots)$ of comparison results, where $s_{ijt} = -s_{jit}$ takes the value $1$, $0$, or $-1$ depending on whether option $i$ wins, draws, or loses the $t$-th comparison against option $j$, respectively. 
For example, these results could represent outcomes of chess games between players, or voting preferences over movies. 
Importantly, the set of non-empty sequences, or {\em pairings}, defines the links of an undirected network with $N$ nodes and no self-links, described by an adjacency matrix $a$ with entries $a_{ij} = a_{ji} \in \{0, 1\}$ for $i,j \in \{0, 1, \dots, n\}$. 
The degree, or the number of links adjacent to node $i$, is $k_i = \sum_j a_{ij} = \sum_j a_{ji}$, and the average degree is $\bar{k} = N^{-1} \sum_i k_i = 2M/N$, where $M = \frac{1}{2} \sum_{ij} a_{ij}$ is the total number of links.

Assuming that the results within a sequence are statistically independent and are identically distributed, the probability for a particular sequence to occur is 
\begin{eqnarray}
\label{eq:v_and_q}
P(s_{ij1},
...,s_{ijt},...)
&=&
P(s_{ij1})
...P(s_{ijt})...
\\
&=&
p_{ij}^{x_{ij}}q_{ij}^{y_{ij}}r_{ij}^{z_{ij}}
\nonumber
\\
&=&
v_{ij}^{x_{ij}}(1-v_{ij})^{z_{ij}}
q_{ij}^{y_{ij}}(1-q_{ij})^{x_{ij}+z_{ij}}
\nonumber
\end{eqnarray}
where $p_{ij}:=P(S_{ijt}=1)$, $q_{ij}:=P(S_{ijt}=0)$ and $r_{ij}:=P(S_{ijt}=-1)$, and $x_{ij}=\sum_t \delta_{s_{ijt},1}$, $y_{ij}=\sum_t \delta_{s_{ijt},0}$ and $z_{ij}=\sum_t \delta_{s_{ijt},-1}$ are the number of wins, draws and losses of $i$ over $j$ for all $t$, respectively.
Moreover, $v_{ij}:=P(S_{ijt}=1|S_{ijt}\neq 0)$ and $1-v_{ij}:=P(S_{ijt}=-1|S_{ijt}\neq 0)$ since $p_{ij}=v_{ij}(1-q_{ij})$ and $r_{ij}=(1-v_{ij})(1-q_{ij})$.
There are different ways to fit the probabilities of Eq.~\ref{eq:v_and_q} to given data.
For instance, in the maximum a posteriori bayesian approach of the multinomial distribution with a uniform prior, the expected value of the probabilities are $p_{ij}=(x_{ij}+1)/(x_{ij}+y_{ij}+z_{ij}+3)$, $q_{ij}=(y_{ij}+1)/(x_{ij}+y_{ij}+z_{ij}+3)$ and $r_{ij}=(z_{ij}+1)/(x_{ij}+y_{ij}+z_{ij}+3)$ and, therefore $v_{ij}=p_{ij}/(1-q_{ij})=(x_{ij}+1)/(x_{ij}+z_{ij}+2)$.
In this way, pairs of items with few comparisons can be handled appropriately.
In particular, if the kind of comparisons under consideration admit no draws, then $q_{ij}=0$ for all $ij$.

\subsection{
A rating method from least-squares
\label{sec:ratings_and_rankings_lsqr}
}

Consider the parameterization of the probabilities $v_{ij}$ in Eq.~\ref{eq:v_and_q} using the logistic function $v = (1 + e^f)^{-1}$. 
This defines a monotonically decreasing bijection between the probabilities $v_{ij} \in (0, 1)$ and the corresponding values $f_{ij} = \ln(v_{ij}^{-1} - 1) \in (-\infty, \infty)$, and motivates the following question.
{\em Is there a vector $w \in \mathbb{R}^N$ with components $w_i$ such that Eq.~\ref{eq:motivating_exact} is satisfied?}
\begin{equation}
\label{eq:motivating_exact}
f_{ij}=w_j-w_i
\end{equation}
If such a vector exists, it could serve as a rating system, where $v_{ij} > 1/2$ if and only if $f_{ij} < 0$, or equivalently, if and only if $w_j < w_i$. 
Additionally, this vector could be used to derive a ranking system, where a permutation $r_0, \dots, r_n$ of $0, 1, \dots, n$ satisfying $w_{r_0} \leq w_{r_1} \leq \dots \leq w_{r_n}$, yields a ranking $r_0, r_1, \dots, r_n$ of the items.
However, such a vector $w$ does not always exist. 
For example, a dataset might exhibit cyclic behavior, where item $i$ defeats $j$, $j$ defeats $k$, and $k$ defeats $i$, leading to inconsistencies in the ranking. 
Cyclical inconsistencies are common in practice.
Fortunatelly, a weakened version of Eq.~\ref{eq:motivating_exact} can still be used to infer useful ratings and rankings. 
Specifically, the exact condition of Eq.~\ref{eq:motivating_exact} can be replaced by the least-squares problem in Eq.~\ref{eq:motivating_lsqr}~\cite{csato2015graph}.
\begin{equation}
\label{eq:motivating_lsqr}
w = \argmin_{w'}
\sum_{ij} a_{ij}\,|f_{ij} - (w'_j-w'_i)|^2
\end{equation}
The idea here is that, although Eq.~\ref{eq:motivating_exact} cannot always be satisfied, the solution to Eq.~\ref{eq:motivating_lsqr} provides the vector $w$ that best approximates the intended behavior in a least-squares sense.

\subsection{HodgeRank}

In this section the HodgeRank method is reviewed~\cite{jiang2011statistical}.
For further reading, some practical applications of HodgeRank are demonstrated by Johnson et al.~\cite{johnson2013discrete} and some numerical challenges emerging from its appliation to complex networks are studied by Hirani et al.~\cite{hirani2015graph}.
Moreover, an elementary review about combinatorial differential topology, geometry and calculus over simplicial complexes is provided in the appendix section, where the formal details are introduced.

An hypergraph is composed of nodes, links and higher order links or hyperlinks such as triangles, tetrahedra and so on.
A simplicial complex is a specific type of hypergraph that satisfies the property of topological closure, making it a topological space. 
This structure allows for the definition of cochains, which are combinatorial versions of differential forms where scalar forms corresponds to 0-cochains, vector forms corresponds to 1-cochains and, more generally, $k$-forms corresponds to $k$-cochains.
In practical terms, a 0-cochain is a vector of scalars $w_i$ associated with nodes $i$, while a 1-cochain is a vector of antisymmetric scalars $f_{ij} = -f_{ji}$ associated with links $ij$. 
More generally, a $k$-cochain is a vector of alternating scalars $t_{i_0 \dots i_k}$ associated with hyperlinks known as $k$-simplices. 
Crucially, combinatorial versions of differential operators such as the gradient, the divergence, and the Laplacian, can be defined over simplicial complexes.
This framework allows the least-squares problem in Eq.~\ref{eq:motivating_lsqr} to be reformulated in terms of the 3-clique complex $K = K_0 \cup K_1 \cup K_2$, derived from the network of adjacency matrix $a$. 
Here, the values $f_{ij}$ are recognized as the components of a 1-cochain $f$, where $f_{ij} = -f_{ji}$ because $v_{ji} = 1 - v_{ij}$, while the differences $w_j - w_i$ correspond to the components of the gradient of a 0-cochain $w$.
More specifically, Eq.~\ref{eq:motivating_lsqr} can be rewritten as:
\begin{equation}
\label{eq:hodge_lsqr}
w=\argmin_{w'\in C_0^*}|f-d_0(w')|^2
\end{equation}
where $w \in C_0^*$ is a 0-cochain, and $d_0 = \grad$ is the gradient operator on the vector space $C_0^*$ of 0-cochains.
Thanks to this identification, the least-squares problem in Eq.~\ref{eq:hodge_lsqr} can be reformulated in terms of the Hodge decomposition of $f$.

\begin{figure}
\includegraphics*[scale=.9]{assets/fig1.pdf}
\caption{
\label{fig:1}
(color online).
An elementary example of a simplicial complex with $N=8$ nodes.
Its set of 0-simplices or nodes is $K_0=\{\{0\},\{1\},...,\{7\}\}$, of 1-simplices or links is $K_1=\{\{0,1\},\{1,2\},\{0,2\},\{2,3\},\{3,4\},\{4,5\},\{1,5\},\{4,6\},\{6,7\}\}$ and of 2-simplices or triangles is $K_2=\{\{0,1,2\}\}$.
A local cyclic flow can only exists in the 2-simplex $\{0,1,2\}$ (depicted in green), while non-local cyclic flows can exists in the closed paths $1\to 2\to 3\to 4\to 5\to 1$ and $0\to 2\to 3\to 4\to 5\to 1\to 0$.
}
\end{figure}

In the Hodge decomposition of a general $k$-cochain $f \in C_k^*$, there exist unique $k$-cochains $s, g, h \in C_k^*$ such that $f = s \oplus h \oplus g$, where $s\in \rng\,d_{k}^*$, $h\in \krn\,L_k$ and $g\in \rng\,d_{k-1}$ are called the coexact, the harmonic and the exact components of $f$.
Here, $\rng\,d_{k}^*$, $\krn\,L_k$ and $\rng\,d_{k-1}$ are the ranges and kernel of the $k$-th dual coboundary operator $d_k^*$, the $k$-th Laplacian $L_k$ and the $(k-1)$-th coboundary operator $d_{k-1}$, respectively, and $C_k^*$ is the vector space of $k$-cochains.

To compute the Hodge decomposition of a $k$-chochain $f$, recall that
$\krn\,d_k=\krn\,L_k\oplus\,\rng\,d_{k-1}$
and 
$\krn\,d_{k-1}^*=\rng\,d_k^*\oplus\, \krn\,L_k$.
In consequence
\begin{eqnarray}
\label{eq:lsqr_u}
d_{k}(f) &=& d_{k}(s + h + g) \\
&=& d_{k}(s) \nonumber \\
&=& d_{k}(d_{k}^*(u)) \nonumber \\
&=& L_{k+1}^{\downarrow}(u) \nonumber \\
\alpha_{k+1} f &=& \alpha_{k+1} \alpha_{k+1}^T u \nonumber
\end{eqnarray}
and
\begin{eqnarray}
\label{eq:lsqr_w}
d_{k-1}^*(f) &=& d_{k-1}^*(s + h + g) \\
&=& d_{k-1}^*(g) \nonumber \\
&=& d_{k-1}^*(d_{k-1}(w)) \nonumber \\
&=& L_{k-1}^{\uparrow}(w) \nonumber \\
\alpha_{k}^T f &=& \alpha_{k}^T \alpha_{k} w. \nonumber
\end{eqnarray}
are singular linear equations for some $u \in C_{k+1}^*$ and $w \in C_{k-1}^*$, that can be solved to obtain $s = \alpha_{k+1}^T u$, $g = \alpha_{k} w$, and $h = f - s - g$.
Although the Hodge Laplacians 
$L_{k+1}^{\downarrow}$
and
$L_{k-1}^{\uparrow}$
are symmetric operators, they are also singular and, therefore, Eqs.~\ref{eq:lsqr_u} and \ref{eq:lsqr_w} still pose numerical challenges. 
Fortunately, these are the normal equations of respective least-squares problems. 
Hence, their singularities can be addressed by seeking the solutions $u$ and $w$ that also minimize the squared norms $|u|^2$ and $|w|^2$, respectively.

The Hodge decomposition provides a solution to Eq.~\ref{eq:hodge_lsqr} when $k=1$.
Namely, for $k=1$, any solution $w$ of Eq.~\ref{eq:lsqr_w} is also a solution to Eq.~\ref{eq:hodge_lsqr}, and vice versa. 
This approach to inferring $w$ from the Hodge decomposition of $f$ is what Jiang et al.~\cite{jiang2011statistical} referred to as {\em HodgeRank}. 
Additionally, drawing an analogy from physics in 3D, the coexact, harmonic, and exact components $s$, $h$, and $g$, are referred to as the solenoidal, harmonic, and gradient components of $f$, respectively, providing insightful observations in the context of rating inference.
Namely, the gradient component $g$ represents a non cyclic flow, which aligns perfectly with a rating vector $w$. 
In contrast, the solenoid and harmonic components, $s$ and $h$, correspond to local and non-local cyclic flows, respectively, which are incompatible with a consistent rating system.
See Fig.~\ref{fig:1} for an example.
These cyclic flows were previously discussed in Section~\ref{sec:ratings_and_rankings_lsqr}. 
The norms $|s|$, $|h|$, and $|g|$ help to gauge the extent to which the inferred ratings are affected by local or non-local cyclic flows.

\begin{figure*}
\includegraphics*[scale=.4]{assets/fig2a.pdf}
\includegraphics*[scale=.4]{assets/fig2b.pdf}
\put(-483,140){\bf a)}
\put(-240,140){\bf b)}
\\
\includegraphics*[scale=.4]{assets/fig2c.pdf}
\includegraphics*[scale=.4]{assets/fig2d.pdf}
\put(-483,140){\bf c)}
\put(-240,140){\bf d)}
\caption{
\label{fig:2}
(color online).
The mean of the average of the absolute differences of ratings $\avrg{\bar{\tau}}$ averaged over 5000 network samples and realizations of the disorder $\eta$ as a function of $\sigma$ for different sizes $N$ (left subpanels) and values of the parameter $\theta$ (right subpanels) of the corresponding network model.
Dashed lines represent guides to the eye.
In panel {\bf a)}, 1D-lattices where $\theta=z$ is the number of neighbors per node.
In panel {\bf b)}, Erd\H{o}s-R\'enyi random networks where $\theta=\bar{k}$ is the average degree.
In panel {\bf c)}, Barab\'asi-Albert scale-free networks where $\theta=q$ is the number of new links for each new node of the growing algorithm.
In panel {\bf d)}, Watts-Strogatz small-world networks of average degree $\bar{k}=4$ where $\theta=p$ is randomization probability per link.
In the inset, the value of $\avrg{\bar{\tau}}$ for fixed $\sigma$ is plotted as a function of $p$.
}
\end{figure*}

In practice, some components $w_i$ of the minimum norm solution to Eq.~\ref{eq:lsqr_w} may be negative. 
Since a rating system with non-negative ratings is often desirable, the components can be shifted by a constant term, $w_i \to w_i - \min_j w_j$. 
This shifted rating vector remains a valid solution to Eqs.~\ref{eq:hodge_lsqr} and facilitates the comparison between the inferred $w$ and the true ratings $\hat{w}$ when the later are available.

\section{
Results
\label{sec:results}
}

In the following numerical experiments, HodgeRank is applied to infer the ratings and rankings of items matched by networks of various topologies: 1D-lattices, Erdős-Rényi (ER) random graphs, Barabási-Albert (BA) scale-free networks, and Watts-Strogatz (WS) small-world networks~\cite{newman2018networks}.

For each network of adjacency matrix $a$, a vector of true ratings $\hat{w}$ is generated according to $\hat{w}_i = i$ for $i = 0, 1, \dots, n$. 
With this choice, the expected true ranking is $\hat{r}_i=i$.
In 1D-lattices and WS networks, nodes are enumerated following the natural order in the lattice and in BA networks nodes are enumerated from the oldest to the newest.
Using the true rating vector $\hat{w}$, a corresponding 1-cochain $f$ is defined by
\begin{equation}
\label{eq:f_and_eta}
f_{ij}
=
\hat{w}_j-\hat{w}_i
+
\eta_{ij},
\end{equation}
where the different components of the 1-cochain $\eta$ are independent and identically distributed random variables drawn from a Gaussian distribution with zero mean and standard deviation $\sigma$. 
Equation~\ref{eq:f_and_eta} is similar to Eq.~\ref{eq:motivating_exact}, but includes the contribution of the additive quenched disorder characterized by $\eta$ that models the effect of statistical fluctuations and the cyclic flows captured by the components $s$ and $h$ of $f$.
When $\sigma = 0$, the inferred ratings are expected to be perfectly aligned with the true ratings, but as $\sigma$ increases, discrepancies are likely to emerge.

\begin{figure*}
\includegraphics*[scale=.4]{assets/fig3a.pdf}
\includegraphics*[scale=.4]{assets/fig3b.pdf}
\put(-483,140){\bf a)}
\put(-240,140){\bf b)}
\\
\includegraphics*[scale=.4]{assets/fig3c.pdf}
\includegraphics*[scale=.4]{assets/fig3d.pdf}
\put(-483,140){\bf c)}
\put(-240,140){\bf d)}
\caption{
\label{fig:3}
(color online).
The mean of the average of the absolute differences of rankings $\avrg{\bar{\tau}}$ as a function of $\sigma$ for the same numerical experiments of Fig.~\ref{fig:2}.
The insets zoom-in certain regions of the main plots.
Dashed lines represent guides to the eye.
}
\end{figure*}

To compare the true ratings $\hat{w}_i$ with the inferred ratings $w_i$, as well as the true rankings $\hat{r}_i$ with the inferred rankings $r_i$, the average absolute differences
$$
\tau
=
\frac{1}{N}
%\sum_{j:j\leq i}
\sum_{i}
|\hat{w}_i-w_i|
$$
and
$$
\rho
=
\frac{1}{N}
%\sum_{j:j\leq i}
\sum_{i}
|\hat{r}_i-r_i|
$$
are used, respectively.
These metrics are zero when the inferred ratings or rankings match the true ones, and greater than zero otherwise.
It is important to note that the means $\bar{\tau}(\sigma, a) = \int dw \, P(w | \sigma, a) \, \tau(w)$ and $\bar{\rho}(\sigma, a) = \int dw \, P(w | \sigma, a) \, \rho(w)$, while similar to order parameters, are more accurately described as {\em disorder parameters}, since they become different from zero when disorder kicks in. 
Here, the distribution $P(w | \sigma, a)$ represents the statistics of $w$ obtained from solving Eq.~\ref{eq:hodge_lsqr} while sampling $\eta$ for a given network structure $a$.
In the experiments, the network-averaged quantities, $\langle \bar{\tau} \rangle (\sigma) = \sum_a P(a | N, \theta) \bar{\tau}(\sigma, a)$ and $\langle \bar{\rho} \rangle (\sigma) = \sum_a P(a | N, \theta) \bar{\rho}(\sigma, a)$, are computed, where $N$ and $\theta$ represent relevant parameters of the network models. 
Specifically, $\theta$ denotes the coordination number $z$ for 1D-lattices, the average degree $\bar{k}$ for ER networks, the number of new links $q$ per new node in BA networks, and the fraction $p$ of randomized links for WS networks with average degree $\bar{k}=4$.

Traditional solvers struggle to provide adequate solutions for Eqs.~\ref{eq:lsqr_u} and~\ref{eq:lsqr_w}. 
To address this, the recently introduced iterative solver \verb+MINRES-QLP+, developed by Choi et al.~\cite{choi2011minresqlp}, is employed to find the minimum norm solutions to the singular least-squares problems. 
This solver is available in the \verb+Krylov.jl+ package~\cite{montoison2023krylov} for the Julia programming language~\cite{bezanson2017julia}. 
A Jupyter notebook with the code used for the experiments can be found online~\cite{github_hon_ranking}.
The performance of \verb+MINRES-QLP+ was evaluated on non-periodic 1D-lattices of varying sizes $N$. 
For $\sigma = 0$, the solver demonstrates satisfactory numerical accuracy for network sizes up to approximately $N \approx 10^5$. 
However, to mitigate the impact of numerical errors, subsequent experiments are conducted on networks of considerably smaller sizes.

The following figures contain four panels, each corresponding to a different network model.
In each figure, panel {\bf a)} presents results for non-periodic 1D-lattices, panel {\bf b)} for ER networks, panel {\bf c)} for BA networks and panel {\bf d)} for WS networks.
Furthermore, each panel is divided into two subpanels.
The left subpanel presents results for $\theta$ fixed and varying network sizes $N$, and the right subpanel for fixed $N$ and varying $\theta$.
The results are averaged over 5000 samples of both $a$ (the network structure) and $\eta$ (the disorder).
Where possible, a phenomenologically obtained approximate scaling relation for the studied quantities is indicated.

In Fig.~\ref{fig:2}, the mean of the average absolute differences, $\avrg{\bar{\tau}}$, between true and inferred ratings is plotted as a function of $\sigma$. 
For small $\sigma$, an approximately linear scaling of $\avrg{\bar{\tau}}$ with $\sigma$ is observed in most cases. 
However, deviations from this behavior emerge and are more pronounced in networks with smaller sizes $N$ or average degrees $\bar{k}$, specially in BA networks.
As shown in the left subpanels, $\avrg{\bar{\tau}}$ scales approximately as $\sim N^{1/2}$ for 1D-lattices, while for all other network models, it appears largely independent of $N$. 
In the right subpanels, it is observed for 1D-lattices that $\avrg{\bar{\tau}}$ scales as $\sim z^{-1.3}$ for sufficiently large values of $z$. 
Similarly, it scales as $\sim \bar{k}^{-2/3}$ for ER networks and $\sim q^{-2/3}$ for BA networks. 
In all cases, deviations from these scalings are observed when networks approach a sparse tree-like structure.
For WS networks, $\avrg{\bar{\tau}}$ decreases with $\ln p$ in a sigmoid-like manner, as shown in the inset of the right subpanel of panel {\bf d)}, where the transition point is found between the small-world and the lattice regimes.

In Fig.~\ref{fig:3}, the mean of the average absolute differences, $\avrg{\bar{\rho}}$, between true and inferred rankings is shown as a function of $\sigma$. 
Interestingly, as it can be seen in more detail in the insets, two distinct regimes are observed.
For small values of $\sigma$, there is a region of perfect retrieval of the true ranking, where $\avrg{\bar{\rho}} = 0$. 
This behavior persists up to a non-trivial threshold, $\sigma^*$, after which $\avrg{\bar{\rho}}$ deviates from zero increasing monotonically with $\sigma$.
As seen in the right subpanels of panels {\bf a)}, {\bf b)}, and {\bf c)}, a transition region emerges in networks with a sufficiently large connectivity, $\bar{k}$. 
In this region, $\avrg{\bar{\rho}}$ shifts from perfect retrieval to a regime where it grows approximately linearly with $\sigma$.
The distinction between these two regimes becomes more pronounced with increasing disorder in the network structure, as illustrated by the WS networks in the right subpanel of panel {\bf d)} as $p$ grows.

To identify a transition point $\sigma_c$ within a transition region, the primitive
$$
\avrg{\bar{\rho}}(\sigma)
=
\frac{A}{B}
\ln
\big(
1+e^{B(\sigma-\sigma_c)}
\big)
$$
of the sigmoid function
$$
\avrg{\bar{\rho}}'(\sigma)
=
\frac{A}{1+e^{-B(\sigma-\sigma_c)}}
$$
is fitted to the curves of Fig.~\ref{fig:3}.
Here, $A$, $B$ and $\sigma_c$ are fitting parameters.
The primitive function approaches linear growth, $\avrg{\bar{\rho}}(\sigma) \approx A(\sigma - \sigma_c)$, for $\sigma \gg \sigma_c$, and approaches zero for $\sigma \ll \sigma_c$. 
The parameter $B^{-1}$ controls the width of the transition region.
The peak of the derivative of the sigmoid function
$$
\avrg{\bar{\rho}}''(\sigma)
=
\frac{ABe^{-B(\sigma-\sigma_c)}}{(1+e^{-B(\sigma-\sigma_c)})^2}
$$
determines the transition point $\sigma_c$. 
The range $[\sigma^*,\sigma^{**}]$ of fitted values for each curve is determined by maximizing the height of the peak, $\avrg{\tilde{\rho}}:=\avrg{\bar{\rho}}''(\sigma_c) = \frac{AB}{4}$, as a function $\sigma^{**}$.
Further details of the fitting procedure can be found in Figs.~1-8 and related text of the Supplemental Material (SM) at~[URL will be inserted by publisher].
The results of these fits are shown in Fig.~\ref{fig:4}, where the transition points $\sigma_c$ and peak heights $\avrg{\tilde{\rho}}$ are plotted as functions of the networks' parameters $N$ and $\theta$.
As seen in the left subpanels, the dependency of these quantities on the network size $N$ is negligible or relatively weak for all network models. 
In contrast, the right subpanels reveal different power-law scalings between these quantities and $\theta$ for the different networks models and $N$ fixed.
More specifically, $\sigma_c$ grows with $z$, $\bar{k}$ and $q$ for 1D-lattices, ER and BA networks, respectively.
In particular, the scaling law has a logarithmic correction for ER networks.
On the other hand, the peak heigth $\avrg{\tilde{\rho}}$ decreases with $z$, $\bar{k}$ and $q$.
For WS networks, both $\sigma_c$ and $\avrg{\tilde{\rho}}$ remain approximately constant in the small-world regime, but undergo a sharp change once the lattice regime is reached.

\begin{figure*}
\includegraphics*[scale=.4]{assets/fig4a.pdf}
\includegraphics*[scale=.4]{assets/fig4b.pdf}
\put(-483,140){\bf a)}
\put(-240,140){\bf b)}
\\
\includegraphics*[scale=.4]{assets/fig4c.pdf}
\includegraphics*[scale=.4]{assets/fig4d.pdf}
\put(-483,140){\bf c)}
\put(-240,140){\bf d)}
\caption{
\label{fig:4}
(color online).
Transition points $\sigma_c$ (green circles) and peak heights $\langle \tilde{\rho} \rangle$ (pink stars) obtained from fits of the primitives of sigmoid functions to the curves $\langle \bar{\rho} \rangle$ vs. $\sigma$ in Fig.~\ref{fig:3}.
Error bars are calculated through error propagation of the errors of the fitted parameters.
The dashed lines represent phenomenological scaling relations fitted to the data points.
}
\end{figure*}

\section{
Discussion
\label{sec:discussion}
}

The scaling of $\avrg{\bar{\tau}}$ vs $N$ indicated in Figs.~\ref{fig:2} for the 1D-lattice differs from the observed in the other network models.
This suggests that the non-local cycles existing in the small-world regime affects the inference of ratings.
The observation is consistent with the behavior of $\langle \bar{\tau} \rangle$ with $\ln p$ in the WS model.
Concerning the scaling of $\langle \bar{\tau} \rangle$ with $z$, $\bar{k}$, and $q$, deviations from such scalings become evident as the networks approach a sparse tree-like regime, where the lack of redundant paths weakens the reinforcement of rating differences between arbitrary items.

The results suggest that the transition from a regime of perfect retrieval $\avrg{\bar{\rho}}=0$ to a regime of partial retrieval $\avrg{\bar{\rho}}>0$ of the true ratings observed in Fig.~\ref{fig:3}, is a consequence of the progressive degradation of the rating differences as the strength $\sigma$ of the quenched dissorder increases.
Namely, this interpretation is supported by the fact that $\avrg{\bar{\tau}}$, $|h|$ and $|s|$ grow almost linearly with $\sigma$ as shown in Fig.~\ref{fig:2} and Figs.~10~and~11 of the SM.
For sufficiently large connectivities, the transition region remains bounded and is approximately defined by $[\sigma^*, \sigma^{**}]$, with a transition point $\sigma_c$ identified by the peak of the primitive of a sigmoid function. 
As connectivity decreases, or as the small-world property weakens with decreasing $p$, the transition region becomes less definite, emphasizing the critical role that redundant non-local paths have in establishing a well-defined ranking.
Finally, Fig.~\ref{fig:4} shows that the properties of the transition are weakly dependent on network size but strongly influenced by the characteristics of the network model and the values of its other parameters.
This observation is further supported by the scaling of the functions defined by $|f|$, $|s|$, $|h|$ and $|g|$ versus $\sigma$, across the different network models, shown for varying $N$ with fixed $\theta$ and varying $\theta$ with fixed $N$ in Figs.~9-12 of the SM.
Additionally, for the sake of completion, scaling relations for the average number of dimensions $\avrg{\kappa_k}$ of the vector spaces $C_k$ of $k$-chains are also shown in Fig.~13 of the SM, and are consistent with the observed behavior for $|s|$ and $|h|$.

\section{
Conclusions
\label{sec:conclusions}
}

This work explores the impact of complex network topologies on the performance of {\em HodgeRank}, a method developed to infer ratings from pairwise comparisons using a combinatorial formulation of differential geometry and Hodge theory over Higher Order Networks. 
The study presents a comprehensive scaling analysis of a variety of metrics, using well-established network models: 1D-lattices, Erdős-Rényi random graphs, Barabási-Albert scale-free networks, and Watts-Strogatz small-world networks. 
Additionally, the analysis incorporates the effects of quenched disorder in the data, which negatively influences the inference of ratings and rankings.
The numerical experiments demonstrate that as the strength of disorder varies, a transition occurs from a regime of perfect retrieval to one of partial retrieval of the true rankings.
The scaling laws identified for the different studied quantities, highlight the critical role of the network structure.
This variability encompasses contributions from both the gradient and cyclic flows of the Hodge decomposition that form the foundation of the HodgeRank method.

Future work could apply more traditional methods from statistical physics to further investigate the properties of the mentioned transition and scaling relations. 
Additionally, this approach could be extended to study the inference of ratings and rankings in empirical data, potentially leading to the development of more realistic models of the network topology and the quenched disorder.
Moreover, future work could study how the quenched disorder affects the inferred ratings and rankings of the individual nodes depending on their local properties.
Finally, using the benchmark experiments here developed, it would be interesting to investigate how HodgeRank compares with alternative methods for the inference of ratings and rankings over a broader range of network models, using modern computer architectures like GPUs or over very large datasets.

\begin{acknowledgments}
The author acknowledges partial support from CONICET under grant PIP2021-2023 number 11220200101100, and from SeCyT (Universidad Nacional de Córdoba, Argentina). Special thanks go to O.V. Billoni for a careful reading of the article, and to the Centro de Cómputo de Alto Desempeño (CCAD) at the Universidad Nacional de Córdoba (UNC), \url{http://ccad.unc.edu.ar/}, which is part of SNCAD – MinCyT, Argentina, for providing computational resources.
\end{acknowledgments}

% Create the reference section using BibTeX:
\bibliography{ref}

% Specify following sections are appendices. Use \appendix* if there
% only one appendix.

\appendix*
\onecolumngrid

\section{\label{appA}Combinatorial differential topology, geometry and calculus}

This section provides an introductory review about combinatorial differential topology, geometry and calculus over simplicial complexes. 
Where appropriate, analogies to the smooth standard version of differential geometry over manifolds are highlighted. 
However, it is important to remark that, despite these similarities, significant differences exits between the combinatorial and smooth frameworks~\cite{ribando2023combinatorial}.

\subsection{Notation}

If  $\{e_i:i\in I\}$ is a basis of a vector space $V$ over a field $\FF$, then the linear decomposition of a vector $v\in V$ is denoted by $v=v^ie_i$ where Einstein's summation convention is assumed and $v^i\in \FF$.
Additionally, if $f\in V^*$ is a covector of the dual vector space of $V$, then $f=f_ie^i$ where $\{e^i:i\in I\}$ is the corresponding dual basis, $f_i\in \FF$ is a coordinate and $e^i(e_j)=\delta^i{}_j$ is Kronecker's delta.
If $a:V\to V$ is a linear operator, then there is a matrix of components $a_i{}^j$ such that $a(e_i)=a_i{}^je_j$.
In this way, $w^je_j=w=a(v)=v^ia(e_i)=v^ia_i{}^je_j$, so $w^j=v^ia_i{}^j=(a^T)^j{}_iv^i$ where $(a^T)^j{}_i$ are the components of the transpose of the matrix of components $a_i{}^j$~\footnote{The first index $i$ of the components $a_{ij}$, $a^{ij}$, $a_i{}^j$ and $a^i{}_j$ of a matrix indicates the row of the entry, while the second $j$ the column. 
There is no predefined relation between the left/right and the upper/lower properties of an index.}.
The kernel of $a$ is denoted by $\krn\,a=\{v\in V:a(v)=0\}$, the range is denoted by $\rng\,a=a(V)=\{a(v):v\in V\}$, the null is denoted by $\nll\,a=\dim\,\krn\,a$ and the rank is denoted by $\rnk\,a=\dim\,\rng\,a$.
An inner product over $V$ maps any two vectors $u,v\in V$ into a scalar $u.v\in \FF$.
It is linear in the first argument if $\FF=\RR$ or antilinear if $\FF=\CC$, and always linear in the second argument.
Moreover, $v.v\geq 0$ for all $v\in V$, so $|v|:=\sqrt{v.v}$ is a norm over $V$ and, in particular, $v.v=0$ implies $v=0$ whenever the inner product is non-degenerate.

\subsection{Simplicial complexes}

An abstract simplicial complex is a set $K$ of subsets of $\{0,1,...,n\}$ that satisfies the property of topological closure: i.e. if $s\in K$ and $r\subseteq s$ is non-empty, then $r\in K$.
The elements $i$ of $\{0,1,...,n\}$ are called nodes.
The elements of $K$ are called simplices.
For each simplex $s\in K$, there exist unique $i_0,...,i_k$ in $\{0,1,...,n\}$ such that $s=\{i_0,i_1,...,i_k\}$ and $i_0<i_1<...<i_k$.
Here, $\dim s=k=|s|-1$ is called the dimension of the simplex $s$ and $\dim K := \max_{s\in K} \dim\, s$ is the dimension of the complex $K$.
Correspondingly, $s$ is called a $k$-simplex and $K$ a $(\dim K)$-complex.
$K_k$ denotes the set of all $k$-simplices in $K$ and its cardinality is $\kappa_k:=|K_k|$, while $\kappa:=\sum_k\kappa_k=|K|$ is the cardinality of the complex $K$.
In particular, $K_{-1}=K_{n+1}=\emptyset$ are defined for convenience.
For any different nodes $i,j,k,l \in \{0,1,...,n\}$, 0-simplices like $\{i\}$ are in one to one correspondence with nodes, 1-simplices like $\{i,j\}$ can be thought of links, 2-simplices like $\{i,j,k\}$ as triangles, 3-simplices like $\{i,j,k,l\}$ as tetrahedra, and so on.
In particular, a simplicial complex of dimension $\dim K=1$ is equivalent to a non-directed graph without self-links.

A simplex $r$ is a face of another $s$ if $r\subset s$.
There are ${k + 1 \choose q + 1}$ $q$-faces within a $k$-simplex.
Moreover, $r$ is a facet of $s$ if $r\subset s$ and $1+\dim\,r=\dim\,s$.
Correspondingly, $s$ is a coface/cofacet of $s$ if $r$ is a face/facet of $s$, respectively.
Additionally, $r$ and $s$ are incident to each other if $r$ is a facet of $s$ or $s$ is a facet of $r$.
Furthermore, $r$ and $s$ are lower-adjacent to each other if they have a facet in common, and are upper-adjacent to each other if they are both different facets of a common simplex.
In particular, two vertices of a common graph are upper-adjacent if they are both different facets of a common edge.
The degree $k_s$ of a simplex $s$ is its number of cofacets.
The degree $k_s$ of a simplex $s$ should not be confused with the dimension $k$ of a $k$-simplex.
The generalized degree $\zeta_{sq}$ of a $k$-simplex $s$ is its number of $q$-cofaces.
In the particular case of common graphs or networks, $k_i=\zeta_{i1}$ for any node $i$.

A simplicial complex $K$ can be obtained from a graph $g$ where each $(k+1)$-clique of $g$ corresponds to a $k$-simplex of $K$.
Such simplicial complex obtained from $g$ is called the {\em clique complex} of $g$.
In particular, the $(k+1)$-clique complex of a graph $g$ is the simplicial complex obtained by considering only the $(q+1)$-cliques of $g$ for $q\leq k$.

\subsection{Chains}

The set of functions from $K$ to some number field $\FF$ is a vector space $C$ of vectors called chains.
The functions $e_s : K\to \FF$ for $s\in K$ such that $e_s(r)=\delta_{sr}$ form a canonical basis of $C$~\footnote{Here, $\delta_{sr}$ is the Kronecker delta.}.
In particular, if $s$ is a $k$-simplex, then $e_s$ is called a $k$-chain.
The subset of linear combinations of $k$-chains form a subspace $C_k$ of $C$.
Crucially,
$$
C=C_0\oplus C_1\oplus ... \oplus C_n
$$
where $C_0=C_n\cong\FF$, i.e. $C_0$ and $C_n$ are isomorphic to $\FF$.
For convenience, $C_{-1}=C_{n+1}=\{0\}$ are also defined.
Note, $m_k=\dim\, C_k\leq {n\choose k}$ for $k=0,1,...,n$, and $\dim\, C_{-1}=\dim\, C_{n+1}=0$. 
Therefore, $\dim\,C\leq \sum_{k=0}^n {n\choose k}=2^n$.
Using Einstein's notation, any chain of $C$ can be represented by a linear combination $c=c^se_s$ with unique coefficients $c^s\in \FF$.

\subsection{Cochains}

The dual space $C^*$ of $C$ is called the vector space of cochains of $K$.
If $C_k^*$ is the dual space of $C_k$, then
$$
C^*=C_0^*\oplus C_1^*\oplus ... \oplus C_n^*
$$
For convenience, $C_{-1}^*=C_{n+1}^*=\{0\}$ are also defined.
The covectors in $C_k^*$ are called $k$-cochains.
In particular, $\{e^s:s\in K\}$ denotes the basis of $C^*$ dual to the canonical basis $\{e_s:s\in K\}$ of $C$.
As a consequence, $e^s(e_r)=\delta^s{}_r$ for all $s,r\in K$.
In this way, any cochain $f\in C^*$ can be represented by a linear combination $f=f_se^s$ of coefficients $f_s\in \mathbb{F}$.

The application of a cochain $f$ to a chain $c$ equals $f(c)=f_se^s(c^re_r)=f_sc^re^s(e_r)=f_sc^r\delta^s{}_r=f_sc^s\in \FF$.
In a sense, $f(c)$ is the combinatorial analogous of the integration $\int_c f$ of form $f$ over a manifold $c$.
It should be remarked, however, that a simplicial complex is not always obtained from the discretization of a manifold.
This is the reason for which, in this work, the term {\em combinatorial} is preferred over the term {\em discrete}.

\subsection{Orientation}

Let $e_{i_0i_1...i_k}:=e_{\{i_0,i_1,...,i_k\}}$ when $0\leq i_0<i_1<...<i_k\leq n$.
Otherwise, let 
$$
e_{j_0j_1...j_k}
=
\left\{
\begin{array}{rl}
e_{i_0i_1...i_k} & \mbox{if $j_0j_1...j_k$ is an even permutation of $i_0i_1...i_k$,} \\
-e_{i_0i_1...i_k} & \mbox{if it is an odd permutation, and} \\
0 & \mbox{otheriwse.}
\end{array}
\right.
$$
The sign between $e_{j_0j_1...j_k}$ and $e_{i_0i_1...i_k}$ is called the orientation of $e_{j_0j_1...j_k}$ relative to $e_{i_0i_1...i_k}$.
Like the orientation of the axis of $\RR^n$, the orientation of the simplices is a matter of convention, which is consisten with the arbitrariness in the enumeration of the nodes.

\subsection{
Boundaries
\label{sec:boundaries}
}

The boundary operator over chains $\partial : C\to C$ is defined by
$$
\partial = \partial_0+\partial_1+...+\partial_n
$$
where $\partial_{-1}=\partial_0=\partial_{n+1}=0$, and
$$
%\partial_k(e_s)
\partial_k(e_{i_0i_1...i_q})
=
\delta_{qk}
\sum_{j=0}^q
(-1)^j 
e_{i_0i_1...\hat{i}_j...i_k}
$$
for $0<k\leq n$, where $e_{i_0i_1...\hat{i}_j...i_k}:=e_{i_0i_1...i_{j-1}i_{j+1}...i_k}$.
Here, the sign of $(-1)^j$ is called the orientation that the $k$-chain $e_{i_0i_1...i_k}$ induces on the $(k-1)$-chain $e_{i_0i_1...\hat{i}_j...i_k}$.

The boundary operator satisfies the following crucial properties.
Firstly, $\partial$ is a linear operator.
Secondly, $\partial_k(C)=\partial(C_k)\subseteq C_{k-1}$, meaning that the boundary of a simplex of dimension $k$ has dimension $k-1$.
Thirdly, $\partial^2=0$, meaning that a boundary has no boundary.
This third statement follows from
$$
\partial^2 
=
\sum_{jk}\,
\partial_j \partial_k
\nonumber
\\
=
\sum_{k}\,
\partial_{k}\partial_{k+1}
$$
and the fact that $\partial_{k}\partial_{k+1}=0$ follows from the alternating definition of $\partial_k$ and that chains are oriented.

In matricial form 
$\partial_k(e_s)=(\alpha_k)_s{}^r e_r$.
In this way, $q^re_r=q=\partial_k(c)=\partial_k(c^se_s)
=c^s\partial_k(e_s)=c^s(\alpha_k)_s{}^re_r$.
Hence, $q^r=c^s(\alpha_k)_s{}^r=(\alpha_k^T)^r{}_sc^s$.
Note, $(\alpha_k)_s{}^r\neq 0$ if and only if $s$ and $r$ are incident to each other.
Hence, $\alpha_k\in \RR^{m_{k}\times m_{k-1}}$ or its transpose $\alpha_k^T$ are usually called the $k$-th incidence matrix of the simplicial complex~\cite{grady2010discrete,newman2018networks,bianconi2021higher}.
In this work, the convention where the incidence matrix is the transpose $\alpha_k^T$ is adopted.

\subsection{Coboundaries}

The coboundary operator $d:C^*\to C^*$ is the combinatorial analogous of the differential operator.
It is defined by
$$
(d(f))(c) = f(\partial(c))    
$$
for all $f\in C^*$ and $c\in C$, which can be recognized as the combinatorial analogous of the generalized Stoke's theorem
$$
\int_c df = \int_{\partial c} f
$$
which in turn is the generalization of the fundamental theorem of calculus to arbitrary dimensions.
Crucially, 
$$
d=d_0+d_1+...+d_n
$$
where $d_{-1}=d_n=d_{n+1}=0$,  $d_k(C^*)=d(C_k^*)\subseteq C_{k+1}^*$,
and $(d_k(f))(c)=f(\partial_{k+1}(c))$ for all $c\in C$.
Moreover, $d^2=0$ and $d_k d_{k-1}=0$ easily follows from $\partial^2=0$.
This fact is the reason for which many differential equations in physics are of second order.

In matricial form 
$d_k(e^s)=(\beta_k)^s{}_r e^r$.
Hence, 
\begin{eqnarray}
(d_k(e^s))(e_r)
&=&
e^s(\partial_{k+1}(e_r))
\nonumber
\\
((\beta_k)^s{}_te^t)(e_r)
&=&
e^s((\alpha_{k+1})_r{}^le_l)
\nonumber
\\
(\beta_k)^s{}_t\,e^t(e_r)
&=&
(\alpha_{k+1})_r{}^l\,e^s(e_l)
\nonumber
\\
(\beta_k)^s{}_t\,\delta^t{}_r
&=&
(\alpha_{k+1})_r{}^l\,\delta^s{}_l
\nonumber
\\
(\beta_k)^s{}_r
&=&
(\alpha_{k+1})_r{}^s
\nonumber
\end{eqnarray}
In other words, $\beta_k=(\alpha_{k+1}^T)$.
Note, $g_re^r=g=d_k(f)=d_k(f_se^s)=f_sd_k(e^s)
=f_s(\beta_k)^s{}_re^r$.
Hence $g_r=f_s(\beta_k)^s{}_r=(\beta_k^T)_r{}^sf_s=(\alpha_{k+1})_r{}^sf_s$.

Note, $\grad = d_0:C_0^*\to C_1^*$ and 
$\curl = d_1:C_1^*\to C_2^*$ are recognized as a combinatorial analogous of the gradient and the curl operators of standard differential calculus~\footnote{Strictly speaking, in standard differential calculus, the gradient operator is defined in a way that is metric dependent.}, respectively.
In particular, the known result $\curl\circ\grad=-d_1 d_0=0$ follows as a particular case of $d_k d_{k-1}=0$.
As an example on how these operators work, note that if $f\in C_0^*$ and $g=\grad\, f=d_0(f)$, then $g_{ij}=g(e_{ij})=(d_0(f))(e_{ij})=f(\partial_1(e_{ij}))=f(e_j-e_i)=f_j-f_i$.

\subsection{Inner products and geometry}

Consider a positive definite inner product from $C$ to $\RR$.
By the Riesz's representation theorem, a positive definite inner product over $C$ induces an isomorphism $C\ni c\to c^{\flat}\in C^*$ called {\em flat} defined by
$$
c^{\flat}(q) = c.q
$$
for all $c,q\in C$.
Its inverse is denoted by $C^*\ni f\to f_{\sharp}\in C$ and is called {\em sharp}.
Together, $\flat$ and $\sharp$ are called musical isomorphisms.
They induce a positive definite inner product over $C^*$ given by $f.g = f_{\sharp}.g_{\sharp}$ for all $f,g\in C^*$.

Consider a non-degenerate inner product over $C$.
Its coordinates in the canonical basis are $m_{sr}=e_s.e_r=e_r.e_s=m_{rs}$ for $s,r\in K$.
Since $e_s^{\flat}\in C^*$, there exists $t_{sr}\in \RR$ such that $e_s^{\flat}=t_{sr}e^r$ and, therefore,
$$
m_{sq} = e_s.e_q = e_s^{\flat}(e_q)=t_{sr}e^r(e_q)=t_{sr}\delta^r{}_q=t_{sq}
$$
In consequence
$$
m_{sr}=e_s.e_r=e_s^{\flat}.e_r^{\flat}=(m_{st}e^t).(m_{rq}e^q)=m_{st}m_{rq}(e^t.e^q)=m_{st}m_{rq}m^{tq}
$$
so $0=m_{st}(\delta^t{}_r-m_{rq}m^{tq})$ and, as a result, $m_{st}m^{tr}=\delta_s{}^r$.
In other words, the matrices of components $m_{st}=e_s.e_t$ and $m^{tr}=e^t.e^r$ are the inverse of each other.
In the context of abstract simplicial complexes, it is often assumed that the canonical basis $\{e_s:s\in K\}$ of $C$ is orthogonal, so $m_{rs}$ and $m^{rs}$ are the components of diagonal matrices.
Following this trend, in the present work, the trivial inner product $m_{rs}=\delta_{rs}$ is considered.

In differential geometry, an inner product over differencial forms is usually defined in terms of a volume form and an associated Hodge's star operator which in turn depend on a metric tensor.
When a simplicial complex is derived from the discretization of manifold, these constructions can be also available, highlighting the intimate relation between the inner product over chains or cochains, and the geometric properties of simplicial complexes.
In some cases, an hermitian inner product over $\CC$ is preferred.
When this is the case, it can be incorporated into the formalism by the complexification of $C$~\cite{roman2005advanced}.

\subsection{Dual boundaries}

The dual boundary operator $\partial^*:C\to C$ is defined as the inner-product adjoint of the boundary operator $\partial$, i.e. by
$$
\partial^*(c).q = c.\partial(q)
$$
for all $c,q\in C$.
Crucially, $(\partial^*)^2=0$ follows from $\partial^2=0$, and
$$
\partial^*=\partial_0^*+\partial_1^*+...+\partial_n^*
$$
where  $\partial_{-1}^*=\partial_{0}^*=\partial_{n+1}^*=0$ and $\partial_k^*$ is the dual of $\partial_k$, so $\partial_k^*(C)=\partial^*(C_{k-1})\subseteq C_k$.

In matricial form 
$\partial_k^*(e_s) = (\mu_k)_s{}^r e_r$.
Hence,
\begin{eqnarray}
\partial_k^*(e_s).e_r
&=&
e_s.\partial_k(e_r)
\nonumber
\\
(\mu_k)_s{}^t\, e_t.e_r
&=&
(\alpha_k)_r{}^p\, e_s.e_p
\nonumber
\\
(\mu_k)_s{}^t\, m_{tr}
&=&
(\alpha_k)_r{}^p\, m_{sp}
\nonumber
\\
(\mu_k)_s{}^t\, m_{tr}m^{rl}
&=&
(\alpha_k)_r{}^p\, m_{sp}m^{rl}
\nonumber
\\
(\mu_k)_s{}^t\, \delta_t{}^l
&=&
(\alpha_k)_r{}^p\, m_{sp}m^{rl}
\nonumber
\\
(\mu_k)_s{}^l
&=&
m^{lr}\,(\alpha_k)_r{}^p\, m_{ps}
\nonumber
\end{eqnarray}
Moreover, in the particular case where $m_{sr}=\delta_{sr}$, so $m^{sr}=\delta^{sr}$, it is found that $(\mu_k)_s{}^t\delta_{tr}=(\alpha_k)_r{}^p\, \delta_{sp}$, so $(\mu_k)_s{}^r=(\alpha_k)_r{}^s$ and, therefore, $\mu_k=\alpha_k^T$.
Note, $q^re_r=q=\partial_k^*(c)=\partial_k^*(c^se_s)
=c^s\partial_k^*(e_s)=c^s(\mu_k)_s{}^re_r$.
Hence, $q^r=c^s(\mu_k)_s{}^r=(\mu_k^T)^r{}_sc^s=(\alpha_k)^r{}_sc^s$.

\subsection{Dual coboundaries}

The dual coboundary operator $d^*:C^*\to C^*$ is defined as the inner-product adjoint of the coboundary operator $d$, i.e. by
$$
d^*(f).g = f.d(g)
$$
for all $f,g\in C^*$.
The dual coboundary operator is the combinatorial analogous of the codifferential operator.
As before, $(d^*)^2=0$ follows from $d^2=0$, and
$$
d^*=d_0^*+d_1^*+...+d_n^*
$$
where $d_{-1}^*=d_{n}^*=d_{n+1}^*=0$ and $d_k^*$ is the dual of $d_k$, so $d_k^*(C^*)=d^*(C_{k+1}^*)\subseteq C_k^*$.

In matricial form
$d_k^*(e^s)=(\nu_k)^s{}_r e^r$.
Hence
\begin{eqnarray}
d_k^*(e^s).e^r
&=&
e^s.d_k(e^r)
\nonumber
\\
(\nu_k)^s{}_t\,e^t.e^r
&=&
(\beta_k)^r{}_p\,e^s.e^p
\nonumber
\\
(\nu_k)^s{}_t\,m^{tr}
&=&
(\beta_k)^r{}_p\,m^{sp}
\nonumber
\\
(\nu_k)^s{}_t\,m^{tr}m_{rl}
&=&
(\beta_k)^r{}_p\,m^{sp}m_{rl}
\nonumber
\\
(\nu_k)^s{}_t\,\delta^t{}_l
&=&
(\beta_k)^r{}_p\,m^{sp}m_{rl}
\nonumber
\\
(\nu_k)^s{}_l
&=&
m_{lr}\,(\beta_k)^r{}_p\,m^{ps}
\nonumber
\end{eqnarray}
Moreover, in the particular case where $m^{sr}=\delta^{sr}$, so $m_{sr}=\delta_{sr}$, it is found that
$(\nu_k)^s{}_t\delta^{tr}=(\beta_k)^r{}_p\delta^{sp}$, so $(\nu_k)^s{}_r=(\beta_k)^r{}_s$ and, therefore, $\nu_k=\beta_k^T=\alpha_{k+1}$.
Note, $g_re^r=g=d_k^*(f)=d_k^*(f_se^s)
=f_sd_k^*(e^s)=f_s(\nu_k)^s{}_re^r$.
Hence, $g_r=f_s(\nu_k)^s{}_r=(\nu_k^T)_r{}^sf_s=(\beta_k)_r{}^sf_s=(\alpha_{k+1}^T)_r{}^sf_s$.

Note, $\dive = -d_0^*:C_1^*\to C_0^*$ can be recognized as the combinatorial analogous of the divergence of standard differential calculus.

\subsection{Hodge decomposition}

Consider vector spaces $U,V$ and $W$ equipped with an inner product and two linear operator $a:V\to W$ and $b:U\to V$ such that 
\begin{equation}
\label{eqA1}
ab=0
\end{equation}
Then $\rng(b)\subseteq \krn(a)$ follows, enabling the definition of the so called (co)homology group~\footnote{Actually, it is a quotient space. But it can be considered an abelian group under addition.}
\begin{equation}
H:=
\krn(a)/\rng(b)
\cong
\krn(a)\cap \krn(b^*)
\cong
\krn\,L
\nonumber
\end{equation}
whose elements are called (co)homology classes.
Here, $L:=a^*a+bb^*:V\to V$ is called the Hodge Laplacian associated to $a$ and $b$.
It can be shown that
\begin{equation}
\label{eqA2}
V=\rng\,a^*\oplus\krn\,L\oplus\rng\,b
\end{equation}
meaning that, for any vector $v\in V$, there exist the unique decomposition
\begin{equation}
\label{eqA3}
v = a^*(w)+h+b(u)
\end{equation}
for some $w\in W$ and $u\in U$, and where 
$h\in \krn\,L\subseteq V$ is the unique vector of the (co)homology class $[v]\in \krn\,a/\rng\,b$ satisfying $h\in(\rng\,b)^{\perp}=\krn\,b^*$.
For this reason, $h$ is called the harmonic representative of $[v]$.
Equations~\ref{eqA2}~and~\ref{eqA3} represent the so called Hodge decomposition of $V$ and $v$, respectively.
Importantly, $\krn\,b^*=\rng\,a^*\oplus \krn\,L$ and $\krn\,a=\krn\,L\oplus\rng\,b$.
Moreover, $L$ is a positive semidefinite operator since $L=L^*$, and $(a^*a)^*=a^*a$, $(bb^*)^*=bb^*$.
Furthermore, if $[X,Y]:=XY-YX$, then $[L,a^*a]=[L,bb^*]=[a^*a,bb^*]=0$.

\subsection{Homology}

In the context of simplicial complexes, homology emerges from the application of Hodge's decomposition using $\partial_{k}$ and $\partial_{k+1}$ in place of $a$ and $b$ respectively, since $ab=\partial_{k}\partial_{k+1}=0$.
As a result, the Hodge Laplacian takes the form
$$
\mathcal{L}_k
=
\mathcal{L}_k^{\downarrow}
\oplus
\mathcal{L}_k^{\uparrow}
:C_k\to C_k
$$
where
$$
\mathcal{L}_k^{\downarrow}=\partial_{k}^*\partial_{k} : C_k\to C_{k-1}\to C_k
\hspace{1cm}
\mbox{and}
\hspace{1cm}
\mathcal{L}_k^{\uparrow}=\partial_{k+1}\partial_{k+1}^*:C_k\to C_{k+1} \to C_k
$$
so
$$
C_k 
= 
\rng\,\partial_k^*
\oplus
\krn\,\mathcal{L}_k
\oplus
\rng\,\partial_{k+1}
$$
and, therefore, for any $c\in C_k$, there exists unique 
$p\in \rng\,\partial_k^*\subseteq C_k$, $h\in \krn\,\mathcal{L}_k\subseteq C_k$ and $q\in \rng\,\partial_{k+1}\subseteq C_k$
such that $c = p+h+q$.
Importantly, $\krn\,\partial_{k+1}^*=\rng\,\partial_{k}^*\oplus \krn\,\mathcal{L}_k$ and $\krn\,\partial_{k}=\krn\,\mathcal{L}_k\oplus\rng\,\partial_{k+1}$.
Often, $\mathcal{L}_k$, $\mathcal{L}_k^{\downarrow}$ and $\mathcal{L}_k^{\uparrow}$ are called the $k$-th, $k$-th lower and $k$-th upward Laplacians, respectively.
These can be aggregated to obtain a Laplacian over $C$
$$
\mathcal{L} = \partial^*\partial \oplus \partial\partial^* = \mathcal{L}_0\oplus...\oplus \mathcal{L}_n : C\to C
$$
Note, $\mathcal{L}_k^{\downarrow}$ and $\mathcal{L}_k^{\uparrow}$ combine the information of $k$-chains with the information of $(k-1)$-chains and $(k+1)$-chains respectively.

The $k$-th Dirac operator over $C$
$$
\mathcal{D}_k=\partial_k^*+\partial_k
$$
can be also introduced, and it is related to the concept of spinor~\cite{bianconi2021topological}.
It satisfies $\mathcal{D}_k^2=\mathcal{L}_k^{\downarrow}\oplus\mathcal{L}_{k-1}^{\uparrow}$, combining the information between $(k-1)$-chains and $k$-chains.
Similarly, $\mathcal{D}=\partial^*+\partial=\mathcal{D}_0+...+ \mathcal{D}_n$ is such that $\mathcal{D}^2=\mathcal{L}$.

The $k$-th Betti numbers $\beta_k=\rnk\, H_k=\rnk\,\krn \,\partial_k - \rnk\,\rng\,\partial_{k+1}$, which are the ranks of the homology groups, are topological invariants  characterizing the properties of topological spaces.
In the present context where $C$ is a vector space, these homology groups are subspaces, so their ranks are actually their dimensions.

Note, when $m_{rs}$ is the trivial metric $\delta_{rs}$, then $(\mathcal{L}_k^{\downarrow})^s{}_s=(\alpha_k)^s{}_r(\alpha_k^T)^r{}_s=k+1$ is the number of facets of $s$.
Similarly, $(\mathcal{L}_k^{\uparrow})^s{}_s=(\alpha_{k+1}^T)^s{}_r(\alpha_{k+1})^r{}_s=k_s=\zeta_{s,k+1}$ is the number of cofacets or degree of $s$.
For a general metric, $(\mathcal{L}_k^{\downarrow})^s{}_s$ and $(\mathcal{L}_k^{\uparrow})^s{}_s$ represent weighted versions of these quantities.

\subsection{Cohomology\label{appA:cohomology}}

Cohomology emerges from the application of  Hodge decomposition over cochains, using $d_{k}$ and $d_{k-1}$ in place of $a$ and $b$ respectively, so $ab = d_kd_{k-1}=0$.
As a result, the Hodge Laplacian takes the form
$$
L_k
=
L_k^{\uparrow}
\oplus
L_k^{\downarrow}
:C_k^*\to C_k^*
$$
where
$$
L_k^{\uparrow}=d_{k}^*d_{k}:C_k^*\to C_{k+1}^* \to C_k^*
\hspace{1cm}
\mbox{and}
\hspace{1cm}
L_k^{\downarrow}=d_{k-1}d_{k-1}^* : C_k^*\to C_{k-1}^*\to C_k^*
$$
so
$$
C_k^*
=
\rng\,d_k^*
\oplus
\krn\,L_k
\oplus
\rng\,d_{k-1}
$$
and, therefore, for any $f\in C_k^*$, there exists unique 
$s\in \rng\, d_k^*\subseteq C_k^*$,
$h\in \krn\, L_k\subseteq C_k^*$ and
$g\in \rng\, d_{k-1}\subseteq C_k^*$
such that $f = s+h+g$.
Usually, $s$, $h$ and $g$ are called the coexact, harmonic and exact components of $f$ respectively.
Importantly, $\krn\,d_{k-1}^*=\rng\,d_k^*\oplus \krn\,L_k$ and $\krn\,d_k=\krn\,L_k\oplus\rng\,d_{k-1}$.
The Laplacians $L_k^{\uparrow}$ and $L_k^{\downarrow}$ combine the information in $k$-cochains with the information in $(k+1)$-cochains and $(k-1)$-cochains respectively.
Moreover, they can be combined to obtain the so called Hodge Laplacian over $C^*$
$$
L = dd^* \oplus d^*d= L_0\oplus...\oplus L_n : C^*\to C^*
$$
Furthermore, in analogy to its topological counterpart $\mathcal{D}_k$, the $k$-th Dirac operator over cochains
$$
D_k = d_k^*+d_k
$$
can be introduced.
It satisfies $D_k^2=L_k^{\uparrow}\oplus L_{k+1}^{\downarrow}$, combining the information between $k$-cochains and $(k+1)$-cochains.
Correspondingly, $D=d^*+d=D_0+...+ D_n$ is such that $D^2=L$.

The Hodge decomposition of chochains is a generalization of the Helmholtz decomposition of vector calculus in 3D, to arbitrary combinatorial topologies and dimensions.
In fact, the coexact, harmonic and exact componets of the Hodge decomposition are recognized as the solenoid, harmonic and gradient components of the Helmholtz decomposition.
Moreover, $L_0=d_0^*d_0+d_{-1}d_{-1}^*=d_0^*d_0+0=-\dive\circ \grad$ can be recognized as the {\em graph Laplacian}, a combinatorial analogous of the Laplace operator over scalar fields, and $L_1=d_0d_0^*+d_1^*d_1 = \grad\circ \grad^*+\curl^*\circ \curl=-\grad\circ \dive+\curl^*\circ \curl$ can be recognized as the {\em graph Helmholtzian} which is the combinatorial analogous of the vector Laplacian~\cite{lim2020hodge}.
When a simplicial complex is obtained from a discretization of an oriented region in euclidean space, then a Hodge star operator can be defined, relating chains and cochains with volume forms.
In particular, when the triangulation is a 3D-space, then $\curl^*=\curl$ by Hodge duality.
This is the reason for which there is no special name for $\curl^*$ in Physics~\cite{grady2010discrete}.

\end{document}
%
% ****** End of file apstemplate.tex ******

### 2025-07-10 code/enumerator.jl
progress

@@ -1,0 +1,18 @@
+"""
+Input:
+    d::Dict{Any,Int} : container
+    k::Any : key
+Returns the number associated to key k if it already exists in d. Ohterwise, creats a new key-value pair in d with key k and value equal to the current size of d plus one.
+"""
+function enum!(d::Dict{Any,Int},k)
+  return get!(d,k,length(d)+1)
+end
+
+function enumerator_test()
+  d = Dict{Any,Int}()
+  @show enum!(d,"a")
+  @show enum!(d,"b")
+  @show enum!(d,"c")
+  @show enum!(d,"b")
+  @show d
+end

### 2025-07-10 code/running-stats.jl

@@ -1,0 +1,45 @@
+mutable struct RunningStat
+  n::Int
+  m::Real # mean
+  s::Real # std
+end
+
+function RunningStat()
+  return RunningStat(0,0.0,0.0)
+end
+
+function Base.push!(rs::RunningStat,v::Real)
+  if rs.n>0
+    rs.n += 1
+    new_m = rs.m + (v-rs.m)/rs.n
+    rs.s += (v-rs.m)*(v-new_m)
+    rs.m = new_m
+  else
+    rs.n = 1
+    rs.m = v
+    rs.s = 0
+  end
+  return rs.m
+end
+
+function Statistics.mean(rs::RunningStat)
+  return rs.m 
+end
+
+function Statistics.var(rs::RunningStat)
+  return rs.s/(rs.n-1)
+end
+
+function Statistics.std(rs::RunningStat)
+  return sqrt(var(rs))
+end
+
+function running_stats_test()
+  list_v = 10 .+ 5.0*randn(1000)
+  rs = RunningStat()
+  for v in list_v
+    push!(rs,v)
+  end
+  @show mean(rs),std(rs)
+  @show mean(list_v),std(list_v)
+end

### 2025-07-10 setup-julia.sh
progress

@@ -1,0 +1,5 @@
+curl -fsSL https://install.julialang.org | sh
+juliaup add 1.11.5
+juliaup default 1.11.5
+$HOME/.juliaup/bin/julialauncher --eval="using Pkg; Pkg.activate(\".\"); Pkg.instantiate()"
+$HOME/.juliaup/bin/julialauncher

### 2025-07-11 draft/draft.tex
progress

@@
 \begin{abstract}
-The inference of rankings plays a central role in the theory of social choice, which seeks to establish preferences from collectively generated data, such as pairwise comparisons.
-Examples include political elections, ranking athletes based on competition results, ordering web pages in search engines using hyperlink networks, and generating recommendations in online stores based on user behavior.
-Various methods have been developed to infer rankings from incomplete or conflicting data.
-One such method, HodgeRank, introduced by Jiang et al.~\cite{jiang2011statistical}, utilizes Hodge decomposition of cochains in Higher Order Networks to disentangle gradient and cyclical components contributing to rating scores, enabling a parsimonious inference of ratings and rankings for lists of items.
-This paper presents a systematic study of HodgeRank's performance under the influence of quenched disorder and across networks with complex topologies generated by four different network models.
-The results reveal a transition from a regime of perfect trieval of true rankings to one of imperfect trieval as the strength of the quenched disorder increases.
-A range of observables are analyzed, and their scaling behavior with respect to the network model parameters is characterized.
-This work advances the understanding of social choice theory and the inference of ratings and rankings within complex network structures.
+Bla bla...
 \end{abstract}
@@
 
-The rating and ranking candidates, agents, items or options based on collectively expressed preferences is deeply ingrained modern human civilization, making the study of inference methods for ratings and rankings under diverse conditions highly significant.
-Ratings and rankings are often derived from pairwise comparisons, which naturally form a network structure. 
-Understanding how the topology of these networks influences the accuracy of rating and ranking inference is of critical importance.
-
-The modern theory of ratings and rankings has its origins in the field of social choice, with early foundations laid in the late 13th century by Ramon Llull in his {\em Ars Electionis}~\cite{colomer2013ramon}.
-Llull's work introduced the concept of pairwise comparisons, the majority principle, and fairness in voting systems. 
-Unfortunately, his contributions were largely forgotten over time. 
-A mathematical revival of these ideas emerged in the late 18th century with the works of Borda and Condorcet~\cite{borda1781memoire,condorcet1785essai}, who laid the groundwork for modern voting and ranking theory.
-Later, in 1929, Zermelo proposed a model for ranking inference using ratings as latent variables within a probabilistic framework~\cite{zermelo1929berechnung}, but his work was also overlooked~\cite{newman2023efficient}.
-In the 1950s, Bradley and Terry reintroduced Zermelo’s model in a broader context~\cite{bradley1952rank}. 
-In the 1960s, a model based on similar principles was introduced and promoted by \'El\H{o}, eventually becoming the default rating system used by the World Chess Federation (FIDE)~\cite{elo1978rating}.
-More recently, the advent of the World Wide Web and search engines highlighted the significance of network topology in ranking systems, exemplified by the success of PageRank, the original algorithm behind Google and introduced Brin and Page at the end of the 90's~\cite{brin1998anatomy}.
-The study of rating and ranking systems remains an active area of research~\cite{jiang2011statistical,csato2017ranking,debacco2018physical,newman2022ranking,jerdee2023luck}, as collective decision-making plays a vital role in modern civilization, particularly in the information age.
-
-In 1736, Euler solved the Seven Bridges of K\"onigsberg problem, establishing the foundations of graph theory and foreshadowing the development of topology. 
-Graphs, or networks, can be viewed as a specific type of topological space. 
-Within the theory of Higher Order Networks~\cite{bianconi2021higher,battiston2021physics,majhi2022dynamics,bick2023what,ferraz2024contagion,muolo2024global,bianconi2024nature,bianconi2024theory}, hypergraphs extend the concept of graphs by incorporating hyperlinks, that may be used to represent many-body interactions or multiary-relations.
-However, unlike graphs, hypergraphs are not necessarily topological spaces, as they may not satisfy the condition of topological closure. 
-For example, a hypergraph may include a hyperlink $\{i,j,k\}$ without containing the link $\{i,j\}$. 
-When hypergraphs satisfy the closure condition, they form a special class of topological spaces known as abstract simplicial complexes, or simply, simplicial complexes.
-
-Although terminology and definitions may vary by context, many concepts from differential geometry on smooth manifolds can be adapted to the discrete or combinatorial setting of simplicial complexes~\cite{bobenko2008discrete,grady2010discrete,aygun2011spectral,tuncer2015spectral,rovelli2019natural,bianconi2021higher}.
-Namely, discrete or combinatorial analogous of forms, gradients, divergences and Laplacians can be defined over simplicial complexes.
-This is the reason for which simplicial complexes are often preferred over general hypergraphs.
-In particular, and thanks to the property of topological closure, Hodge theory can be applied to decompose cochains over simplicial complexes into exact, coexact and harmonic components, much like the analogous decomposition of tensor fields~\cite{grady2010discrete,lim2020hodge}.
-This is the approach followed by Jian et al.~\cite{jiang2011statistical} who developed HodgeRank, a method for the inference of ratings that leverages Hodge theory over simplicial complexes to disentangle different contributions to the inferred ratings.
-The present article takes advantage of HodgeRank to study the inference of ratings and rankings in complex topologies~\cite{newman2018networks,perotti2009emergent}.
-Section~\ref{sec:theory} provides an introduction to the theory of rating and ranking inference, and focuses on HodgeRank.
-Section~\ref{sec:results} presents the results of numerical experiments using HodgeRank applied to networks with complex topologies.
-Section~\ref{sec:discussion} discuss the results.
-Section~\ref{sec:conclusions} summarizes the findings and outlines directions for future research.
+Bla bla...
 
@@
 
-Several methods exist for inferring ratings and rankings from pairwise comparison data~\cite{jiang2011statistical}.
-This section focuses on a particular approach inspired by the foundational works of Zermelo~\cite{zermelo1929berechnung} and Bradley-Terry~\cite{bradley1952rank}, to illustrate how pairwise comparisons can be associated with probabilities that, in turn, relate to rating scores.
-These concepts are then used to introduce the HodgeRank method, followed by a discussion of the numerical challenges that arise within this framework.
-
-Consider a set of $N=n+1$ items indexed by $i \in \{0, 1, \dots, n\}$. 
-These items or agents could represent movies subject to pairwise comparison or even chess players competing in chess tournaments~\cite{perotti2013innovation}. 
-For any two items $i$ and $j$, there exists a (possibly empty) sequence $s_{ij} = (s_{ij1}, s_{ij2}, \dots, s_{ijt}, \dots)$ of comparison results, where $s_{ijt} = -s_{jit}$ takes the value $1$, $0$, or $-1$ depending on whether option $i$ wins, draws, or loses the $t$-th comparison against option $j$, respectively.
-For example, these results could represent outcomes of chess games between players, or voting preferences over movies. 
-Importantly, the set of non-empty sequences, or {\em pairings}, defines the links of an undirected network with $N$ nodes and no self-links, described by an adjacency matrix $a$ with entries $a_{ij} = a_{ji} \in \{0, 1\}$ for $i,j \in \{0, 1, \dots, n\}$.
-The degree, or the number of links adjacent to node $i$, is $k_i = \sum_j a_{ij} = \sum_j a_{ji}$, and the average degree is $\bar{k} = N^{-1} \sum_i k_i = 2M/N$, where $M = \frac{1}{2} \sum_{ij} a_{ij}$ is the total number of links.
-
-Assuming that the results within a sequence are statistically independent and are identically distributed, the probability for a particular sequence to occur is 
-\begin{eqnarray}
-\label{eq:v_and_q}
-P(s_{ij1},
-...,s_{ijt},...)
-&=&
-P(s_{ij1})
-...P(s_{ijt})...
-\\
-&=&
-p_{ij}^{x_{ij}}q_{ij}^{y_{ij}}r_{ij}^{z_{ij}}
-\nonumber
-\\
-&=&
-v_{ij}^{x_{ij}}(1-v_{ij})^{z_{ij}}
-q_{ij}^{y_{ij}}(1-q_{ij})^{x_{ij}+z_{ij}}
-\nonumber
-\end{eqnarray}
-where $p_{ij}:=P(S_{ijt}=1)$, $q_{ij}:=P(S_{ijt}=0)$ and $r_{ij}:=P(S_{ijt}=-1)$, and $x_{ij}=\sum_t \delta_{s_{ijt},1}$, $y_{ij}=\sum_t \delta_{s_{ijt},0}$ and $z_{ij}=\sum_t \delta_{s_{ijt},-1}$ are the number of wins, draws and losses of $i$ over $j$ for all $t$, respectively.
-Moreover, $v_{ij}:=P(S_{ijt}=1|S_{ijt}\neq 0)$ and $1-v_{ij}:=P(S_{ijt}=-1|S_{ijt}\neq 0)$ since $p_{ij}=v_{ij}(1-q_{ij})$ and $r_{ij}=(1-v_{ij})(1-q_{ij})$.
-There are different ways to fit the probabilities of Eq.~\ref{eq:v_and_q} to given data.
-For instance, in the maximum a posteriori bayesian approach of the multinomial distribution with a uniform prior, the expected value of the probabilities are $p_{ij}=(x_{ij}+1)/(x_{ij}+y_{ij}+z_{ij}+3)$, $q_{ij}=(y_{ij}+1)/(x_{ij}+y_{ij}+z_{ij}+3)$ and $r_{ij}=(z_{ij}+1)/(x_{ij}+y_{ij}+z_{ij}+3)$ and, therefore $v_{ij}=p_{ij}/(1-q_{ij})=(x_{ij}+1)/(x_{ij}+z_{ij}+2)$.
-In this way, pairs of items with few comparisons can be handled appropriately.
-In particular, if the kind of comparisons under consideration admit no draws, then $q_{ij}=0$ for all $ij$.
-
-\subsection{
-A rating method from least-squares
-\label{sec:ratings_and_rankings_lsqr}
-}
-
-Consider the parameterization of the probabilities $v_{ij}$ in Eq.~\ref{eq:v_and_q} using the logistic function $v = (1 + e^f)^{-1}$. 
-This defines a monotonically decreasing bijection between the probabilities $v_{ij} \in (0, 1)$ and the corresponding values $f_{ij} = \ln(v_{ij}^{-1} - 1) \in (-\infty, \infty)$, and motivates the following question.
-{\em Is there a vector $w \in \mathbb{R}^N$ with components $w_i$ such that Eq.~\ref{eq:motivating_exact} is satisfied?}
-\begin{equation}
-\label{eq:motivating_exact}
-f_{ij}=w_j-w_i
-\end{equation}
-If such a vector exists, it could serve as a rating system, where $v_{ij} > 1/2$ if and only if $f_{ij} < 0$, or equivalently, if and only if $w_j < w_i$. 
-Additionally, this vector could be used to derive a ranking system, where a permutation $r_0, \dots, r_n$ of $0, 1, \dots, n$ satisfying $w_{r_0} \leq w_{r_1} \leq \dots \leq w_{r_n}$, yields a ranking $r_0, r_1, \dots, r_n$ of the items.
-However, such a vector $w$ does not always exist. 
-For example, a dataset might exhibit cyclic behavior, where item $i$ defeats $j$, $j$ defeats $k$, and $k$ defeats $i$, leading to inconsistencies in the ranking. 
-Cyclical inconsistencies are common in practice.
-Fortunatelly, a weakened version of Eq.~\ref{eq:motivating_exact} can still be used to infer useful ratings and rankings. 
-Specifically, the exact condition of Eq.~\ref{eq:motivating_exact} can be replaced by the least-squares problem in Eq.~\ref{eq:motivating_lsqr}~\cite{csato2015graph}.
-\begin{equation}
-\label{eq:motivating_lsqr}
-w = \argmin_{w'}
-\sum_{ij} a_{ij}\,|f_{ij} - (w'_j-w'_i)|^2
-\end{equation}
-The idea here is that, although Eq.~\ref{eq:motivating_exact} cannot always be satisfied, the solution to Eq.~\ref{eq:motivating_lsqr} provides the vector $w$ that best approximates the intended behavior in a least-squares sense.
-
-\subsection{HodgeRank}
-
-In this section the HodgeRank method is reviewed~\cite{jiang2011statistical}.
-For further reading, some practical applications of HodgeRank are demonstrated by Johnson et al.~\cite{johnson2013discrete} and some numerical challenges emerging from its appliation to complex networks are studied by Hirani et al.~\cite{hirani2015graph}.
-Moreover, an elementary review about combinatorial differential topology, geometry and calculus over simplicial complexes is provided in the appendix section, where the formal details are introduced.
-
-An hypergraph is composed of nodes, links and higher order links or hyperlinks such as triangles, tetrahedra and so on.
-A simplicial complex is a specific type of hypergraph that satisfies the property of topological closure, making it a topological space. 
-This structure allows for the definition of cochains, which are combinatorial versions of differential forms where scalar forms corresponds to 0-cochains, vector forms corresponds to 1-cochains and, more generally, $k$-forms corresponds to $k$-cochains.
-In practical terms, a 0-cochain is a vector of scalars $w_i$ associated with nodes $i$, while a 1-cochain is a vector of antisymmetric scalars $f_{ij} = -f_{ji}$ associated with links $ij$. 
-More generally, a $k$-cochain is a vector of alternating scalars $t_{i_0 \dots i_k}$ associated with hyperlinks known as $k$-simplices. 
-Crucially, combinatorial versions of differential operators such as the gradient, the divergence, and the Laplacian, can be defined over simplicial complexes.
-This framework allows the least-squares problem in Eq.~\ref{eq:motivating_lsqr} to be reformulated in terms of the 3-clique complex $K = K_0 \cup K_1 \cup K_2$, derived from the network of adjacency matrix $a$.
-Here, the values $f_{ij}$ are recognized as the components of a 1-cochain $f$, where $f_{ij} = -f_{ji}$ because $v_{ji} = 1 - v_{ij}$, while the differences $w_j - w_i$ correspond to the components of the gradient of a 0-cochain $w$.
-More specifically, Eq.~\ref{eq:motivating_lsqr} can be rewritten as:
-\begin{equation}
-\label{eq:hodge_lsqr}
-w=\argmin_{w'\in C_0^*}|f-d_0(w')|^2
-\end{equation}
-where $w \in C_0^*$ is a 0-cochain, and $d_0 = \grad$ is the gradient operator on the vector space $C_0^*$ of 0-cochains.
-Thanks to this identification, the least-squares problem in Eq.~\ref{eq:hodge_lsqr} can be reformulated in terms of the Hodge decomposition of $f$.
-
-\begin{figure}
-\includegraphics*[scale=.9]{assets/fig1.pdf}
-\caption{
-\label{fig:1}
-(color online).
-An elementary example of a simplicial complex with $N=8$ nodes.
-Its set of 0-simplices or nodes is $K_0=\{\{0\},\{1\},...,\{7\}\}$, of 1-simplices or links is $K_1=\{\{0,1\},\{1,2\},\{0,2\},\{2,3\},\{3,4\},\{4,5\},\{1,5\},\{4,6\},\{6,7\}\}$ and of 2-simplices or triangles is $K_2=\{\{0,1,2\}\}$.
-A local cyclic flow can only exists in the 2-simplex $\{0,1,2\}$ (depicted in green), while non-local cyclic flows can exists in the closed paths $1\to 2\to 3\to 4\to 5\to 1$ and $0\to 2\to 3\to 4\to 5\to 1\to 0$.
-}
-\end{figure}
-
-In the Hodge decomposition of a general $k$-cochain $f \in C_k^*$, there exist unique $k$-cochains $s, g, h \in C_k^*$ such that $f = s \oplus h \oplus g$, where $s\in \rng\,d_{k}^*$, $h\in \krn\,L_k$ and $g\in \rng\,d_{k-1}$ are called the coexact, the harmonic and the exact components of $f$.
-Here, $\rng\,d_{k}^*$, $\krn\,L_k$ and $\rng\,d_{k-1}$ are the ranges and kernel of the $k$-th dual coboundary operator $d_k^*$, the $k$-th Laplacian $L_k$ and the $(k-1)$-th coboundary operator $d_{k-1}$, respectively, and $C_k^*$ is the vector space of $k$-cochains.
-
-To compute the Hodge decomposition of a $k$-chochain $f$, recall that
-$\krn\,d_k=\krn\,L_k\oplus\,\rng\,d_{k-1}$
-and 
-$\krn\,d_{k-1}^*=\rng\,d_k^*\oplus\, \krn\,L_k$.
-In consequence
-\begin{eqnarray}
-\label{eq:lsqr_u}
-d_{k}(f) &=& d_{k}(s + h + g) \\
-&=& d_{k}(s) \nonumber \\
-&=& d_{k}(d_{k}^*(u)) \nonumber \\
-&=& L_{k+1}^{\downarrow}(u) \nonumber \\
-\alpha_{k+1} f &=& \alpha_{k+1} \alpha_{k+1}^T u \nonumber
-\end{eqnarray}
-and
-\begin{eqnarray}
-\label{eq:lsqr_w}
-d_{k-1}^*(f) &=& d_{k-1}^*(s + h + g) \\
-&=& d_{k-1}^*(g) \nonumber \\
-&=& d_{k-1}^*(d_{k-1}(w)) \nonumber \\
-&=& L_{k-1}^{\uparrow}(w) \nonumber \\
-\alpha_{k}^T f &=& \alpha_{k}^T \alpha_{k} w. \nonumber
-\end{eqnarray}
-are singular linear equations for some $u \in C_{k+1}^*$ and $w \in C_{k-1}^*$, that can be solved to obtain $s = \alpha_{k+1}^T u$, $g = \alpha_{k} w$, and $h = f - s - g$.
-Although the Hodge Laplacians 
-$L_{k+1}^{\downarrow}$
-and
-$L_{k-1}^{\uparrow}$
-are symmetric operators, they are also singular and, therefore, Eqs.~\ref{eq:lsqr_u} and \ref{eq:lsqr_w} still pose numerical challenges. 
-Fortunately, these are the normal equations of respective least-squares problems. 
-Hence, their singularities can be addressed by seeking the solutions $u$ and $w$ that also minimize the squared norms $|u|^2$ and $|w|^2$, respectively.
-
-The Hodge decomposition provides a solution to Eq.~\ref{eq:hodge_lsqr} when $k=1$.
-Namely, for $k=1$, any solution $w$ of Eq.~\ref{eq:lsqr_w} is also a solution to Eq.~\ref{eq:hodge_lsqr}, and vice versa. 
-This approach to inferring $w$ from the Hodge decomposition of $f$ is what Jiang et al.~\cite{jiang2011statistical} referred to as {\em HodgeRank}. 
-Additionally, drawing an analogy from physics in 3D, the coexact, harmonic, and exact components $s$, $h$, and $g$, are referred to as the solenoidal, harmonic, and gradient components of $f$, respectively, providing insightful observations in the context of rating inference.
-Namely, the gradient component $g$ represents a non cyclic flow, which aligns perfectly with a rating vector $w$. 
-In contrast, the solenoid and harmonic components, $s$ and $h$, correspond to local and non-local cyclic flows, respectively, which are incompatible with a consistent rating system.
-See Fig.~\ref{fig:1} for an example.
-These cyclic flows were previously discussed in Section~\ref{sec:ratings_and_rankings_lsqr}. 
-The norms $|s|$, $|h|$, and $|g|$ help to gauge the extent to which the inferred ratings are affected by local or non-local cyclic flows.
-
-\begin{figure*}
-\includegraphics*[scale=.4]{assets/fig2a.pdf}
-\includegraphics*[scale=.4]{assets/fig2b.pdf}
-\put(-483,140){\bf a)}
-\put(-240,140){\bf b)}
-\\
-\includegraphics*[scale=.4]{assets/fig2c.pdf}
-\includegraphics*[scale=.4]{assets/fig2d.pdf}
-\put(-483,140){\bf c)}
-\put(-240,140){\bf d)}
-\caption{
-\label{fig:2}
-(color online).
-The mean of the average of the absolute differences of ratings $\avrg{\bar{\tau}}$ averaged over 5000 network samples and realizations of the disorder $\eta$ as a function of $\sigma$ for different sizes $N$ (left subpanels) and values of the parameter $\theta$ (right subpanels) of the corresponding network model.
-Dashed lines represent guides to the eye.
-In panel {\bf a)}, 1D-lattices where $\theta=z$ is the number of neighbors per node.
-In panel {\bf b)}, Erd\H{o}s-R\'enyi random networks where $\theta=\bar{k}$ is the average degree.
-In panel {\bf c)}, Barab\'asi-Albert scale-free networks where $\theta=q$ is the number of new links for each new node of the growing algorithm.
-In panel {\bf d)}, Watts-Strogatz small-world networks of average degree $\bar{k}=4$ where $\theta=p$ is randomization probability per link.
-In the inset, the value of $\avrg{\bar{\tau}}$ for fixed $\sigma$ is plotted as a function of $p$.
-}
-\end{figure*}
-
-In practice, some components $w_i$ of the minimum norm solution to Eq.~\ref{eq:lsqr_w} may be negative. 
-Since a rating system with non-negative ratings is often desirable, the components can be shifted by a constant term, $w_i \to w_i - \min_j w_j$. 
-This shifted rating vector remains a valid solution to Eqs.~\ref{eq:hodge_lsqr} and facilitates the comparison between the inferred $w$ and the true ratings $\hat{w}$ when the later are available.
+Bla bla...
 
@@
 
-In the following numerical experiments, HodgeRank is applied to infer the ratings and rankings of items matched by networks of various topologies: 1D-lattices, Erdős-Rényi (ER) random graphs, Barabási-Albert (BA) scale-free networks, and Watts-Strogatz (WS) small-world networks~\cite{newman2018networks}.
+The following experiments are carried on.
+\begin{enumerate}
+\item 
+Compute ratings and rankings from the full weighted networks.
 
-For each network of adjacency matrix $a$, a vector of true ratings $\hat{w}$ is generated according to $\hat{w}_i = i$ for $i = 0, 1, \dots, n$. 
-With this choice, the expected true ranking is $\hat{r}_i=i$.
-In 1D-lattices and WS networks, nodes are enumerated following the natural order in the lattice and in BA networks nodes are enumerated from the oldest to the newest.
-Using the true rating vector $\hat{w}$, a corresponding 1-cochain $f$ is defined by
-\begin{equation}
-\label{eq:f_and_eta}
-f_{ij}
-=
-\hat{w}_j-\hat{w}_i
-+
-\eta_{ij},
-\end{equation}
-where the different components of the 1-cochain $\eta$ are independent and identically distributed random variables drawn from a Gaussian distribution with zero mean and standard deviation $\sigma$. 
-Equation~\ref{eq:f_and_eta} is similar to Eq.~\ref{eq:motivating_exact}, but includes the contribution of the additive quenched disorder characterized by $\eta$ that models the effect of statistical fluctuations and the cyclic flows captured by the components $s$ and $h$ of $f$.
-When $\sigma = 0$, the inferred ratings are expected to be perfectly aligned with the true ratings, but as $\sigma$ increases, discrepancies are likely to emerge.
+\item 
+Compute ratings and rankings from sparsified versiond of the weighted networks, by randomly removing a varying fraction of links.
+
+\item 
+Compute ratings and rankings from sparsified versiond of the weighted networks, by removing a varying fraction of links sorted in decreasing order of absolute weight.
+\end{enumerate}
 
 \begin{figure*}
-\includegraphics*[scale=.4]{assets/fig3a.pdf}
-\includegraphics*[scale=.4]{assets/fig3b.pdf}
-\put(-483,140){\bf a)}
-\put(-240,140){\bf b)}
+\includegraphics*[scale=.4]{assets/fig1.pdf}
+\includegraphics*[scale=.4]{assets/fig1.pdf}
+%\put(-483,140){\bf a)}
+%\put(-240,140){\bf b)}
 \\
-\includegraphics*[scale=.4]{assets/fig3c.pdf}
-\includegraphics*[scale=.4]{assets/fig3d.pdf}
-\put(-483,140){\bf c)}
-\put(-240,140){\bf d)}
+\includegraphics*[scale=.4]{assets/fig1.pdf}
+\includegraphics*[scale=.4]{assets/fig1.pdf}
+%\put(-483,140){\bf c)}
+%\put(-240,140){\bf d)}
 \caption{
-\label{fig:3}
+\label{fig:1}
 (color online).
-The mean of the average of the absolute differences of rankings $\avrg{\bar{\tau}}$ as a function of $\sigma$ for the same numerical experiments of Fig.~\ref{fig:2}.
-The insets zoom-in certain regions of the main plots.
-Dashed lines represent guides to the eye.
 }
 \end{figure*}
-
-To compare the true ratings $\hat{w}_i$ with the inferred ratings $w_i$, as well as the true rankings $\hat{r}_i$ with the inferred rankings $r_i$, the average absolute differences
-$$
-\tau
-=
-\frac{1}{N}
-%\sum_{j:j\leq i}
-\sum_{i}
-|\hat{w}_i-w_i|
-$$
-and
-$$
-\rho
-=
-\frac{1}{N}
-%\sum_{j:j\leq i}
-\sum_{i}
-|\hat{r}_i-r_i|
-$$
-are used, respectively.
-These metrics are zero when the inferred ratings or rankings match the true ones, and greater than zero otherwise.
-It is important to note that the means $\bar{\tau}(\sigma, a) = \int dw \, P(w | \sigma, a) \, \tau(w)$ and $\bar{\rho}(\sigma, a) = \int dw \, P(w | \sigma, a) \, \rho(w)$, while similar to order parameters, are more accurately described as {\em disorder parameters}, since they become different from zero when disorder kicks in.
-Here, the distribution $P(w | \sigma, a)$ represents the statistics of $w$ obtained from solving Eq.~\ref{eq:hodge_lsqr} while sampling $\eta$ for a given network structure $a$.
-In the experiments, the network-averaged quantities, $\langle \bar{\tau} \rangle (\sigma) = \sum_a P(a | N, \theta) \bar{\tau}(\sigma, a)$ and $\langle \bar{\rho} \rangle (\sigma) = \sum_a P(a | N, \theta) \bar{\rho}(\sigma, a)$, are computed, where $N$ and $\theta$ represent relevant parameters of the network models.
-Specifically, $\theta$ denotes the coordination number $z$ for 1D-lattices, the average degree $\bar{k}$ for ER networks, the number of new links $q$ per new node in BA networks, and the fraction $p$ of randomized links for WS networks with average degree $\bar{k}=4$.
-
-Traditional solvers struggle to provide adequate solutions for Eqs.~\ref{eq:lsqr_u} and~\ref{eq:lsqr_w}. 
-To address this, the recently introduced iterative solver \verb+MINRES-QLP+, developed by Choi et al.~\cite{choi2011minresqlp}, is employed to find the minimum norm solutions to the singular least-squares problems.
-This solver is available in the \verb+Krylov.jl+ package~\cite{montoison2023krylov} for the Julia programming language~\cite{bezanson2017julia}. 
-A Jupyter notebook with the code used for the experiments can be found online~\cite{github_hon_ranking}.
-The performance of \verb+MINRES-QLP+ was evaluated on non-periodic 1D-lattices of varying sizes $N$. 
-For $\sigma = 0$, the solver demonstrates satisfactory numerical accuracy for network sizes up to approximately $N \approx 10^5$. 
-However, to mitigate the impact of numerical errors, subsequent experiments are conducted on networks of considerably smaller sizes.
-
-The following figures contain four panels, each corresponding to a different network model.
-In each figure, panel {\bf a)} presents results for non-periodic 1D-lattices, panel {\bf b)} for ER networks, panel {\bf c)} for BA networks and panel {\bf d)} for WS networks.
-Furthermore, each panel is divided into two subpanels.
-The left subpanel presents results for $\theta$ fixed and varying network sizes $N$, and the right subpanel for fixed $N$ and varying $\theta$.
-The results are averaged over 5000 samples of both $a$ (the network structure) and $\eta$ (the disorder).
-Where possible, a phenomenologically obtained approximate scaling relation for the studied quantities is indicated.
-
-In Fig.~\ref{fig:2}, the mean of the average absolute differences, $\avrg{\bar{\tau}}$, between true and inferred ratings is plotted as a function of $\sigma$. 
-For small $\sigma$, an approximately linear scaling of $\avrg{\bar{\tau}}$ with $\sigma$ is observed in most cases. 
-However, deviations from this behavior emerge and are more pronounced in networks with smaller sizes $N$ or average degrees $\bar{k}$, specially in BA networks.
-As shown in the left subpanels, $\avrg{\bar{\tau}}$ scales approximately as $\sim N^{1/2}$ for 1D-lattices, while for all other network models, it appears largely independent of $N$. 
-In the right subpanels, it is observed for 1D-lattices that $\avrg{\bar{\tau}}$ scales as $\sim z^{-1.3}$ for sufficiently large values of $z$. 
-Similarly, it scales as $\sim \bar{k}^{-2/3}$ for ER networks and $\sim q^{-2/3}$ for BA networks. 
-In all cases, deviations from these scalings are observed when networks approach a sparse tree-like structure.
-For WS networks, $\avrg{\bar{\tau}}$ decreases with $\ln p$ in a sigmoid-like manner, as shown in the inset of the right subpanel of panel {\bf d)}, where the transition point is found between the small-world and the lattice regimes.
-
-In Fig.~\ref{fig:3}, the mean of the average absolute differences, $\avrg{\bar{\rho}}$, between true and inferred rankings is shown as a function of $\sigma$. 
-Interestingly, as it can be seen in more detail in the insets, two distinct regimes are observed.
-For small values of $\sigma$, there is a region of perfect retrieval of the true ranking, where $\avrg{\bar{\rho}} = 0$. 
-This behavior persists up to a non-trivial threshold, $\sigma^*$, after which $\avrg{\bar{\rho}}$ deviates from zero increasing monotonically with $\sigma$.
-As seen in the right subpanels of panels {\bf a)}, {\bf b)}, and {\bf c)}, a transition region emerges in networks with a sufficiently large connectivity, $\bar{k}$. 
-In this region, $\avrg{\bar{\rho}}$ shifts from perfect retrieval to a regime where it grows approximately linearly with $\sigma$.
-The distinction between these two regimes becomes more pronounced with increasing disorder in the network structure, as illustrated by the WS networks in the right subpanel of panel {\bf d)} as $p$ grows.
-
-To identify a transition point $\sigma_c$ within a transition region, the primitive
-$$
-\avrg{\bar{\rho}}(\sigma)
-=
-\frac{A}{B}
-\ln
-\big(
-1+e^{B(\sigma-\sigma_c)}
-\big)
-$$
-of the sigmoid function
-$$
-\avrg{\bar{\rho}}'(\sigma)
-=
-\frac{A}{1+e^{-B(\sigma-\sigma_c)}}
-$$
-is fitted to the curves of Fig.~\ref{fig:3}.
-Here, $A$, $B$ and $\sigma_c$ are fitting parameters.
-The primitive function approaches linear growth, $\avrg{\bar{\rho}}(\sigma) \approx A(\sigma - \sigma_c)$, for $\sigma \gg \sigma_c$, and approaches zero for $\sigma \ll \sigma_c$. 
-The parameter $B^{-1}$ controls the width of the transition region.
-The peak of the derivative of the sigmoid function
-$$
-\avrg{\bar{\rho}}''(\sigma)
-=
-\frac{ABe^{-B(\sigma-\sigma_c)}}{(1+e^{-B(\sigma-\sigma_c)})^2}
-$$
-determines the transition point $\sigma_c$. 
-The range $[\sigma^*,\sigma^{**}]$ of fitted values for each curve is determined by maximizing the height of the peak, $\avrg{\tilde{\rho}}:=\avrg{\bar{\rho}}''(\sigma_c) = \frac{AB}{4}$, as a function $\sigma^{**}$.
-Further details of the fitting procedure can be found in Figs.~1-8 and related text of the Supplemental Material (SM) at~[URL will be inserted by publisher].
-The results of these fits are shown in Fig.~\ref{fig:4}, where the transition points $\sigma_c$ and peak heights $\avrg{\tilde{\rho}}$ are plotted as functions of the networks' parameters $N$ and $\theta$.
-As seen in the left subpanels, the dependency of these quantities on the network size $N$ is negligible or relatively weak for all network models. 
-In contrast, the right subpanels reveal different power-law scalings between these quantities and $\theta$ for the different networks models and $N$ fixed.
-More specifically, $\sigma_c$ grows with $z$, $\bar{k}$ and $q$ for 1D-lattices, ER and BA networks, respectively.
-In particular, the scaling law has a logarithmic correction for ER networks.
-On the other hand, the peak heigth $\avrg{\tilde{\rho}}$ decreases with $z$, $\bar{k}$ and $q$.
-For WS networks, both $\sigma_c$ and $\avrg{\tilde{\rho}}$ remain approximately constant in the small-world regime, but undergo a sharp change once the lattice regime is reached.
-
-\begin{figure*}
-\includegraphics*[scale=.4]{assets/fig4a.pdf}
-\includegraphics*[scale=.4]{assets/fig4b.pdf}
-\put(-483,140){\bf a)}
-\put(-240,140){\bf b)}
-\\
-\includegraphics*[scale=.4]{assets/fig4c.pdf}
-\includegraphics*[scale=.4]{assets/fig4d.pdf}
-\put(-483,140){\bf c)}
-\put(-240,140){\bf d)}
-\caption{
-\label{fig:4}
-(color online).
-Transition points $\sigma_c$ (green circles) and peak heights $\langle \tilde{\rho} \rangle$ (pink stars) obtained from fits of the primitives of sigmoid functions to the curves $\langle \bar{\rho} \rangle$ vs.
-$\sigma$ in Fig.~\ref{fig:3}.
-Error bars are calculated through error propagation of the errors of the fitted parameters.
-The dashed lines represent phenomenological scaling relations fitted to the data points.
-}
-\end{figure*}
-
-\section{
-Discussion
-\label{sec:discussion}
-}
-
-The scaling of $\avrg{\bar{\tau}}$ vs $N$ indicated in Figs.~\ref{fig:2} for the 1D-lattice differs from the observed in the other network models.
-This suggests that the non-local cycles existing in the small-world regime affects the inference of ratings.
-The observation is consistent with the behavior of $\langle \bar{\tau} \rangle$ with $\ln p$ in the WS model.
-Concerning the scaling of $\langle \bar{\tau} \rangle$ with $z$, $\bar{k}$, and $q$, deviations from such scalings become evident as the networks approach a sparse tree-like regime, where the lack of redundant paths weakens the reinforcement of rating differences between arbitrary items.
-
-The results suggest that the transition from a regime of perfect retrieval $\avrg{\bar{\rho}}=0$ to a regime of partial retrieval $\avrg{\bar{\rho}}>0$ of the true ratings observed in Fig.~\ref{fig:3}, is a consequence of the progressive degradation of the rating differences as the strength $\sigma$ of the quenched dissorder increases.
-Namely, this interpretation is supported by the fact that $\avrg{\bar{\tau}}$, $|h|$ and $|s|$ grow almost linearly with $\sigma$ as shown in Fig.~\ref{fig:2} and Figs.~10~and~11 of the SM.
-For sufficiently large connectivities, the transition region remains bounded and is approximately defined by $[\sigma^*, \sigma^{**}]$, with a transition point $\sigma_c$ identified by the peak of the primitive of a sigmoid function.
-As connectivity decreases, or as the small-world property weakens with decreasing $p$, the transition region becomes less definite, emphasizing the critical role that redundant non-local paths have in establishing a well-defined ranking.
-Finally, Fig.~\ref{fig:4} shows that the properties of the transition are weakly dependent on network size but strongly influenced by the characteristics of the network model and the values of its other parameters.
-This observation is further supported by the scaling of the functions defined by $|f|$, $|s|$, $|h|$ and $|g|$ versus $\sigma$, across the different network models, shown for varying $N$ with fixed $\theta$ and varying $\theta$ with fixed $N$ in Figs.~9-12 of the SM.
-Additionally, for the sake of completion, scaling relations for the average number of dimensions $\avrg{\kappa_k}$ of the vector spaces $C_k$ of $k$-chains are also shown in Fig.~13 of the SM, and are consistent with the observed behavior for $|s|$ and $|h|$.
 
@@
 }
-
-This work explores the impact of complex network topologies on the performance of {\em HodgeRank}, a method developed to infer ratings from pairwise comparisons using a combinatorial formulation of differential geometry and Hodge theory over Higher Order Networks.
-The study presents a comprehensive scaling analysis of a variety of metrics, using well-established network models: 1D-lattices, Erdős-Rényi random graphs, Barabási-Albert scale-free networks, and Watts-Strogatz small-world networks.
-Additionally, the analysis incorporates the effects of quenched disorder in the data, which negatively influences the inference of ratings and rankings.
-The numerical experiments demonstrate that as the strength of disorder varies, a transition occurs from a regime of perfect retrieval to one of partial retrieval of the true rankings.
-The scaling laws identified for the different studied quantities, highlight the critical role of the network structure.
-This variability encompasses contributions from both the gradient and cyclic flows of the Hodge decomposition that form the foundation of the HodgeRank method.
-
-Future work could apply more traditional methods from statistical physics to further investigate the properties of the mentioned transition and scaling relations. 
-Additionally, this approach could be extended to study the inference of ratings and rankings in empirical data, potentially leading to the development of more realistic models of the network topology and the quenched disorder.
-Moreover, future work could study how the quenched disorder affects the inferred ratings and rankings of the individual nodes depending on their local properties.
-Finally, using the benchmark experiments here developed, it would be interesting to investigate how HodgeRank compares with alternative methods for the inference of ratings and rankings over a broader range of network models, using modern computer architectures like GPUs or over very large datasets.
 
@@
 
-\appendix*
-\onecolumngrid
+%\appendix*
+%\onecolumngrid
 
-\section{\label{appA}Combinatorial differential topology, geometry and calculus}
-
-This section provides an introductory review about combinatorial differential topology, geometry and calculus over simplicial complexes. 
-Where appropriate, analogies to the smooth standard version of differential geometry over manifolds are highlighted. 
-However, it is important to remark that, despite these similarities, significant differences exits between the combinatorial and smooth frameworks~\cite{ribando2023combinatorial}.
-
-\subsection{Notation}
-
-If  $\{e_i:i\in I\}$ is a basis of a vector space $V$ over a field $\FF$, then the linear decomposition of a vector $v\in V$ is denoted by $v=v^ie_i$ where Einstein's summation convention is assumed and $v^i\in \FF$.
-Additionally, if $f\in V^*$ is a covector of the dual vector space of $V$, then $f=f_ie^i$ where $\{e^i:i\in I\}$ is the corresponding dual basis, $f_i\in \FF$ is a coordinate and $e^i(e_j)=\delta^i{}_j$ is Kronecker's delta.
-If $a:V\to V$ is a linear operator, then there is a matrix of components $a_i{}^j$ such that $a(e_i)=a_i{}^je_j$.
-In this way, $w^je_j=w=a(v)=v^ia(e_i)=v^ia_i{}^je_j$, so $w^j=v^ia_i{}^j=(a^T)^j{}_iv^i$ where $(a^T)^j{}_i$ are the components of the transpose of the matrix of components $a_i{}^j$~\footnote{The first index $i$ of the components $a_{ij}$, $a^{ij}$, $a_i{}^j$ and $a^i{}_j$ of a matrix indicates the row of the entry, while the second $j$ the column.
-There is no predefined relation between the left/right and the upper/lower properties of an index.}.
-The kernel of $a$ is denoted by $\krn\,a=\{v\in V:a(v)=0\}$, the range is denoted by $\rng\,a=a(V)=\{a(v):v\in V\}$, the null is denoted by $\nll\,a=\dim\,\krn\,a$ and the rank is denoted by $\rnk\,a=\dim\,\rng\,a$.
-An inner product over $V$ maps any two vectors $u,v\in V$ into a scalar $u.v\in \FF$.
-It is linear in the first argument if $\FF=\RR$ or antilinear if $\FF=\CC$, and always linear in the second argument.
-Moreover, $v.v\geq 0$ for all $v\in V$, so $|v|:=\sqrt{v.v}$ is a norm over $V$ and, in particular, $v.v=0$ implies $v=0$ whenever the inner product is non-degenerate.
-
-\subsection{Simplicial complexes}
-
-An abstract simplicial complex is a set $K$ of subsets of $\{0,1,...,n\}$ that satisfies the property of topological closure: i.e. if $s\in K$ and $r\subseteq s$ is non-empty, then $r\in K$.
-The elements $i$ of $\{0,1,...,n\}$ are called nodes.
-The elements of $K$ are called simplices.
-For each simplex $s\in K$, there exist unique $i_0,...,i_k$ in $\{0,1,...,n\}$ such that $s=\{i_0,i_1,...,i_k\}$ and $i_0<i_1<...<i_k$.
-Here, $\dim s=k=|s|-1$ is called the dimension of the simplex $s$ and $\dim K := \max_{s\in K} \dim\, s$ is the dimension of the complex $K$.
-Correspondingly, $s$ is called a $k$-simplex and $K$ a $(\dim K)$-complex.
-$K_k$ denotes the set of all $k$-simplices in $K$ and its cardinality is $\kappa_k:=|K_k|$, while $\kappa:=\sum_k\kappa_k=|K|$ is the cardinality of the complex $K$.
-In particular, $K_{-1}=K_{n+1}=\emptyset$ are defined for convenience.
-For any different nodes $i,j,k,l \in \{0,1,...,n\}$, 0-simplices like $\{i\}$ are in one to one correspondence with nodes, 1-simplices like $\{i,j\}$ can be thought of links, 2-simplices like $\{i,j,k\}$ as triangles, 3-simplices like $\{i,j,k,l\}$ as tetrahedra, and so on.
-In particular, a simplicial complex of dimension $\dim K=1$ is equivalent to a non-directed graph without self-links.
-
-A simplex $r$ is a face of another $s$ if $r\subset s$.
-There are ${k + 1 \choose q + 1}$ $q$-faces within a $k$-simplex.
-Moreover, $r$ is a facet of $s$ if $r\subset s$ and $1+\dim\,r=\dim\,s$.
-Correspondingly, $s$ is a coface/cofacet of $s$ if $r$ is a face/facet of $s$, respectively.
-Additionally, $r$ and $s$ are incident to each other if $r$ is a facet of $s$ or $s$ is a facet of $r$.
-Furthermore, $r$ and $s$ are lower-adjacent to each other if they have a facet in common, and are upper-adjacent to each other if they are both different facets of a common simplex.
-In particular, two vertices of a common graph are upper-adjacent if they are both different facets of a common edge.
-The degree $k_s$ of a simplex $s$ is its number of cofacets.
-The degree $k_s$ of a simplex $s$ should not be confused with the dimension $k$ of a $k$-simplex.
-The generalized degree $\zeta_{sq}$ of a $k$-simplex $s$ is its number of $q$-cofaces.
-In the particular case of common graphs or networks, $k_i=\zeta_{i1}$ for any node $i$.
-
-A simplicial complex $K$ can be obtained from a graph $g$ where each $(k+1)$-clique of $g$ corresponds to a $k$-simplex of $K$.
-Such simplicial complex obtained from $g$ is called the {\em clique complex} of $g$.
-In particular, the $(k+1)$-clique complex of a graph $g$ is the simplicial complex obtained by considering only the $(q+1)$-cliques of $g$ for $q\leq k$.
-
-\subsection{Chains}
-
-The set of functions from $K$ to some number field $\FF$ is a vector space $C$ of vectors called chains.
-The functions $e_s : K\to \FF$ for $s\in K$ such that $e_s(r)=\delta_{sr}$ form a canonical basis of $C$~\footnote{Here, $\delta_{sr}$ is the Kronecker delta.}.
-In particular, if $s$ is a $k$-simplex, then $e_s$ is called a $k$-chain.
-The subset of linear combinations of $k$-chains form a subspace $C_k$ of $C$.
-Crucially,
-$$
-C=C_0\oplus C_1\oplus ... \oplus C_n
-$$
-where $C_0=C_n\cong\FF$, i.e. $C_0$ and $C_n$ are isomorphic to $\FF$.
-For convenience, $C_{-1}=C_{n+1}=\{0\}$ are also defined.
-Note, $m_k=\dim\, C_k\leq {n\choose k}$ for $k=0,1,...,n$, and $\dim\, C_{-1}=\dim\, C_{n+1}=0$. 
-Therefore, $\dim\,C\leq \sum_{k=0}^n {n\choose k}=2^n$.
-Using Einstein's notation, any chain of $C$ can be represented by a linear combination $c=c^se_s$ with unique coefficients $c^s\in \FF$.
-
-\subsection{Cochains}
-
-The dual space $C^*$ of $C$ is called the vector space of cochains of $K$.
-If $C_k^*$ is the dual space of $C_k$, then
-$$
-C^*=C_0^*\oplus C_1^*\oplus ... \oplus C_n^*
-$$
-For convenience, $C_{-1}^*=C_{n+1}^*=\{0\}$ are also defined.
-The covectors in $C_k^*$ are called $k$-cochains.
-In particular, $\{e^s:s\in K\}$ denotes the basis of $C^*$ dual to the canonical basis $\{e_s:s\in K\}$ of $C$.
-As a consequence, $e^s(e_r)=\delta^s{}_r$ for all $s,r\in K$.
-In this way, any cochain $f\in C^*$ can be represented by a linear combination $f=f_se^s$ of coefficients $f_s\in \mathbb{F}$.
-
-The application of a cochain $f$ to a chain $c$ equals $f(c)=f_se^s(c^re_r)=f_sc^re^s(e_r)=f_sc^r\delta^s{}_r=f_sc^s\in \FF$.
-In a sense, $f(c)$ is the combinatorial analogous of the integration $\int_c f$ of form $f$ over a manifold $c$.
-It should be remarked, however, that a simplicial complex is not always obtained from the discretization of a manifold.
-This is the reason for which, in this work, the term {\em combinatorial} is preferred over the term {\em discrete}.
-
-\subsection{Orientation}
-
-Let $e_{i_0i_1...i_k}:=e_{\{i_0,i_1,...,i_k\}}$ when $0\leq i_0<i_1<...<i_k\leq n$.
-Otherwise, let 
-$$
-e_{j_0j_1...j_k}
-=
-\left\{
-\begin{array}{rl}
-e_{i_0i_1...i_k} & \mbox{if $j_0j_1...j_k$ is an even permutation of $i_0i_1...i_k$,} \\
--e_{i_0i_1...i_k} & \mbox{if it is an odd permutation, and} \\
-0 & \mbox{otheriwse.}
-\end{array}
-\right.
-$$
-The sign between $e_{j_0j_1...j_k}$ and $e_{i_0i_1...i_k}$ is called the orientation of $e_{j_0j_1...j_k}$ relative to $e_{i_0i_1...i_k}$.
-Like the orientation of the axis of $\RR^n$, the orientation of the simplices is a matter of convention, which is consisten with the arbitrariness in the enumeration of the nodes.
-
-\subsection{
-Boundaries
-\label{sec:boundaries}
-}
-
-The boundary operator over chains $\partial : C\to C$ is defined by
-$$
-\partial = \partial_0+\partial_1+...+\partial_n
-$$
-where $\partial_{-1}=\partial_0=\partial_{n+1}=0$, and
-$$
-%\partial_k(e_s)
-\partial_k(e_{i_0i_1...i_q})
-=
-\delta_{qk}
-\sum_{j=0}^q
-(-1)^j 
-e_{i_0i_1...\hat{i}_j...i_k}
-$$
-for $0<k\leq n$, where $e_{i_0i_1...\hat{i}_j...i_k}:=e_{i_0i_1...i_{j-1}i_{j+1}...i_k}$.
-Here, the sign of $(-1)^j$ is called the orientation that the $k$-chain $e_{i_0i_1...i_k}$ induces on the $(k-1)$-chain $e_{i_0i_1...\hat{i}_j...i_k}$.
-
-The boundary operator satisfies the following crucial properties.
-Firstly, $\partial$ is a linear operator.
-Secondly, $\partial_k(C)=\partial(C_k)\subseteq C_{k-1}$, meaning that the boundary of a simplex of dimension $k$ has dimension $k-1$.
-Thirdly, $\partial^2=0$, meaning that a boundary has no boundary.
-This third statement follows from
-$$
-\partial^2 
-=
-\sum_{jk}\,
-\partial_j \partial_k
-\nonumber
-\\
-=
-\sum_{k}\,
-\partial_{k}\partial_{k+1}
-$$
-and the fact that $\partial_{k}\partial_{k+1}=0$ follows from the alternating definition of $\partial_k$ and that chains are oriented.
-
-In matricial form 
-$\partial_k(e_s)=(\alpha_k)_s{}^r e_r$.
-In this way, $q^re_r=q=\partial_k(c)=\partial_k(c^se_s)
-=c^s\partial_k(e_s)=c^s(\alpha_k)_s{}^re_r$.
-Hence, $q^r=c^s(\alpha_k)_s{}^r=(\alpha_k^T)^r{}_sc^s$.
-Note, $(\alpha_k)_s{}^r\neq 0$ if and only if $s$ and $r$ are incident to each other.
-Hence, $\alpha_k\in \RR^{m_{k}\times m_{k-1}}$ or its transpose $\alpha_k^T$ are usually called the $k$-th incidence matrix of the simplicial complex~\cite{grady2010discrete,newman2018networks,bianconi2021higher}.
-In this work, the convention where the incidence matrix is the transpose $\alpha_k^T$ is adopted.
-
-\subsection{Coboundaries}
-
-The coboundary operator $d:C^*\to C^*$ is the combinatorial analogous of the differential operator.
-It is defined by
-$$
-(d(f))(c) = f(\partial(c))    
-$$
-for all $f\in C^*$ and $c\in C$, which can be recognized as the combinatorial analogous of the generalized Stoke's theorem
-$$
-\int_c df = \int_{\partial c} f
-$$
-which in turn is the generalization of the fundamental theorem of calculus to arbitrary dimensions.
-Crucially, 
-$$
-d=d_0+d_1+...+d_n
-$$
-where $d_{-1}=d_n=d_{n+1}=0$,  $d_k(C^*)=d(C_k^*)\subseteq C_{k+1}^*$,
-and $(d_k(f))(c)=f(\partial_{k+1}(c))$ for all $c\in C$.
-Moreover, $d^2=0$ and $d_k d_{k-1}=0$ easily follows from $\partial^2=0$.
-This fact is the reason for which many differential equations in physics are of second order.
-
-In matricial form 
-$d_k(e^s)=(\beta_k)^s{}_r e^r$.
-Hence, 
-\begin{eqnarray}
-(d_k(e^s))(e_r)
-&=&
-e^s(\partial_{k+1}(e_r))
-\nonumber
-\\
-((\beta_k)^s{}_te^t)(e_r)
-&=&
-e^s((\alpha_{k+1})_r{}^le_l)
-\nonumber
-\\
-(\beta_k)^s{}_t\,e^t(e_r)
-&=&
-(\alpha_{k+1})_r{}^l\,e^s(e_l)
-\nonumber
-\\
-(\beta_k)^s{}_t\,\delta^t{}_r
-&=&
-(\alpha_{k+1})_r{}^l\,\delta^s{}_l
-\nonumber
-\\
-(\beta_k)^s{}_r
-&=&
-(\alpha_{k+1})_r{}^s
-\nonumber
-\end{eqnarray}
-In other words, $\beta_k=(\alpha_{k+1}^T)$.
-Note, $g_re^r=g=d_k(f)=d_k(f_se^s)=f_sd_k(e^s)
-=f_s(\beta_k)^s{}_re^r$.
-Hence $g_r=f_s(\beta_k)^s{}_r=(\beta_k^T)_r{}^sf_s=(\alpha_{k+1})_r{}^sf_s$.
-
-Note, $\grad = d_0:C_0^*\to C_1^*$ and 
-$\curl = d_1:C_1^*\to C_2^*$ are recognized as a combinatorial analogous of the gradient and the curl operators of standard differential calculus~\footnote{Strictly speaking, in standard differential calculus, the gradient operator is defined in a way that is metric dependent.}, respectively.
-In particular, the known result $\curl\circ\grad=-d_1 d_0=0$ follows as a particular case of $d_k d_{k-1}=0$.
-As an example on how these operators work, note that if $f\in C_0^*$ and $g=\grad\, f=d_0(f)$, then $g_{ij}=g(e_{ij})=(d_0(f))(e_{ij})=f(\partial_1(e_{ij}))=f(e_j-e_i)=f_j-f_i$.
-
-\subsection{Inner products and geometry}
-
-Consider a positive definite inner product from $C$ to $\RR$.
-By the Riesz's representation theorem, a positive definite inner product over $C$ induces an isomorphism $C\ni c\to c^{\flat}\in C^*$ called {\em flat} defined by
-$$
-c^{\flat}(q) = c.q
-$$
-for all $c,q\in C$.
-Its inverse is denoted by $C^*\ni f\to f_{\sharp}\in C$ and is called {\em sharp}.
-Together, $\flat$ and $\sharp$ are called musical isomorphisms.
-They induce a positive definite inner product over $C^*$ given by $f.g = f_{\sharp}.g_{\sharp}$ for all $f,g\in C^*$.
-
-Consider a non-degenerate inner product over $C$.
-Its coordinates in the canonical basis are $m_{sr}=e_s.e_r=e_r.e_s=m_{rs}$ for $s,r\in K$.
-Since $e_s^{\flat}\in C^*$, there exists $t_{sr}\in \RR$ such that $e_s^{\flat}=t_{sr}e^r$ and, therefore,
-$$
-m_{sq} = e_s.e_q = e_s^{\flat}(e_q)=t_{sr}e^r(e_q)=t_{sr}\delta^r{}_q=t_{sq}
-$$
-In consequence
-$$
-m_{sr}=e_s.e_r=e_s^{\flat}.e_r^{\flat}=(m_{st}e^t).(m_{rq}e^q)=m_{st}m_{rq}(e^t.e^q)=m_{st}m_{rq}m^{tq}
-$$
-so $0=m_{st}(\delta^t{}_r-m_{rq}m^{tq})$ and, as a result, $m_{st}m^{tr}=\delta_s{}^r$.
-In other words, the matrices of components $m_{st}=e_s.e_t$ and $m^{tr}=e^t.e^r$ are the inverse of each other.
-In the context of abstract simplicial complexes, it is often assumed that the canonical basis $\{e_s:s\in K\}$ of $C$ is orthogonal, so $m_{rs}$ and $m^{rs}$ are the components of diagonal matrices.
-Following this trend, in the present work, the trivial inner product $m_{rs}=\delta_{rs}$ is considered.
-
-In differential geometry, an inner product over differencial forms is usually defined in terms of a volume form and an associated Hodge's star operator which in turn depend on a metric tensor.
-When a simplicial complex is derived from the discretization of manifold, these constructions can be also available, highlighting the intimate relation between the inner product over chains or cochains, and the geometric properties of simplicial complexes.
-In some cases, an hermitian inner product over $\CC$ is preferred.
-When this is the case, it can be incorporated into the formalism by the complexification of $C$~\cite{roman2005advanced}.
-
-\subsection{Dual boundaries}
-
-The dual boundary operator $\partial^*:C\to C$ is defined as the inner-product adjoint of the boundary operator $\partial$, i.e. by
-$$
-\partial^*(c).q = c.\partial(q)
-$$
-for all $c,q\in C$.
-Crucially, $(\partial^*)^2=0$ follows from $\partial^2=0$, and
-$$
-\partial^*=\partial_0^*+\partial_1^*+...+\partial_n^*
-$$
-where  $\partial_{-1}^*=\partial_{0}^*=\partial_{n+1}^*=0$ and $\partial_k^*$ is the dual of $\partial_k$, so $\partial_k^*(C)=\partial^*(C_{k-1})\subseteq C_k$.
-
-In matricial form 
-$\partial_k^*(e_s) = (\mu_k)_s{}^r e_r$.
-Hence,
-\begin{eqnarray}
-\partial_k^*(e_s).e_r
-&=&
-e_s.\partial_k(e_r)
-\nonumber
-\\
-(\mu_k)_s{}^t\, e_t.e_r
-&=&
-(\alpha_k)_r{}^p\, e_s.e_p
-\nonumber
-\\
-(\mu_k)_s{}^t\, m_{tr}
-&=&
-(\alpha_k)_r{}^p\, m_{sp}
-\nonumber
-\\
-(\mu_k)_s{}^t\, m_{tr}m^{rl}
-&=&
-(\alpha_k)_r{}^p\, m_{sp}m^{rl}
-\nonumber
-\\
-(\mu_k)_s{}^t\, \delta_t{}^l
-&=&
-(\alpha_k)_r{}^p\, m_{sp}m^{rl}
-\nonumber
-\\
-(\mu_k)_s{}^l
-&=&
-m^{lr}\,(\alpha_k)_r{}^p\, m_{ps}
-\nonumber
-\end{eqnarray}
-Moreover, in the particular case where $m_{sr}=\delta_{sr}$, so $m^{sr}=\delta^{sr}$, it is found that $(\mu_k)_s{}^t\delta_{tr}=(\alpha_k)_r{}^p\, \delta_{sp}$, so $(\mu_k)_s{}^r=(\alpha_k)_r{}^s$ and, therefore, $\mu_k=\alpha_k^T$.
-Note, $q^re_r=q=\partial_k^*(c)=\partial_k^*(c^se_s)
-=c^s\partial_k^*(e_s)=c^s(\mu_k)_s{}^re_r$.
-Hence, $q^r=c^s(\mu_k)_s{}^r=(\mu_k^T)^r{}_sc^s=(\alpha_k)^r{}_sc^s$.
-
-\subsection{Dual coboundaries}
-
-The dual coboundary operator $d^*:C^*\to C^*$ is defined as the inner-product adjoint of the coboundary operator $d$, i.e. by
-$$
-d^*(f).g = f.d(g)
-$$
-for all $f,g\in C^*$.
-The dual coboundary operator is the combinatorial analogous of the codifferential operator.
-As before, $(d^*)^2=0$ follows from $d^2=0$, and
-$$
-d^*=d_0^*+d_1^*+...+d_n^*
-$$
-where $d_{-1}^*=d_{n}^*=d_{n+1}^*=0$ and $d_k^*$ is the dual of $d_k$, so $d_k^*(C^*)=d^*(C_{k+1}^*)\subseteq C_k^*$.
-
-In matricial form
-$d_k^*(e^s)=(\nu_k)^s{}_r e^r$.
-Hence
-\begin{eqnarray}
-d_k^*(e^s).e^r
-&=&
-e^s.d_k(e^r)
-\nonumber
-\\
-(\nu_k)^s{}_t\,e^t.e^r
-&=&
-(\beta_k)^r{}_p\,e^s.e^p
-\nonumber
-\\
-(\nu_k)^s{}_t\,m^{tr}
-&=&
-(\beta_k)^r{}_p\,m^{sp}
-\nonumber
-\\
-(\nu_k)^s{}_t\,m^{tr}m_{rl}
-&=&
-(\beta_k)^r{}_p\,m^{sp}m_{rl}
-\nonumber
-\\
-(\nu_k)^s{}_t\,\delta^t{}_l
-&=&
-(\beta_k)^r{}_p\,m^{sp}m_{rl}
-\nonumber
-\\
-(\nu_k)^s{}_l
-&=&
-m_{lr}\,(\beta_k)^r{}_p\,m^{ps}
-\nonumber
-\end{eqnarray}
-Moreover, in the particular case where $m^{sr}=\delta^{sr}$, so $m_{sr}=\delta_{sr}$, it is found that
-$(\nu_k)^s{}_t\delta^{tr}=(\beta_k)^r{}_p\delta^{sp}$, so $(\nu_k)^s{}_r=(\beta_k)^r{}_s$ and, therefore, $\nu_k=\beta_k^T=\alpha_{k+1}$.
-Note, $g_re^r=g=d_k^*(f)=d_k^*(f_se^s)
-=f_sd_k^*(e^s)=f_s(\nu_k)^s{}_re^r$.
-Hence, $g_r=f_s(\nu_k)^s{}_r=(\nu_k^T)_r{}^sf_s=(\beta_k)_r{}^sf_s=(\alpha_{k+1}^T)_r{}^sf_s$.
-
-Note, $\dive = -d_0^*:C_1^*\to C_0^*$ can be recognized as the combinatorial analogous of the divergence of standard differential calculus.
-
-\subsection{Hodge decomposition}
-
-Consider vector spaces $U,V$ and $W$ equipped with an inner product and two linear operator $a:V\to W$ and $b:U\to V$ such that 
-\begin{equation}
-\label{eqA1}
-ab=0
-\end{equation}
-Then $\rng(b)\subseteq \krn(a)$ follows, enabling the definition of the so called (co)homology group~\footnote{Actually, it is a quotient space.
-But it can be considered an abelian group under addition.}
-\begin{equation}
-H:=
-\krn(a)/\rng(b)
-\cong
-\krn(a)\cap \krn(b^*)
-\cong
-\krn\,L
-\nonumber
-\end{equation}
-whose elements are called (co)homology classes.
-Here, $L:=a^*a+bb^*:V\to V$ is called the Hodge Laplacian associated to $a$ and $b$.
-It can be shown that
-\begin{equation}
-\label{eqA2}
-V=\rng\,a^*\oplus\krn\,L\oplus\rng\,b
-\end{equation}
-meaning that, for any vector $v\in V$, there exist the unique decomposition
-\begin{equation}
-\label{eqA3}
-v = a^*(w)+h+b(u)
-\end{equation}
-for some $w\in W$ and $u\in U$, and where 
-$h\in \krn\,L\subseteq V$ is the unique vector of the (co)homology class $[v]\in \krn\,a/\rng\,b$ satisfying $h\in(\rng\,b)^{\perp}=\krn\,b^*$.
-For this reason, $h$ is called the harmonic representative of $[v]$.
-Equations~\ref{eqA2}~and~\ref{eqA3} represent the so called Hodge decomposition of $V$ and $v$, respectively.
-Importantly, $\krn\,b^*=\rng\,a^*\oplus \krn\,L$ and $\krn\,a=\krn\,L\oplus\rng\,b$.
-Moreover, $L$ is a positive semidefinite operator since $L=L^*$, and $(a^*a)^*=a^*a$, $(bb^*)^*=bb^*$.
-Furthermore, if $[X,Y]:=XY-YX$, then $[L,a^*a]=[L,bb^*]=[a^*a,bb^*]=0$.
-
-\subsection{Homology}
-
-In the context of simplicial complexes, homology emerges from the application of Hodge's decomposition using $\partial_{k}$ and $\partial_{k+1}$ in place of $a$ and $b$ respectively, since $ab=\partial_{k}\partial_{k+1}=0$.
-As a result, the Hodge Laplacian takes the form
-$$
-\mathcal{L}_k
-=
-\mathcal{L}_k^{\downarrow}
-\oplus
-\mathcal{L}_k^{\uparrow}
-:C_k\to C_k
-$$
-where
-$$
-\mathcal{L}_k^{\downarrow}=\partial_{k}^*\partial_{k} : C_k\to C_{k-1}\to C_k
-\hspace{1cm}
-\mbox{and}
-\hspace{1cm}
-\mathcal{L}_k^{\uparrow}=\partial_{k+1}\partial_{k+1}^*:C_k\to C_{k+1} \to C_k
-$$
-so
-$$
-C_k 
-= 
-\rng\,\partial_k^*
-\oplus
-\krn\,\mathcal{L}_k
-\oplus
-\rng\,\partial_{k+1}
-$$
-and, therefore, for any $c\in C_k$, there exists unique 
-$p\in \rng\,\partial_k^*\subseteq C_k$, $h\in \krn\,\mathcal{L}_k\subseteq C_k$ and $q\in \rng\,\partial_{k+1}\subseteq C_k$
-such that $c = p+h+q$.
-Importantly, $\krn\,\partial_{k+1}^*=\rng\,\partial_{k}^*\oplus \krn\,\mathcal{L}_k$ and $\krn\,\partial_{k}=\krn\,\mathcal{L}_k\oplus\rng\,\partial_{k+1}$.
-Often, $\mathcal{L}_k$, $\mathcal{L}_k^{\downarrow}$ and $\mathcal{L}_k^{\uparrow}$ are called the $k$-th, $k$-th lower and $k$-th upward Laplacians, respectively.
-These can be aggregated to obtain a Laplacian over $C$
-$$
-\mathcal{L} = \partial^*\partial \oplus \partial\partial^* = \mathcal{L}_0\oplus...\oplus \mathcal{L}_n : C\to C
-$$
-Note, $\mathcal{L}_k^{\downarrow}$ and $\mathcal{L}_k^{\uparrow}$ combine the information of $k$-chains with the information of $(k-1)$-chains and $(k+1)$-chains respectively.
-
-The $k$-th Dirac operator over $C$
-$$
-\mathcal{D}_k=\partial_k^*+\partial_k
-$$
-can be also introduced, and it is related to the concept of spinor~\cite{bianconi2021topological}.
-It satisfies $\mathcal{D}_k^2=\mathcal{L}_k^{\downarrow}\oplus\mathcal{L}_{k-1}^{\uparrow}$, combining the information between $(k-1)$-chains and $k$-chains.
-Similarly, $\mathcal{D}=\partial^*+\partial=\mathcal{D}_0+...+ \mathcal{D}_n$ is such that $\mathcal{D}^2=\mathcal{L}$.
-
-The $k$-th Betti numbers $\beta_k=\rnk\, H_k=\rnk\,\krn \,\partial_k - \rnk\,\rng\,\partial_{k+1}$, which are the ranks of the homology groups, are topological invariants  characterizing the properties of topological spaces.
-In the present context where $C$ is a vector space, these homology groups are subspaces, so their ranks are actually their dimensions.
-
-Note, when $m_{rs}$ is the trivial metric $\delta_{rs}$, then $(\mathcal{L}_k^{\downarrow})^s{}_s=(\alpha_k)^s{}_r(\alpha_k^T)^r{}_s=k+1$ is the number of facets of $s$.
-Similarly, $(\mathcal{L}_k^{\uparrow})^s{}_s=(\alpha_{k+1}^T)^s{}_r(\alpha_{k+1})^r{}_s=k_s=\zeta_{s,k+1}$ is the number of cofacets or degree of $s$.
-For a general metric, $(\mathcal{L}_k^{\downarrow})^s{}_s$ and $(\mathcal{L}_k^{\uparrow})^s{}_s$ represent weighted versions of these quantities.
-
-\subsection{Cohomology\label{appA:cohomology}}
-
-Cohomology emerges from the application of  Hodge decomposition over cochains, using $d_{k}$ and $d_{k-1}$ in place of $a$ and $b$ respectively, so $ab = d_kd_{k-1}=0$.
-As a result, the Hodge Laplacian takes the form
-$$
-L_k
-=
-L_k^{\uparrow}
-\oplus
-L_k^{\downarrow}
-:C_k^*\to C_k^*
-$$
-where
-$$
-L_k^{\uparrow}=d_{k}^*d_{k}:C_k^*\to C_{k+1}^* \to C_k^*
-\hspace{1cm}
-\mbox{and}
-\hspace{1cm}
-L_k^{\downarrow}=d_{k-1}d_{k-1}^* : C_k^*\to C_{k-1}^*\to C_k^*
-$$
-so
-$$
-C_k^*
-=
-\rng\,d_k^*
-\oplus
-\krn\,L_k
-\oplus
-\rng\,d_{k-1}
-$$
-and, therefore, for any $f\in C_k^*$, there exists unique 
-$s\in \rng\, d_k^*\subseteq C_k^*$,
-$h\in \krn\, L_k\subseteq C_k^*$ and
-$g\in \rng\, d_{k-1}\subseteq C_k^*$
-such that $f = s+h+g$.
-Usually, $s$, $h$ and $g$ are called the coexact, harmonic and exact components of $f$ respectively.
-Importantly, $\krn\,d_{k-1}^*=\rng\,d_k^*\oplus \krn\,L_k$ and $\krn\,d_k=\krn\,L_k\oplus\rng\,d_{k-1}$.
-The Laplacians $L_k^{\uparrow}$ and $L_k^{\downarrow}$ combine the information in $k$-cochains with the information in $(k+1)$-cochains and $(k-1)$-cochains respectively.
-Moreover, they can be combined to obtain the so called Hodge Laplacian over $C^*$
-$$
-L = dd^* \oplus d^*d= L_0\oplus...\oplus L_n : C^*\to C^*
-$$
-Furthermore, in analogy to its topological counterpart $\mathcal{D}_k$, the $k$-th Dirac operator over cochains
-$$
-D_k = d_k^*+d_k
-$$
-can be introduced.
-It satisfies $D_k^2=L_k^{\uparrow}\oplus L_{k+1}^{\downarrow}$, combining the information between $k$-cochains and $(k+1)$-cochains.
-Correspondingly, $D=d^*+d=D_0+...+ D_n$ is such that $D^2=L$.
-
-The Hodge decomposition of chochains is a generalization of the Helmholtz decomposition of vector calculus in 3D, to arbitrary combinatorial topologies and dimensions.
-In fact, the coexact, harmonic and exact componets of the Hodge decomposition are recognized as the solenoid, harmonic and gradient components of the Helmholtz decomposition.
-Moreover, $L_0=d_0^*d_0+d_{-1}d_{-1}^*=d_0^*d_0+0=-\dive\circ \grad$ can be recognized as the {\em graph Laplacian}, a combinatorial analogous of the Laplace operator over scalar fields, and $L_1=d_0d_0^*+d_1^*d_1 = \grad\circ \grad^*+\curl^*\circ \curl=-\grad\circ \dive+\curl^*\circ \curl$ can be recognized as the {\em graph Helmholtzian} which is the combinatorial analogous of the vector Laplacian~\cite{lim2020hodge}.
-When a simplicial complex is obtained from a discretization of an oriented region in euclidean space, then a Hodge star operator can be defined, relating chains and cochains with volume forms.
-In particular, when the triangulation is a 3D-space, then $\curl^*=\curl$ by Hodge duality.
-This is the reason for which there is no special name for $\curl^*$ in Physics~\cite{grady2010discrete}.
+%\section{
+%Combinatorial differential topology, geometry and calculus
+%\label{appA}
+%}
 

### 2025-07-11 draft/draft.tex
progress

@@
 \section{
+Data
+\label{sec:data}
+}
+
+There are 5 leagues: England, France, Germany, Italy and Spain.
+Each has 20 teams, with the exception of Germany.
+In each league, each pair of teams plays two games, one as the local team and the other as the visitor.
+In each game between the $i$-th and the $j$-th team, different metrics are measured.
+With each metric, a weighted graphs of entries $w_{ij}\in \RR$ is obtained.
+
+\section{
 Results
@@
 \item 
-Compute ratings and rankings from the full weighted networks.
+Compute ratings and rankings from the full weighted networks $w$ defining $f_{ij}=w_j-w_i$ for each pair $ij$ of a given league.
 
@@
 Compute ratings and rankings from sparsified versiond of the weighted networks, by removing a varying fraction of links sorted in decreasing order of absolute weight.
+
+\item 
+Compute ratings defining the $f_{ij}=\phi_j-\phi_i$ where the strength $\phi_i = \sum_j w_{ij}$ defines a local potential for the $i$-th team.
 \end{enumerate}

### 2025-07-12 draft/draft.tex
progress

--- 2025-07-11 draft/draft.tex
@@
 \section{
+Data
+\label{sec:data}
+}
+
+The data comprehends 5 soccer leagues from England, France, Germany, Italy and Spain.
+Each has 20 teams, with the exception of Germany which has 18.
+Each pair of teams of each league plays two games, one as local and another the visitor.
+For each game between two teams $i$ and $j$ of a given league, a score $\phi_{ij}\in \RR$ is provided.
+
+Each matrix $\phi$ of scores $\phi_{ij}$ defines a fully connected undirected network for the teams in a league.
+In some experiments, some fraction of links is removed from the network according to some criterium, defining a graph $\mathcal{G}$ that is not fully connected.
+The graph is promoted to a simplicial complex of 0-simplices, 1-simplices and 2-simplices composed by the nodes, the links and the triplets of connected nodes in $\mathcal{G}$, respectively.
+A triplet $ijk$ of the nodes $i$, $j$ and $k$ of $\mathcal{G}$ is connected if the links $ij$, $ik$ and $jk$ belong to $\mathcal{G}$.
+
+From a score $\phi$, a corresponding cochain $f\in C_1^*$ of coordinates $f_{ij}=\phi_{ji}-\phi_{ij}$ is defined.
+Also, the cochain $f'\in C_1^*$ of coordinates $f'_{ij}=\phi_j-\phi_i$ with team strengths $\phi_i=\sum_j \phi_{ij}$ is considered.
+
+\section{
 Results
@@
 \item 
-Compute ratings and rankings from the full weighted networks.
+Compute ratings and rankings from the full weighted network of scores $\phi_{ij}$.
 
 \item 
-Compute ratings and rankings from sparsified versiond of the weighted networks, by randomly removing a varying fraction of links.
+Compute ratings and rankings from sparsified versiond of the weighted networks of schores $\phi_{ij}$, by randomly removing a varying fraction of links.
 
 \item 
-Compute ratings and rankings from sparsified versiond of the weighted networks, by removing a varying fraction of links sorted in decreasing order of absolute weight.
+Compute ratings and rankings from sparsified versiond of the weighted networks, by removing a varying fraction of the weighted links of the weighted matrix of scores $\phi_{ij}$ sorted in decreasing order of absolute weight.
+
+\item 
+Compute ratings defining the $f_{ij}=\phi_j-\phi_i$ where the strength $\phi_i = \sum_j w_{ij}$ defines a local potential for the $i$-th team.
 \end{enumerate}

### 2025-07-13 draft/draft.tex
progress

@@
 Each has 20 teams, with the exception of Germany which has 18.
-Each pair of teams of each league plays two games, one as local and another the visitor.
-For each game between two teams $i$ and $j$ of a given league, a score $\phi_{ij}\in \RR$ is provided.
+For each pair of teams $i$ and $j$ of a given league, a score $\phi_{ij}\in \RR$ is provided, which is the aggregate of the two games between them, one as a local and another as a visitor.
+There are ten different types of scores:
+\begin{enumerate}
+\item {\bf crossings:}
+\item {\bf counterattacks:}
+\item {\bf pressure loss:}
+\item {\bf build-up time:}
+\item {\bf direct play:}
+\item {\bf pressure points:}
+\item {\bf shots:}
+\item {\bf flow rate:}
+\item {\bf maintenance time:}
+\item {\bf middle-zone time:}
+\end{enumerate}
 

### 2025-07-15 draft/draft.tex
progress

@@
 \item 
-Compute ratings defining the $f_{ij}=\phi_j-\phi_i$ where the strength $\phi_i = \sum_j w_{ij}$ defines a local potential for the $i$-th team.
+Compute ratings defining the 1-cochain coordiantes $f'_{ij}=\phi_j-\phi_i$, where the strength $\phi_i = \sum_j w_{ij}$ defines a local potential for the $i$-th team.
 \end{enumerate}

### 2025-09-02 draft/draft.tex
progress

@@
 
-Bla bla...
+See~\cite{perotti2025analysis,chacoma2025data}.
 
@@
 \section{
+Methods
+\label{sec:methods}
+}
+
+\subsection{
 Data
@@
 
-The data comprehends 5 soccer leagues from England, France, Germany, Italy and Spain.
-Each has 20 teams, with the exception of Germany which has 18.
-For each pair of teams $i$ and $j$ of a given league, a score $\phi_{ij}\in \RR$ is provided, which is the aggregate of the two games between them, one as a local and another as a visitor.
-There are ten different types of scores:
+The dataset covers five soccer leagues: England, France, Germany, Italy, and Spain.
+Each league consists of 20 teams, except for Germany, which has 18.
+Every pair of teams $i,j$ plays two matches: one with $i$ as the home team and $j$ as the away team, and the other with the reverse condition.
+The teams are indexed according to their position in that year’s final standings.
+For each pair of teams $i$ and $j$ in a given league, a metric $m_{ij}\in \mathbb{R}$ is provided, which takes a positive value if it presumably favors team $i$ over team $j$, and a negative value otherwise.
+Different types of empirically determined metrics are considered:
 \begin{enumerate}
@@
 \item {\bf maintenance time:}
-\item {\bf middle-zone time:}
+\item {\bf middle-zone time:} bla bla,
 \end{enumerate}
+and metric-styles are also considered~\cite{chacoma2025data}.
+Additionally, in order to test the inference of ratings, two types of ground truth metrics are also defined $m_{ij}=\frac{1}{2}(i-j)$ and $m_{ij}=\frac{1}{2}(\sigma_j-\sigma_i)$ using the rankings $i$ and scores $\sigma_i$ at the end of the season.
+Note, $i<j$ if $\sigma_i>\sigma_j$ by definition.
+
+{\color{red}[HERE...]}
 

### 2025-09-03 draft/draft.tex
progress

--- 2025-07-15 draft/draft.tex
@@
 
-The data comprehends 5 soccer leagues from England, France, Germany, Italy and Spain.
-Each has 20 teams, with the exception of Germany which has 18.
-For each pair of teams $i$ and $j$ of a given league, a score $\phi_{ij}\in \RR$ is provided, which is the aggregate of the two games between them, one as a local and another as a visitor.
-There are ten different types of scores:
+The dataset covers five soccer leagues: England, France, Germany, Italy, and Spain.
+Each league consists of 20 teams, except for Germany, which has 18.
+Every pair of teams $i,j$ plays two matches: one with $i$ as the home team and $j$ as the away team, and in the other match it is the other way around.
+The teams are indexed or ranked according to their position in that year’s final standings.
+The standings is characterized by some tournament points $\pi_i$ for each team $i$ in each league.
+Therefore, the condition $i<j$ holds if and only if $\pi_i>\pi_j$ for any two teams $i$ and $j$ in a given league.
+
+For each pair of teams $i$ and $j$ in a given league, several pairwise metrics $m_{ij}\in \mathbb{R}$ are provided.
+Each metric takes a positive value if it presumably favors team $i$ over team $j$, and a negative value otherwise.
+The different types of empirically determined metrics being considered are:
 \begin{enumerate}
@@
 \item {\bf maintenance time:}
-\item {\bf middle-zone time:}
+\item {\bf middle-zone time:} bla bla.
 \end{enumerate}
+Namely, $m_{ij}$ corresponds to $i$ being the home team and $j$ being the away team.
+Additionally, metric-styles derived from the Singular Value Decomposition of previous metrics are also considered~\cite{chacoma2025data}.
+Moreover, in order to test the inference of ratings, two types of ground-truth metrics are also defined.
+Namely, the rank-based metric $m_{ij}=\frac{1}{2}(i-j)$ and the tournament point metric $m_{ij}=\frac{1}{2}(\sigma_j-\sigma_i)$.
+Notice, $m_{ij}$ is generally independent of the metric $m_{ji}$ since they belong to different matches between $i$ and $j$.
+In the case of the ground-truth metrics, it holds that $m_{ij}=-m_{ji}$.
+
+{\color{red}[HERE...]}
 

### 2025-09-03 draft/draft.tex
progress

@@
 The dataset covers five soccer leagues: England, France, Germany, Italy, and Spain.
-Each league consists of 20 teams, except for Germany, which has 18.
+Each league consists of $N=20$ teams, except for Germany, which has $N=18$.
 Every pair of teams $i,j$ plays two matches: one with $i$ as the home team and $j$ as the away team, and in the other match it is the other way around.
@@
 Additionally, metric-styles derived from the Singular Value Decomposition of previous metrics are also considered~\cite{chacoma2025data}.
-Moreover, in order to test the inference of ratings, two types of ground-truth metrics are also defined.
-Namely, the rank-based metric $m_{ij}=\frac{1}{2}(i-j)$ and the tournament point metric $m_{ij}=\frac{1}{2}(\sigma_j-\sigma_i)$.
 Notice, $m_{ij}$ is generally independent of the metric $m_{ji}$ since they belong to different matches between $i$ and $j$.
-In the case of the ground-truth metrics, it holds that $m_{ij}=-m_{ji}$.
+As a consequence, $m_{ij}+m_{ji}$ is generally different from zero and is the net contribution between teams $i$ and $j$.
 
-{\color{red}[HERE...]}
-
-Each matrix $\phi$ of scores $\phi_{ij}$ defines a fully connected undirected network for the teams in a league.
-In some experiments, some fraction of links is removed from the network according to some criterium, defining a graph $\mathcal{G}$ that is not fully connected.
-The graph is promoted to a simplicial complex of 0-simplices, 1-simplices and 2-simplices composed by the nodes, the links and the triplets of connected nodes in $\mathcal{G}$, respectively.
-A triplet $ijk$ of the nodes $i$, $j$ and $k$ of $\mathcal{G}$ is connected if the links $ij$, $ik$ and $jk$ belong to $\mathcal{G}$.
+With each metric, a 1-cochain is defined as
+$$
+f_{ij}
+=
+\frac{1}{2}(m_{ij}+m_{ji})
+$$
+Moreover, in order to test the inference method, two types of ground-truth 1-cochains are also defined.
+Namely, the rank-based 1-cochain $f_{ij}=\frac{1}{2}(i-j)$ and the tournament point 1-cochain $f_{ij}=\frac{1}{2}(\pi_j-\pi_i)$.
+Note, $f_{ij}=-f_{ji}$ holds by definition.
+These 1-cochains $f$ define a fully connected network or graph, since all pairs of teams play two matches against each other in each league.
+Each of these fully connected graphs of $N=n+1$ nodes is promoted to a simplicial complex compose of 0-simplices labeled by $i$, 1-simplices labeled by $ij$ and 2-simplices labeled by $ijk$ where $i,j,k$ are different elements in $\{0,1,2,...,n\}$.
 

### 2025-09-03 draft/draft.tex
progress

@@
 
+\subsection{Definition of 1-cochains}
+
 With each metric, a 1-cochain is defined as
@@
 Moreover, in order to test the inference method, two types of ground-truth 1-cochains are also defined.
-Namely, the rank-based 1-cochain $f_{ij}=\frac{1}{2}(i-j)$ and the tournament point 1-cochain $f_{ij}=\frac{1}{2}(\pi_j-\pi_i)$.
+Namely, the rank-based 1-cochain $f_{ij}=i-j$ and the tournament points based 1-cochain $f_{ij}=\pi_j-\pi_i$.
 Note, $f_{ij}=-f_{ji}$ holds by definition.
-These 1-cochains $f$ define a fully connected network or graph, since all pairs of teams play two matches against each other in each league.
-Each of these fully connected graphs of $N=n+1$ nodes is promoted to a simplicial complex compose of 0-simplices labeled by $i$, 1-simplices labeled by $ij$ and 2-simplices labeled by $ijk$ where $i,j,k$ are different elements in $\{0,1,2,...,n\}$.
+These 1-cochains $f$ define a fully connected network or graph since, as already mentioned, all pairs of teams play two matches against each other in each league.
+Each of these fully connected graphs of $N=n+1$ nodes is promoted to a simplicial complex composed of 0-simplices labeled by $i$, 1-simplices labeled by $ij$ and 2-simplices labeled by $ijk$ where $i,j$ and $k$ run over all choices in $\{0,1,2,...,n\}$ such that $i<j$ or $i<j<k$, depending on the case.
 

### 2025-09-03 draft/draft.tex
progress

--- 2025-09-03 draft/draft.tex
@@
 
-With each metric, a 1-cochain is defined as
+\subsection{Inference of ratings and rankings}
+
+With each metric, the coefficients
 $$
@@
 $$
-Moreover, in order to test the inference method, two types of ground-truth 1-cochains are also defined.
-Namely, the rank-based 1-cochain $f_{ij}=\frac{1}{2}(i-j)$ and the tournament point 1-cochain $f_{ij}=\frac{1}{2}(\pi_j-\pi_i)$.
-Note, $f_{ij}=-f_{ji}$ holds by definition.
-These 1-cochains $f$ define a fully connected network or graph, since all pairs of teams play two matches against each other in each league.
-Each of these fully connected graphs of $N=n+1$ nodes is promoted to a simplicial complex compose of 0-simplices labeled by $i$, 1-simplices labeled by $ij$ and 2-simplices labeled by $ijk$ where $i,j,k$ are different elements in $\{0,1,2,...,n\}$.
+are defined for each $i$ and $j$ in $\{0,1,...,n\}$ such that $i<j$ and where $N=n+1$ is the number of teams in the corresponding league.
+Moreover, in order to test the inference method, two types of ground-truth coefficients are also defined.
+Namely, the rank-based coefficients $f_{ij}=i-j$ and the tournament points based coefficients $f_{ij}=\pi_j-\pi_i$.
+These coefficients correspond to a fully connected network or graph with $N$ nodes since, as already mentioned, all pairs of teams play two matches against each other in each league.
+Next, each fully connected graphs is promoted to a simplicial complex composed of 0-simplices labeled by $i$, 1-simplices labeled by $ij$ with $i<j$ and 2-simplices labeled by $ijk$ with $i<j<k$.
+Finally, the coefficients $f_{ij}$ are promoted to be the components of a 1-cochain  $f$, where the anti-symmetric relation $f_{ji}=-f_{ij}$ must hold by definition.
 
-From a score $\phi$, a corresponding cochain $f\in C_1^*$ of coordinates $f_{ij}=\phi_{ji}-\phi_{ij}$ is defined.
-Also, the cochain $f'\in C_1^*$ of coordinates $f'_{ij}=\phi_j-\phi_i$ with team strengths $\phi_i=\sum_j \phi_{ij}$ is considered.
+%From a score $\phi$, a corresponding cochain $f\in C_1^*$ of coordinates $f_{ij}=\phi_{ji}-\phi_{ij}$ is defined.
+%Also, the cochain $f'\in C_1^*$ of coordinates $f'_{ij}=\phi_j-\phi_i$ with team strengths $\phi_i=\sum_j \phi_{ij}$ is considered.
+
+{\color{red}[JUAN: Write here the inference problem.]}
 

### 2025-09-03 draft/draft.tex
progress

@@
 
-Bla bla...
+\subsection{Inference of ratings and rankings}
+
+Consider a set of $N=n+1$ teams labeled by $\{0,1,...,n\}$ to be rated and ranked.
+For this aim, assume that some set of pairwise comparison scores $\psi_{ij}=-\psi_{ji}$ is provided for each $ij$ of different labels $i$ and $j$.
+Informally, the score $\psi_{ij}$ should be positive if the pairwise comparison indicates that team $i$ should be rated higher than team $j$ and it should be non-positive otherwise.
+One way to obtain rating scores somehow consistent with the pairwise comparisons scores, consist in finding ratings $w_i$ for $i=0,1,..,n$ such that minimize the sum of squared differences
+$$
+\sum_{ij:i<j} |\psi_{ij}-(w_i-w_j)|^2.
+$$
 

### 2025-09-03 draft/draft.tex
progress

@@
 Informally, the score $\psi_{ij}$ should be positive if the pairwise comparison indicates that team $i$ should be rated higher than team $j$ and it should be non-positive otherwise.
-One way to obtain rating scores somehow consistent with the pairwise comparisons scores, consist in finding ratings $w_i$ for $i=0,1,..,n$ such that minimize the sum of squared differences
+One way to infer rating scores somehow consistent with the pairwise comparisons scores, consist in finding ratings $w_i$ for $i=0,1,..,n$ that minimize the sum of squared differences
 $$
@@
 $$
+This is a well defined mathematical problem that can be conveniently reformulated in the language of cochains over simplicial complexes.
+In fact, this reformulation is known as the HodgeRank method for the inference of ratings and rankings starting from pairwise comparison scores~\cite{jiang2011statistical,
+perotti2025analysis}.
 

### 2025-09-03 draft/draft.tex
progress

@@
 
+To define HodgeRank, it is convenient to introduce a simplicial complex composed of a 0-simplex $i$ for each team, a 1-simplex $ij$ for each pairwise comparison $\psi_{ij}$ between teams and a 2-simplex $ijk$ for each existing triple $(ij,jk,ki)$ in the set of 1-simplices.
+Recall here that $ij$ and $ji$ are both labels of the same simplex, as well as any permutation of $ijk$ is also a label of the same 2-simplex.
+
 \section{

### 2025-09-03 draft/draft.tex
progress

--- 2025-09-03 draft/draft.tex
@@
 
-\subsection{Inference of ratings and rankings}
+\subsection{HodgeRank for inference of team's ratings and rankings}
 
-Consider a set of $N=n+1$ teams labeled by $\{0,1,...,n\}$ to be rated and ranked.
-For this aim, assume that some set of pairwise comparison scores $\psi_{ij}=-\psi_{ji}$ is provided for each $ij$ of different labels $i$ and $j$.
-Informally, the score $\psi_{ij}$ should be positive if the pairwise comparison indicates that team $i$ should be rated higher than team $j$ and it should be non-positive otherwise.
-One way to obtain rating scores somehow consistent with the pairwise comparisons scores, consist in finding ratings $w_i$ for $i=0,1,..,n$ such that minimize the sum of squared differences
+%Consider a set of $N=n+1$ teams labeled by $\{0,1,...,n\}$ to be rated and ranked.
+%For this aim, assume that some set of pairwise comparison scores $\psi_{ij}=-\psi_{ji}$ is provided for each pair $ij$ of different teams $i$ and $j$.
+%Informally, the score $\psi_{ij}$ should be positive if the comparison indicates that team $i$ should preferably be rated higher than team $j$ and the opposite otherwise.
+%
+%One way to infer rating scores somehow consistent with the pairwise comparisons scores $\phi_{ij}$, consist in finding ratings $w_i$ for $i=0,1,..,n$ that minimize the sum of squared differences
+%$$
+%\sum_{ij:i<j} |\psi_{ij}-(w_i-w_j)|^2.
+%$$
+%This is a well defined mathematical problem that can be conveniently reformulated in the language of cochains over simplicial complexes, conceiving a method for the inference of ratings and rankings that is named  HodgeRank~\cite{jiang2011statistical,
+%perotti2025analysis}.
+
+Consider a set of $N=n+1$ teams labeled $\{0,1,\dots,n\}$ to be rated and ranked.
+Suppose that for each distinct pair of teams $i$ and $j$, a pairwise comparison score $\psi_{ij}=-\psi_{ji}$ is given.
+Informally, $\psi_{ij}>0$ suggests that team $i$ should be rated higher than team $j$, while $\psi_{ij}<0$ suggests the opposite.
+
+A natural approach to inferring rating scores consistent with the pairwise comparisons is to assign ratings $w_i$ (for $i=0,1,\dots,n$) that minimize the sum of squared discrepancies
 $$
-\sum_{ij:i<j} |\psi_{ij}-(w_i-w_j)|^2.
+\sum_{i<j} \bigl(\psi_{ij}-(w_i-w_j)\bigr)^2.
 $$
+This yields a well-posed optimization problem that admits an elegant reformulation in the language of cochains over simplicial complexes.
+The resulting framework for the inference of ratings and rankings is known as {\em HodgeRank}~\cite{jiang2011statistical, perotti2025analysis}.
+
+%To define HodgeRank, it is convenient to introduce a simplicial complex composed of a 0-simplex $i$ for each team, a 1-simplex $ij$ for each pairwise comparison $\psi_{ij}$ between teams and a 2-simplex $ijk$ for each existing triple $(ij,jk,ki)$ in the set of 1-simplices.
+%Recall here that $ij$ and $ji$ are both labels of the same simplex, as well as any permutation of $ijk$ is also a label of the same 2-simplex.
+
+To define HodgeRank, an abstract simplicial complex has to be defined.
+For this, let the vertex set be $V=\{0,1,\dots,n\}$, representing the teams, and let the edge set $E$ consist of the unordered pairs $\{i,j\}$ for which a pairwise comparison score $\psi_{ij}$ is available.
+Then, the associated abstract simplicial complex is
+$$
+K = K_0 \cup K_1 \cup K_2,
+$$
+where
+\begin{eqnarray}
+K_0 &=& \{\{i\} : i \in V\}, \notag \\  
+K_1 &=& \{\{i,j\} : \{i,j\} \in E\}, \notag \\
+K_2 &=& \{\{i,j,k\} : \{i,j\},\{j,k\},\{k,i\} \in E\} \notag
+\end{eqnarray}
+are the sets of 0, 1 and 2 simplices, respectively.
+Thus, each team corresponds to a 0-simplex, each comparison to a 1-simplex, and each closed triple of comparisons to a 2-simplex.
+By convention, $\{i,j\}=\{j,i\}$ denotes the same 1-simplex, and any permutation of $\{i,j,k\}$ denotes the same 2-simplex.
 

### 2025-09-05 draft/draft.tex
progress

@@
 Consider a set of $N=n+1$ teams labeled $\{0,1,\dots,n\}$ to be rated and ranked.
-Suppose that for each distinct pair of teams $i$ and $j$, a pairwise comparison score $\psi_{ij}=-\psi_{ji}$ is given.
-Informally, $\psi_{ij}>0$ suggests that team $i$ should be rated higher than team $j$, while $\psi_{ij}<0$ suggests the opposite.
+Suppose that for each distinct pair of teams $i$ and $j$, a pairwise comparison score $f_{ij}=-f_{ji}$ is given.
+Informally, $f_{ij}>0$ suggests that team $i$ should be rated higher than team $j$, while $f_{ij}<0$ suggests the opposite.
 
@@
 $$
-\sum_{i<j} \bigl(\psi_{ij}-(w_i-w_j)\bigr)^2.
+\sum_{i<j} \bigl(f_{ij}-(w_i-w_j)\bigr)^2.
 $$
@@
 %Recall here that $ij$ and $ji$ are both labels of the same simplex, as well as any permutation of $ijk$ is also a label of the same 2-simplex.
+
+\subsection{Discrete calculus on simplicial complexes}
 
@@
 By convention, $\{i,j\}=\{j,i\}$ denotes the same 1-simplex, and any permutation of $\{i,j,k\}$ denotes the same 2-simplex.
+
+Next, the simplicial complex is equipped with a vector space $C$ of vectors called chains, whose canonical basis is denoted by $\{e_s:s\in K\}$.
+
+The vector space of chains $C$ is equipped with the trivial inner product
+$$
+e_r.e_s = \delta_{rs}
+$$
+for each pair $r,s \in K$.
+Using Einstein's implicit summation convention, an arbitrary chain $c$ in $C$ can be linearly decomposed in the canonical basis as $c=c^se_s$ where the $c^s$ are coefficients in $\RR$.
+The vector space of chains admits the following decomposition in direct sums
+$$
+C = C_0\oplus C_1\oplus C_2
+$$
+where $C_k$ is the subspace of $C$ spanned by the basis of $k$-chains, i.e. by $\{e_s:s\in K_k\}$.
+
+The covectors $f$ in the dual space $C^*$ of $C$ are called cochains.
+The dual basis is denoted by $\{e^s:s\in K\}$ and satisfies
+$$
+e^r(e_s) = \delta^r{}_s
+$$
+for each $s,r\in K$.
+An arbitrary cochain $f$ in $C^*$ can be linearly decomposed in the dual basis as $f=f_se^s$ where the $f_s$ are coefficients in $\RR$.
+The vector space of cochains admits the following decomposition in direct sums
+$$
+C^* = C_0^*\oplus C_1^*\oplus C_2^*
+$$
+where $C_k^*$ is the subspace of $C^*$ spanned by the dual basis of $k$-cochains, i.e. by $\{e^s:s\in K_k\}$.
+
+Riesz's theorem establishes that for any cochain $f\in C^*$, there is an unique chain in $C$ denoted by $f^{\flat}$ such that $f(q) = f^{\flat}.c$ for any chain $c\in C$.
+This results, induces an inner product on on $C^*$ given by
+$$
+f.g = f^{\flat}.g^{\flat}
+$$
+for any pair of cochains $f,g\in C^*$.
+The map $C^*\ni f\to f^{\flat}\in C$ is an isomorphism, whose inverse is denoted by $C\ni c\to c^{\sharp}\in C^*$.
+
+A notion of orientation is imposed to the vector space of chains, by requering that
+$$
+e_s
+=
+e_{i_0...i_k}
+=
+(-1)^p
+e_{j_0...j_k}
+$$
+for each simplex $s=\{i_0,...,i_k\}$ of $K$
+when $0\leq i_0<i_1<...<i_k\leq n$ and $p$ is the parity of the permutation taking the sequence $i_0...i_k$ into the sequence $j_0...j_k$.
+
+Thanks to the notion of orientation, a boundary linear operator $\partial:C\to C$ can be defined by its action on the vectors of the canonical basis, i.e. by
+$$
+\partial e_{i_0...i_k}
+=
+\sum_{j=0}^k
+(-1)^j
+e_{i_0...i_{j-1}i_{j+1}...i_k}
+$$
+for any $e_{i_0...i_k}\in C$.
+Note that $\partial(C_k) \subseteq C_{k-1}$, where $C_{-1}=\{0\}$ denotes the trivial subspace of $C$ of dimension 0.
+
+Since the application of any cochain $f$ to any chain $c$ is a bilinear form, the coboundary linear operator $d:C^*\to C^*$ is univocally defined by the relation
+$$
+df(c) = f(\partial c)
+$$
+for any $f\in C^*$ and $c\in C$.
+Note, $d(C_k^*)\subset C_{k+1}^*$ where $C_3^*=\{0\}$ denotes the trivial subspace of $C^*$ of dimension 0.
+
+Since the inner products on $C$ and $C^*$ are also bilinear forms, then the dual boundary $\partial^*:C\to C$ and the dual coboundary $d^*:C^*\to C^*$ linear operators can be univocally defined by
+$$
+\partial^*c.q
+=
+c.\partial q
+$$
+and
+$$
+d^*f.g
+=
+f.d g,
+$$
+for any chains $c,q\in C$ and any cochains $f,g\in C^*$, respectively.
+Note, $\partial^*(C_{k-1})\subseteq C_k$ and $d^*(C_{k+1}^*)\subseteq C_k^*$.
 

### 2025-09-05 draft/draft.tex
progress

@@
 Note that $\partial(C_k) \subseteq C_{k-1}$, where $C_{-1}=\{0\}$ denotes the trivial subspace of $C$ of dimension 0.
+This induces the following direct sum decomposition
+$$
+\partial = \partial_0 \oplus \partial_1 \oplus \partial_2
+$$
+where $\partial_k(C)=\partial(C_k)\subseteq C_{k-1}$ and, in particular, $\partial_0=0$.
 
@@
 Note, $d(C_k^*)\subset C_{k+1}^*$ where $C_3^*=\{0\}$ denotes the trivial subspace of $C^*$ of dimension 0.
+As before, this induces the following direct sum decomposition
+$$
+d = d_0 \oplus d_1 \oplus d_2
+$$
+where $d_k(C^*)=d(C_k^*)\subseteq C_{k+1}^*$ and, in particular, $d_2=0$.
 
@@
 for any chains $c,q\in C$ and any cochains $f,g\in C^*$, respectively.
-Note, $\partial^*(C_{k-1})\subseteq C_k$ and $d^*(C_{k+1}^*)\subseteq C_k^*$.
+Note, that $\partial^*(C_{k-1})\subseteq C_k$ and $d^*(C_{k+1}^*)\subseteq C_k^*$, indicing the follow direct sum decompositions
+$$
+\partial^* = \partial_0^* \oplus \partial_1^* \oplus \partial_2^*
+$$
+and
+$$
+d^* = d_0^* \oplus d_1^* \oplus d_2^*
+$$
+where $\partial_k^*(C)=\partial^*(C_{k-1})\subseteq C_k$ and $d_k^*(C^*)=d(C_{k+1}^*)\subseteq C_k^*$.
 

### 2025-09-05 draft/draft.tex
progress

@@
 
+The boundary and coboundary operators and their duals can be used to define the so called Hodge Laplacians
+$$
+\mathcal{L} = (\partial + \partial^*)^2 = \partial \partial^* + \partial^* \partial
+: C\to C
+$$
+and
+$$
+L = (d+d^*)^2 = dd^* + d^*d : C^*\to C^*
+$$
+where the fact that $\partial^2 = (\partial^*)^2 = 0$ and $d^2=(d^*)^2=0$ is used.
+The first identity $\partial^2=0$ is implied by the alternating properties of the definition of $\partial$, from where the other identities follow.
+These Hodge Laplacians satisfy the following decomposition in direct sums
+$$
+\mathcal{L}
+=
+\mathcal{L}_0
+\oplus
+\mathcal{L}_1
+\oplus
+\mathcal{L}_2
+$$
+and
+$$
+L = L_0 \oplus L_1 \oplus L_2,
+$$
+where in turn
+$$
+\mathcal{L}_k = \mathcal{L}_k^{\downarrow} \oplus \mathcal{L}_k^{\uparrow} : C_k\to C_k,
+$$
+$$
+L_k = L_k^{\downarrow} \oplus L_k^{\uparrow} : C_k^* \to C_k^*,
+$$
+$$
+\mathcal{L}_k^{\downarrow} = \partial_k^*\partial_k : C_k \to C_{k-1} \to C_k,
+$$
+$$
+\mathcal{L}_k^{\uparrow} = \partial_{k+1}\partial_{k+1}^* : C_k \to C_{k+1} \to C_k,
+$$
+$$
+L_k^{\uparrow} = d_k^*d_k : C_k^* \to C_{k+1}^* \to C_k^*,
+$$
+and
+$$
+L_k^{\downarrow} = d_{k-1}d_{k-1}^* : C_k^* \to C_{k-1}^* \to C_k^*.
+$$
+
 \section{

### 2025-09-06 draft/draft.tex
progress

@@
 
-A natural approach to inferring rating scores consistent with the pairwise comparisons is to assign ratings $w_i$ (for $i=0,1,\dots,n$) that minimize the sum of squared discrepancies
+A natural approach to inferring rating scores consistent with the pairwise comparisons is to assign ratings $f_i$ (for $i=0,1,\dots,n$) that minimize the sum of squared discrepancies
 $$
-\sum_{i<j} \bigl(f_{ij}-(w_i-w_j)\bigr)^2.
+\sum_{i<j} \bigl(f_{ij}-(f_i-f_j)\bigr)^2.
 $$
@@
 
+Any $k$-cochain $f$ admits the so called Hodge decomposition
+$$
+f = g\oplus h\oplus s
+$$
+where the direct sum components $g\in \rng d_{k-1}$, $h\in \krn L_k$ and $s\in \rng d_k^*$ are unique and are called the gradient, the harmonic and the solenoid components, respectively.
+As a consequence, there exists non-unique cochains $w\in C_{k-1}^*$ and $u\in C_{k+1}^*$ such that $g=d_{k-1}w$ and $s=d_k^*u$, that can be found solving for
+\begin{eqnarray}
+d_kf 
+&=&
+d_k(g+h+s)
+\notag
+\\
+&=&
+d_k s
+\notag
+\\
+&=&
+d_k d_k^* u
+\notag
+\\
+&=&
+L_{k+1}^{\downarrow} u
+\notag
+\end{eqnarray}
+and
+\begin{eqnarray}
+d_{k-1}^* f 
+&=&
+d_{k-1}^*(g+h+s)
+\notag
+\\
+&=&
+d_{k-1}^* g
+\notag
+\\
+&=&
+d_{k-1}^*d_{k-1} w
+\notag
+\\
+&=&
+L_{k-1}^{\uparrow} w,
+\notag
+\end{eqnarray}
+while the harmonic component can be found from $h=f-g-s$.
+
+\subsection{Discrete calculus reformulation of the infererence of ratings}
+
 \section{

### 2025-09-06 draft/draft.tex
progress

@@
 
-A natural approach to inferring rating scores consistent with the pairwise comparisons is to assign ratings $f_i$ (for $i=0,1,\dots,n$) that minimize the sum of squared discrepancies
-$$
-\sum_{i<j} \bigl(f_{ij}-(f_i-f_j)\bigr)^2.
-$$
+A natural approach to inferring rating scores consistent with the pairwise comparisons is to assign ratings $w_i$ (for $i=0,1,\dots,n$) that minimize the sum of squared discrepancies
+\begin{equation}
+\label{eq:1}
+\sum_{i<j} \bigl(f_{ij}-(w_i-w_j)\bigr)^2.
+\end{equation}
 This yields a well-posed optimization problem that admits an elegant reformulation in the language of cochains over simplicial complexes.
@@
 
-\subsection{Discrete calculus on simplicial complexes}
+\subsection{Discrete exterior calculus on simplicial complexes}
 
-To define HodgeRank, an abstract simplicial complex has to be defined.
-For this, let the vertex set be $V=\{0,1,\dots,n\}$, representing the teams, and let the edge set $E$ consist of the unordered pairs $\{i,j\}$ for which a pairwise comparison score $\psi_{ij}$ is available.
-Then, the associated abstract simplicial complex is
+To review the HodgeRank method for the infrence of ratings and rankings, the theory of discrete exterior calculus on simplicial complexes has to be reviewed first.
+For this, consider a vertex set $V=\{0,1,\dots,n\}$.
+A simplicial complex is a set
 $$
-K = K_0 \cup K_1 \cup K_2,
+K = K_0 \cup K_1 \cup ... \cup K_n
 $$
-where
-\begin{eqnarray}
-K_0 &=& \{\{i\} : i \in V\}, \notag \\  
-K_1 &=& \{\{i,j\} : \{i,j\} \in E\}, \notag \\
-K_2 &=& \{\{i,j,k\} : \{i,j\},\{j,k\},\{k,i\} \in E\} \notag
-\end{eqnarray}
-are the sets of 0, 1 and 2 simplices, respectively.
-Thus, each team corresponds to a 0-simplex, each comparison to a 1-simplex, and each closed triple of comparisons to a 2-simplex.
-By convention, $\{i,j\}=\{j,i\}$ denotes the same 1-simplex, and any permutation of $\{i,j,k\}$ denotes the same 2-simplex.
+of $k$-simplices
+$$
+K_k \subseteq \{\{i_0,...,i_k\} : i_0,...,i_k \in V\}
+$$
+that satisfies the closure condition: if $s\in K$ and $r\subset s$ is non-empty, then $r\in K$.
 
-Next, the simplicial complex is equipped with a vector space $C$ of vectors called chains, whose canonical basis is denoted by $\{e_s:s\in K\}$.
+To define discrete calculus, the simplicial complex is equipped with a vector space $C$ of vectors called chains, whose canonical basis is denoted by $\{e_s:s\in K\}$.
 
@@
 for each pair $r,s \in K$.
-Using Einstein's implicit summation convention, an arbitrary chain $c$ in $C$ can be linearly decomposed in the canonical basis as $c=c^se_s$ where the $c^s$ are coefficients in $\RR$.
+Using Einstein's convention where repeated indices assume an implicit summation, an arbitrary chain $c$ in $C$ can be linearly decomposed in the canonical basis as $c=c^se_s$ where the $c^s$ are coefficients in $\RR$.
 The vector space of chains admits the following decomposition in direct sums
@@
 $$
-for each $s,r\in K$.
+denotes Kronecker's delta for each $s,r\in K$.
 An arbitrary cochain $f$ in $C^*$ can be linearly decomposed in the dual basis as $f=f_se^s$ where the $f_s$ are coefficients in $\RR$.
@@
 \begin{eqnarray}
+\label{eq:2}
 d_kf 
@@
 d_k(g+h+s)
-\notag
+%\notag
 \\
@@
 \begin{eqnarray}
+\label{eq:3}
 d_{k-1}^* f 
@@
 d_{k-1}^*(g+h+s)
-\notag
+%\notag
 \\
@@
 \end{eqnarray}
-while the harmonic component can be found from $h=f-g-s$.
+while
+ the harmonic component can be found from $h=f-g-s$.
 
-\subsection{Discrete calculus reformulation of the infererence of ratings}
+\subsection{HodgeRank}
+
+Expression~\ref{eq:1} can be reformulated in the language of discrete calculuas as
+\begin{equation}
+\label{eq:4}
+|f-d_0(w)|
+\end{equation}
+if the scalar components $f_{ij}$ of the 1-cochain $f=f_{ij}e^{ij}$ are identified with the pairwise comparison socres and the 0-cochain scalar components $w_i$ are identified with the ratings.
+In fact, the vector of ratings minimizing expression~\ref{eq:4} is the solution of Eq.~\ref{eq:3} for the case $k=1$.
+For this, the set of 0-simplices $K_0$ is identified with a set $V$ of teams on a league, the set of 1-simplices $K_1$ is identified with the set $E$ of links or pairwise comparison of teams in the league.
+Additionally, to compute the 2-cochain $u$ in order to obtain the solenoid component of the Hodge decomposition of $f$ from Eq.~\ref{eq:2}, the set of 2-simplices is identified with the set of connected triples in the set of comparisons; i.e.
+$$
+K_2
+=
+\{
+\{i,j,k\}
+:
+\{i,j\}\in E,
+\{j,k\}\in E,
+\{i,k\}\in E
+\}.
+$$
 

### 2025-09-06 draft/draft.tex
progress

@@
 $$
+In this setting, $K_k=\emptyset$ for  $k>2$.
 
@@
 \frac{1}{2}(m_{ij}+m_{ji})
+=
+- f_{ji}
 $$

### 2025-09-08 draft/draft.tex
progress

--- 2025-09-06 draft/draft.tex
@@
 
-\section{
-Theory
+\section{Theory}
 \label{sec:theory}
-}
 
-\subsection{HodgeRank for inference of team's ratings and rankings}
+\subsection{Least-squares inference of team ratings and rankings}
 
-%Consider a set of $N=n+1$ teams labeled by $\{0,1,...,n\}$ to be rated and ranked.
-%For this aim, assume that some set of pairwise comparison scores $\psi_{ij}=-\psi_{ji}$ is provided for each pair $ij$ of different teams $i$ and $j$.
-%Informally, the score $\psi_{ij}$ should be positive if the comparison indicates that team $i$ should preferably be rated higher than team $j$ and the opposite otherwise.
-%
-%One way to infer rating scores somehow consistent with the pairwise comparisons scores $\phi_{ij}$, consist in finding ratings $w_i$ for $i=0,1,..,n$ that minimize the sum of squared differences
-%$$
-%\sum_{ij:i<j} |\psi_{ij}-(w_i-w_j)|^2.
-%$$
-%This is a well defined mathematical problem that can be conveniently reformulated in the language of cochains over simplicial complexes, conceiving a method for the inference of ratings and rankings that is named  HodgeRank~\cite{jiang2011statistical,
-%perotti2025analysis}.
+Consider a set of $N=n+1$ teams labeled $\{0,1,\dots,n\}$.
+For each distinct pair of teams $i$ and $j$, assume that a pairwise comparison score $f_{ij}=-f_{ji}$ is available.
+Intuitively, $f_{ij}>0$ indicates that team $i$ should be rated higher than team $j$, whereas $f_{ij}<0$ suggests the opposite.
 
-Consider a set of $N=n+1$ teams labeled $\{0,1,\dots,n\}$ to be rated and ranked.
-Suppose that for each distinct pair of teams $i$ and $j$, a pairwise comparison score $f_{ij}=-f_{ji}$ is given.
-Informally, $f_{ij}>0$ suggests that team $i$ should be rated higher than team $j$, while $f_{ij}<0$ suggests the opposite.
+A natural approach to inferring ratings consistent with these comparisons is to assign scores $w_i$ (for $i=0,1,\dots,n$) that minimize the sum of squared discrepancies:
+\begin{equation}
+\label{eq:1}
+\sum_{i<j} \bigl(f_{ij}-(w_i-w_j)\bigr)^2.
+\end{equation}
+Once the team ratings $w_0,w_1,\dots,w_n$ have been inferred, the ranking $r_0,r_1,\dots,r_n$ is obtained by requiring that $r_i<r_j$ whenever $w_i>w_j$.
+This setup leads to a well-posed optimization problem that admits an elegant reformulation in terms of cochains over simplicial complexes.
+The resulting framework for rating and ranking inference is known as \emph{HodgeRank}~\cite{jiang2011statistical, perotti2025analysis}.
 
-A natural approach to inferring rating scores consistent with the pairwise comparisons is to assign ratings $f_i$ (for $i=0,1,\dots,n$) that minimize the sum of squared discrepancies
-$$
-\sum_{i<j} \bigl(f_{ij}-(f_i-f_j)\bigr)^2.
-$$
-This yields a well-posed optimization problem that admits an elegant reformulation in the language of cochains over simplicial complexes.
-The resulting framework for the inference of ratings and rankings is known as {\em HodgeRank}~\cite{jiang2011statistical, perotti2025analysis}.
+\subsection{Discrete exterior calculus on simplicial complexes}
 
-%To define HodgeRank, it is convenient to introduce a simplicial complex composed of a 0-simplex $i$ for each team, a 1-simplex $ij$ for each pairwise comparison $\psi_{ij}$ between teams and a 2-simplex $ijk$ for each existing triple $(ij,jk,ki)$ in the set of 1-simplices.
-%Recall here that $ij$ and $ji$ are both labels of the same simplex, as well as any permutation of $ijk$ is also a label of the same 2-simplex.
+To review the HodgeRank method, we summarize the basic elements of discrete exterior calculus on simplicial complexes.  
+Let $V=\{0,1,\dots,n\}$ denote a set of vertices.  
+A simplicial complex is a collection
+\[
+K = K_0 \cup K_1 \cup \dots \cup K_n
+\]
+where
+\[
+K_k \subseteq \bigl\{\{i_0,\dots,i_k\} : 0\leq i_0<\dots<i_k\leq n,\ i_j\in V\bigr\},
+\]
+of $k$-simplices, i.e. unordered subsets $\{i_0,\dots,i_k\}$ of the vertex set satisfying the closure condition: if $s\in K$ and $r\subset s$ with $r\neq \emptyset$, then $r\in K$.
 
-\subsection{Discrete calculus on simplicial complexes}
+The set $K$ of simplices is associated with a vector space $C$ of vectors called chains, whose canonical basis is denoted by $\{e_s:s\in K\}$.
+We equip $C$ with the trivial inner product
+\[
+e_r \cdot e_s = \delta_{rs}, \qquad r,s \in K.
+\]
+Using Einstein summation convention, any chain $c\in C$ can be written as $c=c^se_s$ with coefficients $c^s\in \mathbb{R}$.
+The chain space decomposes into
+\[
+C = C_0 \oplus C_1 \oplus C_2,
+\]
+where $C_k$ is the subspace spanned by $\{e_s:s\in K_k\}$.
 
-To define HodgeRank, an abstract simplicial complex has to be defined.
-For this, let the vertex set be $V=\{0,1,\dots,n\}$, representing the teams, and let the edge set $E$ consist of the unordered pairs $\{i,j\}$ for which a pairwise comparison score $\psi_{ij}$ is available.
-Then, the associated abstract simplicial complex is
-$$
-K = K_0 \cup K_1 \cup K_2,
-$$
-where
-\begin{eqnarray}
-K_0 &=& \{\{i\} : i \in V\}, \notag \\  
-K_1 &=& \{\{i,j\} : \{i,j\} \in E\}, \notag \\
-K_2 &=& \{\{i,j,k\} : \{i,j\},\{j,k\},\{k,i\} \in E\} \notag
-\end{eqnarray}
-are the sets of 0, 1 and 2 simplices, respectively.
-Thus, each team corresponds to a 0-simplex, each comparison to a 1-simplex, and each closed triple of comparisons to a 2-simplex.
-By convention, $\{i,j\}=\{j,i\}$ denotes the same 1-simplex, and any permutation of $\{i,j,k\}$ denotes the same 2-simplex.
+Dually, the space of cochains $C^*$ is spanned by $\{e^s:s\in K\}$, where $e^r(e_s)=\delta^r{}_s$.
+Any $f\in C^*$ decomposes as $f=f_se^s$ with coefficients $f_s\in \mathbb{R}$, and
+\[
+C^* = C_0^* \oplus C_1^* \oplus C_2^*.
+\]
 
-Next, the simplicial complex is equipped with a vector space $C$ of vectors called chains, whose canonical basis is denoted by $\{e_s:s\in K\}$.
+By Riesz’s theorem, every cochain $f\in C^*$ corresponds uniquely to a chain $f^\flat \in C$ such that $f(c) = f^\flat \cdot c$ for all $c\in C$.
+This induces an inner product on $C^*$:
+\[
+f\cdot g = f^\flat \cdot g^\flat.
+\]
+The map $f\mapsto f^\flat$ is an isomorphism, with inverse denoted by $c\mapsto c^\sharp$.
 
-The vector space of chains $C$ is equipped with the trivial inner product
-$$
-e_r.e_s = \delta_{rs}
-$$
-for each pair $r,s \in K$.
-Using Einstein's implicit summation convention, an arbitrary chain $c$ in $C$ can be linearly decomposed in the canonical basis as $c=c^se_s$ where the $c^s$ are coefficients in $\RR$.
-The vector space of chains admits the following decomposition in direct sums
-$$
-C = C_0\oplus C_1\oplus C_2
-$$
-where $C_k$ is the subspace of $C$ spanned by the basis of $k$-chains, i.e. by $\{e_s:s\in K_k\}$.
+\medskip
 
-The covectors $f$ in the dual space $C^*$ of $C$ are called cochains.
-The dual basis is denoted by $\{e^s:s\in K\}$ and satisfies
-$$
-e^r(e_s) = \delta^r{}_s
-$$
-for each $s,r\in K$.
-An arbitrary cochain $f$ in $C^*$ can be linearly decomposed in the dual basis as $f=f_se^s$ where the $f_s$ are coefficients in $\RR$.
-The vector space of cochains admits the following decomposition in direct sums
-$$
-C^* = C_0^*\oplus C_1^*\oplus C_2^*
-$$
-where $C_k^*$ is the subspace of $C^*$ spanned by the dual basis of $k$-cochains, i.e. by $\{e^s:s\in K_k\}$.
+\noindent\textbf{Orientation and boundary.}  
+An orientation is imposed on chains by requiring
+\[
+e_s = e_{i_0\dots i_k} = (-1)^p e_{j_0\dots j_k},
+\]
+for each simplex $s=\{i_0,\dots,i_k\}\in K$, where $0\leq i_0<\dots<i_k\leq n$ and $p$ is the parity of the permutation taking $(i_0,\dots,i_k)$ into $(j_0,\dots,j_k)$.
 
-Riesz's theorem establishes that for any cochain $f\in C^*$, there is an unique chain in $C$ denoted by $f^{\flat}$ such that $f(q) = f^{\flat}.c$ for any chain $c\in C$.
-This results, induces an inner product on on $C^*$ given by
-$$
-f.g = f^{\flat}.g^{\flat}
-$$
-for any pair of cochains $f,g\in C^*$.
-The map $C^*\ni f\to f^{\flat}\in C$ is an isomorphism, whose inverse is denoted by $C\ni c\to c^{\sharp}\in C^*$.
+The boundary operator $\partial:C\to C$ is defined on basis elements as
+\[
+\partial e_{i_0\dots i_k}
+=
+\sum_{j=0}^k (-1)^j e_{i_0\dots i_{j-1}i_{j+1}\dots i_k}.
+\]
+Thus $\partial(C_k)\subseteq C_{k-1}$, with $C_{-1}=\{0\}$.
+This yields the decomposition
+\[
+\partial = \partial_0 \oplus \partial_1 \oplus \partial_2,
+\]
+where $\partial_k(C_k)\subseteq C_{k-1}$ and $\partial_0=0$.
 
-A notion of orientation is imposed to the vector space of chains, by requering that
-$$
-e_s
-=
-e_{i_0...i_k}
-=
-(-1)^p
-e_{j_0...j_k}
-$$
-for each simplex $s=\{i_0,...,i_k\}$ of $K$
-when $0\leq i_0<i_1<...<i_k\leq n$ and $p$ is the parity of the permutation taking the sequence $i_0...i_k$ into the sequence $j_0...j_k$.
+\medskip
 
-Thanks to the notion of orientation, a boundary linear operator $\partial:C\to C$ can be defined by its action on the vectors of the canonical basis, i.e. by
-$$
-\partial e_{i_0...i_k}
-=
-\sum_{j=0}^k
-(-1)^j
-e_{i_0...i_{j-1}i_{j+1}...i_k}
-$$
-for any $e_{i_0...i_k}\in C$.
-Note that $\partial(C_k) \subseteq C_{k-1}$, where $C_{-1}=\{0\}$ denotes the trivial subspace of $C$ of dimension 0.
-This induces the following direct sum decomposition
-$$
-\partial = \partial_0 \oplus \partial_1 \oplus \partial_2
-$$
-where $\partial_k(C)=\partial(C_k)\subseteq C_{k-1}$ and, in particular, $\partial_0=0$.
+\noindent\textbf{Coboundary.}  
+The coboundary operator $d:C^*\to C^*$ is defined by
+\[
+df(c) = f(\partial c),
+\]
+for all $f\in C^*$ and $c\in C$.
+Thus $d(C_k^*) \subseteq C_{k+1}^*$, with $C_3^*=\{0\}$, and
+\[
+d = d_0 \oplus d_1 \oplus d_2, \qquad d_2=0.
+\]
 
-Since the application of any cochain $f$ to any chain $c$ is a bilinear form, the coboundary linear operator $d:C^*\to C^*$ is univocally defined by the relation
-$$
-df(c) = f(\partial c)
-$$
-for any $f\in C^*$ and $c\in C$.
-Note, $d(C_k^*)\subset C_{k+1}^*$ where $C_3^*=\{0\}$ denotes the trivial subspace of $C^*$ of dimension 0.
-As before, this induces the following direct sum decomposition
-$$
-d = d_0 \oplus d_1 \oplus d_2
-$$
-where $d_k(C^*)=d(C_k^*)\subseteq C_{k+1}^*$ and, in particular, $d_2=0$.
+\medskip
 
-Since the inner products on $C$ and $C^*$ are also bilinear forms, then the dual boundary $\partial^*:C\to C$ and the dual coboundary $d^*:C^*\to C^*$ linear operators can be univocally defined by
-$$
-\partial^*c.q
-=
-c.\partial q
-$$
-and
-$$
-d^*f.g
-=
-f.d g,
-$$
-for any chains $c,q\in C$ and any cochains $f,g\in C^*$, respectively.
-Note, that $\partial^*(C_{k-1})\subseteq C_k$ and $d^*(C_{k+1}^*)\subseteq C_k^*$, indicing the follow direct sum decompositions
-$$
-\partial^* = \partial_0^* \oplus \partial_1^* \oplus \partial_2^*
-$$
-and
-$$
-d^* = d_0^* \oplus d_1^* \oplus d_2^*
-$$
-where $\partial_k^*(C)=\partial^*(C_{k-1})\subseteq C_k$ and $d_k^*(C^*)=d(C_{k+1}^*)\subseteq C_k^*$.
+\noindent\textbf{Adjoint operators.}  
+Since the inner products on $C$ and $C^*$ are bilinear, the adjoint operators $\partial^*:C\to C$ and $d^*:C^*\to C^*$ are defined by
+\[
+\partial^*c \cdot q = c\cdot \partial q, 
+\qquad 
+d^*f \cdot g = f\cdot dg,
+\]
+for all $c,q\in C$ and $f,g\in C^*$.
+These satisfy
+\[
+\partial^* = \partial_0^* \oplus \partial_1^* \oplus \partial_2^*, 
+\qquad 
+d^* = d_0^* \oplus d_1^* \oplus d_2^*.
+\]
 
-The boundary and coboundary operators and their duals can be used to define the so called Hodge Laplacians
-$$
-\mathcal{L} = (\partial + \partial^*)^2 = \partial \partial^* + \partial^* \partial
-: C\to C
-$$
-and
-$$
-L = (d+d^*)^2 = dd^* + d^*d : C^*\to C^*
-$$
-where the fact that $\partial^2 = (\partial^*)^2 = 0$ and $d^2=(d^*)^2=0$ is used.
-The first identity $\partial^2=0$ is implied by the alternating properties of the definition of $\partial$, from where the other identities follow.
-These Hodge Laplacians satisfy the following decomposition in direct sums
-$$
-\mathcal{L}
-=
-\mathcal{L}_0
-\oplus
-\mathcal{L}_1
-\oplus
-\mathcal{L}_2
-$$
-and
-$$
+\medskip
+
+\noindent\textbf{Hodge Laplacians.}  
+The boundary and coboundary operators define the Hodge Laplacians
+\[
+\mathcal{L} = (\partial+\partial^*)^2 = \partial\partial^* + \partial^*\partial : C\to C,
+\]
+\[
+L = (d+d^*)^2 = dd^* + d^*d : C^*\to C^*,
+\]
+where $\partial^2=(\partial^*)^2=d^2=(d^*)^2=0$.
+These decompose as
+\[
+\mathcal{L} = \mathcal{L}_0 \oplus \mathcal{L}_1 \oplus \mathcal{L}_2, 
+\qquad 
 L = L_0 \oplus L_1 \oplus L_2,
-$$
-where in turn
-$$
-\mathcal{L}_k = \mathcal{L}_k^{\downarrow} \oplus \mathcal{L}_k^{\uparrow} : C_k\to C_k,
-$$
-$$
-L_k = L_k^{\downarrow} \oplus L_k^{\uparrow} : C_k^* \to C_k^*,
-$$
-$$
-\mathcal{L}_k^{\downarrow} = \partial_k^*\partial_k : C_k \to C_{k-1} \to C_k,
-$$
-$$
-\mathcal{L}_k^{\uparrow} = \partial_{k+1}\partial_{k+1}^* : C_k \to C_{k+1} \to C_k,
-$$
-$$
-L_k^{\uparrow} = d_k^*d_k : C_k^* \to C_{k+1}^* \to C_k^*,
-$$
-and
-$$
-L_k^{\downarrow} = d_{k-1}d_{k-1}^* : C_k^* \to C_{k-1}^* \to C_k^*.
-$$
+\]
+with
+\[
+\mathcal{L}_k = \mathcal{L}_k^{\downarrow}\oplus \mathcal{L}_k^{\uparrow}, 
+\qquad 
+L_k = L_k^{\downarrow}\oplus L_k^{\uparrow},
+\]
+\[
+\mathcal{L}_k^{\downarrow} = \partial_k^*\partial_k, \quad 
+\mathcal{L}_k^{\uparrow} = \partial_{k+1}\partial_{k+1}^*, 
+\]
+\[
+L_k^{\uparrow} = d_k^*d_k, \quad 
+L_k^{\downarrow} = d_{k-1}d_{k-1}^*.
+\]
 
-Any $k$-cochain $f$ admits the so called Hodge decomposition
-$$
-f = g\oplus h\oplus s
-$$
-where the direct sum components $g\in \rng d_{k-1}$, $h\in \krn L_k$ and $s\in \rng d_k^*$ are unique and are called the gradient, the harmonic and the solenoid components, respectively.
-As a consequence, there exists non-unique cochains $w\in C_{k-1}^*$ and $u\in C_{k+1}^*$ such that $g=d_{k-1}w$ and $s=d_k^*u$, that can be found solving for
-\begin{eqnarray}
-d_kf 
-&=&
-d_k(g+h+s)
-\notag
-\\
-&=&
-d_k s
-\notag
-\\
-&=&
-d_k d_k^* u
-\notag
-\\
-&=&
-L_{k+1}^{\downarrow} u
-\notag
-\end{eqnarray}
-and
-\begin{eqnarray}
-d_{k-1}^* f 
-&=&
-d_{k-1}^*(g+h+s)
-\notag
-\\
-&=&
-d_{k-1}^* g
-\notag
-\\
-&=&
-d_{k-1}^*d_{k-1} w
-\notag
-\\
-&=&
-L_{k-1}^{\uparrow} w,
-\notag
-\end{eqnarray}
-while the harmonic component can be found from $h=f-g-s$.
+\medskip
 
-\subsection{Discrete calculus reformulation of the infererence of ratings}
+\noindent\textbf{Hodge decomposition.}  
+Every $k$-cochain $f$ admits the Hodge decomposition
+\[
+f = g \oplus h \oplus s,
+\]
+where $g\in \operatorname{im} d_{k-1}$ (gradient), $h\in \ker L_k$ (harmonic), and $s\in \operatorname{im} d_k^*$ (solenoidal).
+Equivalently, there exist $w\in C_{k-1}^*$ and $u\in C_{k+1}^*$ such that
+\[
+g = d_{k-1}w, \qquad s = d_k^* u,
+\]
+with $w$ and $u$ determined by
+\begin{align}
+\label{eq:2}
+d_k f &= L_{k+1}^{\downarrow} u, \\
+\label{eq:3}
+d_{k-1}^* f &= L_{k-1}^{\uparrow} w,
+\end{align}
+while $h=f-g-s$.
+
+\subsection{HodgeRank}
+
+Expression~\eqref{eq:1} can be reformulated in the language of discrete calculus as
+\begin{equation}
+\label{eq:4}
+\lVert f - d_0(w) \rVert,
+\end{equation}
+where the scalar components $f_{ij}$ of the $1$-cochain $f=f_{ij}e^{ij}$ encode the pairwise comparison scores, and the components $w_i$ of the $0$-cochain $w=w_ie^i$ represent the ratings.
+The minimizer of~\eqref{eq:4} coincides with the solution of Eq.~\eqref{eq:3} for $k=1$.
+
+In this setting, the set of $0$-simplices $K_0$ corresponds to the teams $V$, the set of $1$-simplices $K_1$ to the comparisons $E$, and the set of $2$-simplices $K_2$ to the closed triples of comparisons:
+\[
+K_2 = \bigl\{ \{i,j,k\} : \{i,j\},\{j,k\},\{i,k\}\in E \bigr\}.
+\]
+For $k>2$, $K_k=\emptyset$.
 
@@
 
+The dataset covers five soccer leagues: England, France, Germany, Italy and Spain.
+Each league consists of $N=20$ teams, except for Germany, which has $N=18$.
+
+in the major European soccer leagues (e.g., Premier League, La Liga, Serie A, Bundesliga, Ligue 1), the final standings are determined by a points system based on the results of matches during the season.
+In the standard system, a win adds 3 points, a draw 1 point and a loss 0 points.
+Each team plays every other team in the league twice (home and away) in what is called a round-robin format. 
+Over the course of the season, the $i$-th team accumulates a total of $\pi_i$ points.
+At the end of the season, the teams are ranked in descending order of total points.
+If two or more teams have the same number of points, tiebreakers are applied. The exact rules vary slightly by league, but usually the order is something like:
+Goal difference (GD): goals scored minus goals conceded.
+Goals scored (GF): total number of goals scored during the season.
+Head-to-head record: results of the matches between the tied teams (some leagues use this before GD).
+Playoff or extra match: in rare cases (e.g., for the title, relegation, or European qualification) if all other criteria fail.
+So the final standing (the ranking table at the end of the season) is determined by total points, with tiebreakers applied as needed.
+
 \subsection{
@@
 
-The dataset covers five soccer leagues: England, France, Germany, Italy, and Spain.
-Each league consists of $N=20$ teams, except for Germany, which has $N=18$.
-Every pair of teams $i,j$ plays two matches: one with $i$ as the home team and $j$ as the away team, and in the other match it is the other way around.
-The teams are indexed or ranked according to their position in that year’s final standings.
-The standings is characterized by some tournament points $\pi_i$ for each team $i$ in each league.
-Therefore, the condition $i<j$ holds if and only if $\pi_i>\pi_j$ for any two teams $i$ and $j$ in a given league.
-
-For each pair of teams $i$ and $j$ in a given league, several pairwise metrics $m_{ij}\in \mathbb{R}$ are provided.
-Each metric takes a positive value if it presumably favors team $i$ over team $j$, and a negative value otherwise.
+For each pair $\{i,j\}$ of teams in a given league, several pairwise metrics $m_{ij}$ and $m_{ji}$ are provided.
+Each metric $m_{ij}$ takes a positive value when it presumably favors the home team $i$ over away team $j$, and takes a negative value otherwise.
 The different types of empirically determined metrics being considered are:
@@
 \end{enumerate}
-Namely, $m_{ij}$ corresponds to $i$ being the home team and $j$ being the away team.
-Additionally, metric-styles derived from the Singular Value Decomposition of previous metrics are also considered~\cite{chacoma2025data}.
-Notice, $m_{ij}$ is generally independent of the metric $m_{ji}$ since they belong to different matches between $i$ and $j$.
-As a consequence, $m_{ij}+m_{ji}$ is generally different from zero and is the net contribution between teams $i$ and $j$.
+Moreover, the Singular Value Decomposition of previous emprically determined metrics are considered in order to define the so-called {\em styles} which are also some kind of metric~\cite{chacoma2025data}.
 
@@
 \frac{1}{2}(m_{ij}+m_{ji})
+=
+- f_{ji}
 $$
-are defined for each $i$ and $j$ in $\{0,1,...,n\}$ such that $i<j$ and where $N=n+1$ is the number of teams in the corresponding league.
-Moreover, in order to test the inference method, two types of ground-truth coefficients are also defined.
+are defined for each $i$ and $j$ such that $i<j$.
+Moreover, in order to test the HodgeRank inference method, two types of ground-truth coefficients are also defined.
 Namely, the rank-based coefficients $f_{ij}=i-j$ and the tournament points based coefficients $f_{ij}=\pi_j-\pi_i$.
-These coefficients correspond to a fully connected network or graph with $N$ nodes since, as already mentioned, all pairs of teams play two matches against each other in each league.
-Next, each fully connected graphs is promoted to a simplicial complex composed of 0-simplices labeled by $i$, 1-simplices labeled by $ij$ with $i<j$ and 2-simplices labeled by $ijk$ with $i<j<k$.
-Finally, the coefficients $f_{ij}$ are promoted to be the components of a 1-cochain  $f$, where the anti-symmetric relation $f_{ji}=-f_{ij}$ must hold by definition.
+In all cases there is an underlying fully connected network or graph with $N$ nodes, that is promoted to a simplicial complex composed of 0-simplices labeled by $i$, 1-simplices labeled by $ij$ with $i<j$ and 2-simplices labeled by $ijk$ with $i<j<k$.
+Correspondingly, the coefficients $f_{ij}$ are promoted to be the components of a 1-cochain  $f$, where the anti-symmetric relation $f_{ji}=-f_{ij}$ must hold by definition.
 

### 2025-09-08 draft/draft.tex
progress

@@
 Consider a set of $N=n+1$ teams labeled $\{0,1,\dots,n\}$.
-For each distinct pair of teams $i$ and $j$, assume that a pairwise comparison score $f_{ij}=-f_{ji}$ is available.
-Intuitively, $f_{ij}>0$ indicates that team $i$ should be rated higher than team $j$, whereas $f_{ij}<0$ suggests the opposite.
+For each distinct pair of teams $i$ and $j$, assume a pairwise comparison score $f_{ij}=-f_{ji}$ is available.
+Informally, $f_{ij}>0$ indicates that team $i$ is rated higher than team $j$, whereas $f_{ij}<0$ suggests the opposite.
 
-A natural approach to inferring ratings consistent with these comparisons is to assign scores $w_i$ (for $i=0,1,\dots,n$) that minimize the sum of squared discrepancies:
+A natural approach to inferring consistent ratings is to assign scores $w_i$ ($i=0,1,\dots,n$) that minimize the total squared discrepancy
 \begin{equation}
@@
 \end{equation}
-Once the team ratings $w_0,w_1,\dots,w_n$ have been inferred, the ranking $r_0,r_1,\dots,r_n$ is obtained by requiring that $r_i<r_j$ whenever $w_i>w_j$.
-This setup leads to a well-posed optimization problem that admits an elegant reformulation in terms of cochains over simplicial complexes.
+Once the team ratings $w_0,w_1,\dots,w_n$ have been obtained, the ranking $r_0,r_1,\dots,r_n$ follows from the ordering condition $r_i<r_j$ whenever $w_i>w_j$.
+This least-squares formulation defines a well-posed optimization problem that admits an elegant reformulation in terms of cochains over simplicial complexes.
 The resulting framework for rating and ranking inference is known as \emph{HodgeRank}~\cite{jiang2011statistical, perotti2025analysis}.
@@
 
-To review the HodgeRank method, we summarize the basic elements of discrete exterior calculus on simplicial complexes.  
-Let $V=\{0,1,\dots,n\}$ denote a set of vertices.  
-A simplicial complex is a collection
+To review the HodgeRank method, we recall the basic elements of discrete exterior calculus on simplicial complexes.  
+Let $V=\{0,1,\dots,n\}$ denote a set of vertices, and define a simplicial complex as a collection
 \[
-K = K_0 \cup K_1 \cup \dots \cup K_n
+K = K_0 \cup K_1 \cup \dots \cup K_n,
 \]
@@
 \[
-K_k \subseteq \bigl\{\{i_0,\dots,i_k\} : 0\leq i_0<\dots<i_k\leq n,\ i_j\in V\bigr\},
+K_k
+\subseteq
+\bigl\{
+\{i_0,\dots,i_k\}
+:
+0\leq i_0<\dots<i_k\leq n
+\bigr\}
 \]
-of $k$-simplices, i.e. unordered subsets $\{i_0,\dots,i_k\}$ of the vertex set satisfying the closure condition: if $s\in K$ and $r\subset s$ with $r\neq \emptyset$, then $r\in K$.
+is the set of $k$-simplices, i.e., unordered subsets $\{i_0,\dots,i_k\}$ of vertices satisfying the closure condition: if $s\in K$ and $r\subset s$ with $r\neq \emptyset$, then $r\in K$.
 
-The set $K$ of simplices is associated with a vector space $C$ of vectors called chains, whose canonical basis is denoted by $\{e_s:s\in K\}$.
-We equip $C$ with the trivial inner product
+The set $K$ of simplices is associated with a vector space $C$ of \emph{chains}, with canonical basis $\{e_s:s\in K\}$.
+We equip $C$ with the inner product
 \[
@@
 \]
-Using Einstein summation convention, any chain $c\in C$ can be written as $c=c^se_s$ with coefficients $c^s\in \mathbb{R}$.
-The chain space decomposes into
+With Einstein summation convention, any chain $c\in C$ can be written as $c=c^s e_s$ with coefficients $c^s\in \mathbb{R}$.
+The chain space decomposes as
 \[
@@
 \]
-where $C_k$ is the subspace spanned by $\{e_s:s\in K_k\}$.
+where $C_k$ is spanned by $\{e_s:s\in K_k\}$.
 
-Dually, the space of cochains $C^*$ is spanned by $\{e^s:s\in K\}$, where $e^r(e_s)=\delta^r{}_s$.
-Any $f\in C^*$ decomposes as $f=f_se^s$ with coefficients $f_s\in \mathbb{R}$, and
+The dual space $C^*$ of \emph{cochains} is spanned by $\{e^s:s\in K\}$, where $e^r(e_s)=\delta^r{}_s$.
+Any $f\in C^*$ decomposes as $f=f_s e^s$ with coefficients $f_s\in \mathbb{R}$, and
 \[
@@
 \]
-The map $f\mapsto f^\flat$ is an isomorphism, with inverse denoted by $c\mapsto c^\sharp$.
+The map $f\mapsto f^\flat$ is an isomorphism, with inverse denoted $c\mapsto c^\sharp$.
 
@@
 \noindent\textbf{Orientation and boundary.}  
-An orientation is imposed on chains by requiring
+An orientation is fixed by requiring
 \[
-e_s = e_{i_0\dots i_k} = (-1)^p e_{j_0\dots j_k},
+e_{i_0\dots i_k} = (-1)^p e_{j_0\dots j_k},
 \]
-for each simplex $s=\{i_0,\dots,i_k\}\in K$, where $0\leq i_0<\dots<i_k\leq n$ and $p$ is the parity of the permutation taking $(i_0,\dots,i_k)$ into $(j_0,\dots,j_k)$.
+for each simplex $s=\{i_0,\dots,i_k\}\in K$, where $p$ is the parity of the permutation mapping $(i_0,\dots,i_k)$ to $(j_0,\dots,j_k)$.
 
@@
 \]
-Thus $\partial(C_k)\subseteq C_{k-1}$, with $C_{-1}=\{0\}$.
-This yields the decomposition
+Thus $\partial(C_k)\subseteq C_{k-1}$, with $C_{-1}=\{0\}$, yielding the decomposition
 \[
 \partial = \partial_0 \oplus \partial_1 \oplus \partial_2,
+\qquad \partial_k(C_k)\subseteq C_{k-1}, \ \partial_0=0.
 \]
-where $\partial_k(C_k)\subseteq C_{k-1}$ and $\partial_0=0$.
 
@@
 \[
-df(c) = f(\partial c),
+df(c) = f(\partial c), \qquad f\in C^*, \ c\in C.
 \]
-for all $f\in C^*$ and $c\in C$.
-Thus $d(C_k^*) \subseteq C_{k+1}^*$, with $C_3^*=\{0\}$, and
+Hence $d(C_k^*) \subseteq C_{k+1}^*$, with $C_3^*=\{0\}$, and
 \[
@@
 \noindent\textbf{Adjoint operators.}  
-Since the inner products on $C$ and $C^*$ are bilinear, the adjoint operators $\partial^*:C\to C$ and $d^*:C^*\to C^*$ are defined by
+The adjoints $\partial^*:C\to C$ and $d^*:C^*\to C^*$ are defined by
 \[
@@
 for all $c,q\in C$ and $f,g\in C^*$.
-These satisfy
+They decompose as
 \[
@@
 
-\medskip
-
 \noindent\textbf{Hodge Laplacians.}  
 The boundary and coboundary operators define the Hodge Laplacians
-\[
-\mathcal{L} = (\partial+\partial^*)^2 = \partial\partial^* + \partial^*\partial : C\to C,
-\]
-\[
-L = (d+d^*)^2 = dd^* + d^*d : C^*\to C^*,
-\]
-where $\partial^2=(\partial^*)^2=d^2=(d^*)^2=0$.
+\begin{eqnarray}
+\mathcal{L} 
+&=&
+(\partial+\partial^*)^2 
+=
+\partial \partial^* + \partial^* \partial : C\to C,
+\mbox{and}
+\notag
+\\
+L &=& (d+d^*)^2 = dd^* + d^*d : C^*\to C^*,
+\notag
+\end{eqnarray}
+with $\partial^2=(\partial^*)^2=d^2=(d^*)^2=0$.
 These decompose as
@@
 
-\medskip
-
 \noindent\textbf{Hodge decomposition.}  
@@
 
-\subsection{HodgeRank}
+%\subsection{HodgeRank}
 
-Expression~\eqref{eq:1} can be reformulated in the language of discrete calculus as
+The least-squares problem~\eqref{eq:1} admits a reformulation in the language of discrete calculus:
 \begin{equation}
@@
 \end{equation}
-where the scalar components $f_{ij}$ of the $1$-cochain $f=f_{ij}e^{ij}$ encode the pairwise comparison scores, and the components $w_i$ of the $0$-cochain $w=w_ie^i$ represent the ratings.
+where the scalar entries $f_{ij}$ of the $1$-cochain $f=f_{ij}e^{ij}$ encode the pairwise comparison scores, and the components $w_i$ of the $0$-cochain $w=w_ie^i$ represent the team ratings.
 The minimizer of~\eqref{eq:4} coincides with the solution of Eq.~\eqref{eq:3} for $k=1$.
 
-In this setting, the set of $0$-simplices $K_0$ corresponds to the teams $V$, the set of $1$-simplices $K_1$ to the comparisons $E$, and the set of $2$-simplices $K_2$ to the closed triples of comparisons:
+In this setting, the set of $0$-simplices $K_0$ corresponds to the teams $V$, the set of $1$-simplices $K_1$ to the pairwise comparisons $E$, and the set of $2$-simplices $K_2$ to the closed triples of comparisons:
 \[
@@
 
-\section{
-Methods
+\section{Methods}
 \label{sec:methods}
-}
 
-The dataset covers five soccer leagues: England, France, Germany, Italy and Spain.
-Each league consists of $N=20$ teams, except for Germany, which has $N=18$.
+The dataset covers five major soccer leagues: England, France, Germany, Italy, and Spain.
+Each league consists of $N=20$ teams, except Germany, which has $N=18$.
 
-in the major European soccer leagues (e.g., Premier League, La Liga, Serie A, Bundesliga, Ligue 1), the final standings are determined by a points system based on the results of matches during the season.
-In the standard system, a win adds 3 points, a draw 1 point and a loss 0 points.
-Each team plays every other team in the league twice (home and away) in what is called a round-robin format. 
-Over the course of the season, the $i$-th team accumulates a total of $\pi_i$ points.
-At the end of the season, the teams are ranked in descending order of total points.
-If two or more teams have the same number of points, tiebreakers are applied. The exact rules vary slightly by league, but usually the order is something like:
-Goal difference (GD): goals scored minus goals conceded.
-Goals scored (GF): total number of goals scored during the season.
-Head-to-head record: results of the matches between the tied teams (some leagues use this before GD).
-Playoff or extra match: in rare cases (e.g., for the title, relegation, or European qualification) if all other criteria fail.
-So the final standing (the ranking table at the end of the season) is determined by total points, with tiebreakers applied as needed.
+In these competitions (e.g., Premier League, La Liga, Serie A, Bundesliga, Ligue 1), the final standings are determined by a points system based on match results throughout the season.
+In the standard system, a win yields 3 points, a draw 1 point, and a loss 0 points.
+Each team plays every other team twice (home and away) in a round-robin format. 
+Over the season, the $i$-th team accumulates $\pi_i$ total points.
+At the end of the season, teams are ranked in descending order of $\pi_i$.
+If two or more teams tie on points, tiebreakers are applied, typically in the following order (with minor variations across leagues):  
+goal difference (GD: goals scored minus goals conceded),  
+goals scored (GF),  
+head-to-head record,  
+and, in rare cases, a playoff or extra match.  
+Thus, the final table is determined primarily by total points, with tiebreakers resolving ties.
 
-\subsection{
-Data
+%\subsection{Data}
 \label{sec:data}
-}
 
-For each pair $\{i,j\}$ of teams in a given league, several pairwise metrics $m_{ij}$ and $m_{ji}$ are provided.
-Each metric $m_{ij}$ takes a positive value when it presumably favors the home team $i$ over away team $j$, and takes a negative value otherwise.
-The different types of empirically determined metrics being considered are:
+For each pair $\{i,j\}$ of teams in a given league, several pairwise metrics $m_{ij}$ and $m_{ji}$ are recorded.
+Each metric $m_{ij}$ takes a positive value when it favors team $i$ over $j$, and a negative value otherwise.
+The metrics considered include:
 \begin{enumerate}
-\item {\bf crossings:}
-\item {\bf counterattacks:}
-\item {\bf pressure loss:}
-\item {\bf build-up time:}
-\item {\bf direct play:}
-\item {\bf pressure points:}
-\item {\bf shots:}
-\item {\bf flow rate:}
-\item {\bf maintenance time:}
-\item {\bf middle-zone time:} bla bla.
+\item \textbf{crossings}
+\item \textbf{counterattacks}
+\item \textbf{pressure loss}
+\item \textbf{build-up time}
+\item \textbf{direct play}
+\item \textbf{pressure points}
+\item \textbf{shots}
+\item \textbf{flow rate}
+\item \textbf{maintenance time}
+\item \textbf{middle-zone time} (etc.)
 \end{enumerate}
-Moreover, the Singular Value Decomposition of previous emprically determined metrics are considered in order to define the so-called {\em styles} which are also some kind of metric~\cite{chacoma2025data}.
+Additionally, the Singular Value Decomposition of these empirical metrics is considered to define higher-level descriptors called \emph{styles}~\cite{chacoma2025data}.
 
-\subsection{Inference of ratings and rankings}
+\subsection{HodgeRank inference via metrics}
 
-With each metric, the coefficients
-$$
+Given a metric, the coefficients
+\[
 f_{ij}
 =
+-
 \frac{1}{2}(m_{ij}+m_{ji})
-=
-- f_{ji}
-$$
-are defined for each $i$ and $j$ such that $i<j$.
-Moreover, in order to test the HodgeRank inference method, two types of ground-truth coefficients are also defined.
-Namely, the rank-based coefficients $f_{ij}=i-j$ and the tournament points based coefficients $f_{ij}=\pi_j-\pi_i$.
-In all cases there is an underlying fully connected network or graph with $N$ nodes, that is promoted to a simplicial complex composed of 0-simplices labeled by $i$, 1-simplices labeled by $ij$ with $i<j$ and 2-simplices labeled by $ijk$ with $i<j<k$.
-Correspondingly, the coefficients $f_{ij}$ are promoted to be the components of a 1-cochain  $f$, where the anti-symmetric relation $f_{ji}=-f_{ij}$ must hold by definition.
+\]
+are defined for each $i<j$.
+The minus sign is imposed to identify ratings as potentials~\cite{perotti2025analysis}.
+To test the HodgeRank inference method, two types of ground-truth coefficients are also introduced:  
+(i) \emph{rank-based} coefficients $f_{ij}=i-j$, and  
+(ii) \emph{points-based} coefficients $f_{ij}=\pi_j-\pi_i$.
 
-%From a score $\phi$, a corresponding cochain $f\in C_1^*$ of coordinates $f_{ij}=\phi_{ji}-\phi_{ij}$ is defined.
-%Also, the cochain $f'\in C_1^*$ of coordinates $f'_{ij}=\phi_j-\phi_i$ with team strengths $\phi_i=\sum_j \phi_{ij}$ is considered.
-
-{\color{red}[JUAN: Write here the inference problem.]}
+In all cases, the underlying structure is a fully connected graph with $N$ nodes, promoted to a simplicial complex composed of $0$-simplices labeled by $i$, $1$-simplices labeled by $ij$ with $i<j$, and $2$-simplices labeled by $ijk$ with $i<j<k$.
+Correspondingly, the coefficients $f_{ij}$ form the components of a $1$-cochain $f$, where the antisymmetry $f_{ji}=-f_{ij}$ is imposed by definition.
+There are no $k$-simplices for $k>2$.
 

### 2025-10-21 PAPER/main.tex
agrego carpeta paper

\documentclass[]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[switch]{lineno}
\usepackage{amsfonts,amsmath,amssymb}
\usepackage{graphicx,dcolumn,bm,xcolor,ulem}
\usepackage{hyperref,url}

\title{Ranking fútbol}
\author{}
\date{}

\begin{document}

\maketitle

\begin{abstract}
	We ...
\end{abstract}

\section{Introduction}

\section{Data}

\subsection{Recopilación de métricas}

En este trabajo utilizamos la base de datos de evento provista por L. Pappalardo et al en \cite{pappalardo2019public}. 
En ese articulo, los autores visualizan todos los partidos de la temporada 2017-2018 de las principales ligas de fútbol europeas: La Liga (España), Premier league (Inglaterra), Seria A (Italia), Bundesliga (Alemania), Ligue 1 (Francia). Por cada partido detectan, clasifican, y localizan en tiempo y espacio todos los eventos: goles, tiros al arco, pases, saques de esquina, faltas, etc. 
%
En el sistema de referencia que utilizan para ubicar los eventos en tiempo y espacio, $t$ expresa el tiempo transcurridos desde el inicio del partido, la coordenada $x$ expresa la distancia respecto del arco del equipo creador del evento, y la coordenada $y$ la distancia respecto a la banda lateral derecha. Las unidades de las coordenadas espaciales están dadas en porcentajes del campo juego, siendo por ejemplo $x=0$, $x=50$ y $x=100$ las posiciones de la linea de meta propia, la linea de centro del campo y la linea de meta del rival, respectivamente.

En este marco, definimos intervalo de posesión de pelota (BPI) como al conjunto dado por una secuencia continua de eventos generados por un equipo. Note que cada BPI contiene información de un solo equipo.
Recopilamos todos los BPI de todos los equipos de cada liga, y sobre esos datos extrajimos algunas métricas que nos permiten detectar los recursos tácticos que están utilizando los equipos en esa ventana temporal del partido. 
Las métricas recopiladas para nuestro análisis están basadas en las propuestas por J. Fernandez-Navarro en \cite{fernandez2018influence}.
A continuación describimos en detalle a cada una de estas,

\begin{enumerate}
    \item {\it Direct play.} Cada vez que hay un pase o un tiro libre en un BPI, medimos la velocidad media en la dirección de ataque, dada por el cociente entre la distancia recorrida por la pelota en el eje $x$ y el tiempo transcurrido. De cada BPI tomamos el valor máximo. 
    Esto nos permite detectar que tan directo hacia la portería rival es el movimiento de la pelota en el equipo.

    \item {\it Counterattack.} Dado dos eventos consecutivos en un BPI, si el primero se observa en $x_1<40$ y el segundo en $x_2>60$ a una diferencia temporal $\Delta t$, se informa la velocidad como $v=\frac{x_2-x_1}{\Delta t}$. En otro caso se informa $0$.
    Esto es una medida de que tan rápido un equipo pasa de una posición defensiva en su campo a una ofensiva en campo rival.

    \item {\it Maintenance.} Dado un BPI, se calcula el promedio de las posiciones en el eje $x$, donde se generaron todos los eventos. Si se cumple $\bar{x}<40$, es decir si los eventos se desarrollaron mayoritariamente en la zona defensiva del equipo,
    se informa el tiempo total de la posesión. En otro caso se informa $0$. Esto nos permite detectar que tanto un equipo decide mantener y construir su juego desde su propio campo.

    \item {\it Build up.} Si en un BPI se verifica que $\bar{x}>60$, es decir la posesión se desarrolla mayoritariamente en campo rival, se informa el tiempo total de la posesión. En otro caso se informa $0$. Esta métrica informa el tiempo de posesión en situaciones donde el equipo invade fuertemente el campo rival.

    \item {\it Midfield play.} Si en un BPI se observa que $\bar{x}\leq60$ y $\bar{x}\geq40$, es decir la posesión de desarrolla mayoritariamente por el centro del campo de juego, se informa el tiempo total de la posesión, en otro caso se informa $0$.La idea de esta variable es medir el tiempo que el equipo pasa en el sector medio del campo de juego.

    \item {\it Flow rate.} En cada BPI donde $\bar{x}\geq50$ se toma la diferencia temporal entre todos los eventos, y se calcula el valor medio, $\bar{dt}$. Luego se define la métrica como $1/\bar{dt}$. De esta manera se tiene una medida de que tan rápido el equipo mueve la pelota en el campo rival.

    \item {\it Crossing.} Si en un BPI se observa un evento centro, se informa $1$, en otro caso se informa $0$. Esta métrica sirve para contabilizar los intentos de llegada por vía aérea.

    \item {\it Pressure point.} De cada BPI se toma el primer evento, y se extrae la posición en la variable $x$, es decir donde el equipo comienza su posición. Esto nos permite medir si el equipo esta recuperando la pelota en su campo, en la zona media o en el campo rival.

    \item {\it Pressure loss.} Si un BPI comienza en un evento donde $x>40$, se informa el tiempo total de la posesión del adversario del BPI inmediatamente anterior. Este métrica es útil para observar si el equipo esta relajando o aumentando el nivel de presión que ejerce sobre el juego del rival en las zonas medias y altas del campo de juego.

    \item {\it Shots.} Si en un BPI se registra un evento ``Shot", se informa 1, en otro caso se informa $0$. Esta métrica permite contabilizar los tiros al arco de cada equipo.

\end{enumerate}

Para nuestro análisis,se descartaron todas los BPI con menos de 3 eventos y con tiempo total menor a 2 segundos. La idea de esto es descartar pequeñas recuperaciones pasajeras y quedarnos con posesiones consolidadas.
Del proceso de recopilación se obtuvieron $215681$ BPI.
%
%Luego de calcular los valores de las métricas en cada uno de estos, estudiamos la distribución de los datos. Observamos que las métricas parecen seguir una distribución tipo log-normal, por lo tanto decidimos transformar los datos como $x \rightarrow log(1+x)$ para trabajar con distribuciones aproximadamente normales. 
%
Luego, agrupamos la información por partido y por equipo, y sumamos los valores obtenidos en cada métrica. De esta manera por ejemplo el feature {\it Shots} cuantifica la cantidad de tiros al arco ejecutados por el equipo en ese partido.
Asimismo el feature {\it Build-up} cuantifica la cantidad de tiempo neto en la cual un equipo sostuvo una posición de ataque frente al rival en ese partido.
%
Note que en la temporada 2017/2018, en las ligas Española, Inglesa, Francesa e Italiana los equipos jugaron 38 partidos. Por lo tanto, al tomar los datos de los primeros 4 equipos cada liga aporta un total de $38\times4=152$ muestras al archivo de datos. Asimismo, en la liga alemana, al ver menos equipos, se jugaron 34 partidos, por lo tanto esta liga aporta 136 muestras. 
%
En consecuencia, DS es una matriz con 744 filas y 10 columnas. 
Por ultimo, en un dataset aparte recopilamos meta data asociada a cada muestra, útil luego para realizar el análisis: a que equipo pertenece esa muestra, cual es la liga de pertenencia, y el resultado final en la tabla de posiciones.

\subsection{Representación en redes complejas}

En lo que sigue, presentamos nuestra propuesta para representar las métricas de rendimiento en términos de redes complejas.
Definimos $M(i, j, g)$ como la métrica de rendimiento correspondiente al equipo $i$ cuando enfrenta al equipo $j$ en el partido $g$. Por ejemplo, puede representar la cantidad de tiros al arco realizados por el FC Barcelona al jugar contra el Real Madrid en el primer encuentro del torneo español {\it La Liga}.
%
En nuestro conjunto de datos, todos los equipos participaron en un formato de liga todos contra todos, enfrentándose dos veces: un partido de ida ($g_1$) y otro de vuelta ($g_2$). Utilizando la información de ambos encuentros, definimos una métrica agregada que resume el desempeño observado entre esos dos equipos a lo largo del torneo:

$$
M(i,j) = \sum_{g_1, g_2} M(i,j,g).
$$
%
En el ejemplo anterior, $M(i,j)$ representa la cantidad total de tiros al arco realizados por el FC Barcelona contra el Real Madrid en ambos partidos del torneo.
%
Calculando $M(i,j)$ para cada par de equipos en una liga $L$, es posible representar estas relaciones de desempeño mediante un grafo dirigido y ponderado $G(L,M)$, cuyos pesos se definen como

$$
f_{ij} = M(j,i) - M(i,j).
$$
%
Nótese que, en esta representación, $f_{ij} < 0$ indica que el equipo $i$ superó al equipo $j$ en la métrica considerada.
%
En este marco, construimos diez grafos por liga, cada uno asociado a una métrica de rendimiento distinta.

\section{Resultados}

\subsection{Caracterización estructural de las redes}

\subsection{Calculo de rating mediante descomposición de Hodge }

\subsection{Comparación con el rating real en cada liga}

\section{Conclusion}

\begin{figure}[t!]
\centering
\includegraphics[width=1.\textwidth]{metricas.pdf}
\caption{}
\label{}
\end{figure}

\bibliographystyle{unsrt}
\bibliography{biblio}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2025-10-21 juan/draft/draft.tex
progress

--- 2025-09-08 draft/draft.tex
@@
 Consider a set of $N=n+1$ teams labeled $\{0,1,\dots,n\}$.
-For each distinct pair of teams $i$ and $j$, assume a pairwise comparison score $f_{ij}=-f_{ji}$ is available.
+For each distinct pair of teams $i$ and $j$, assume a pairwise antisymmetric comparison score $f_{ij}=-f_{ji}$ is available.
 Informally, $f_{ij}>0$ indicates that team $i$ is rated higher than team $j$, whereas $f_{ij}<0$ suggests the opposite.
@@
 
-For each pair $\{i,j\}$ of teams in a given league, several pairwise metrics $m_{ij}$ and $m_{ji}$ are recorded.
-Each metric $m_{ij}$ takes a positive value when it favors team $i$ over $j$, and a negative value otherwise.
-The metrics considered include:
+For each pair $\{i,j\}$ of teams in a given league, several pairwise antisymmetric comparison scores $f_{ij}$ are defined:
 \begin{enumerate}
-\item \textbf{crossings}
-\item \textbf{counterattacks}
-\item \textbf{pressure loss}
-\item \textbf{build-up time}
-\item \textbf{direct play}
-\item \textbf{pressure points}
-\item \textbf{shots}
-\item \textbf{flow rate}
-\item \textbf{maintenance time}
-\item \textbf{middle-zone time} (etc.)
+\item \textbf{crossings:} ...
+\item \textbf{counterattacks:} ...
+\item \textbf{pressure loss:} ...
+\item \textbf{build-up time:} ...
+\item \textbf{direct play:} ...
+\item \textbf{pressure points:} ...
+\item \textbf{shots:} ...
+\item \textbf{flow rate:} ...
+\item \textbf{maintenance time:} ...
+\item \textbf{middle-zone time:} ...
+\item \textbf{style:} the Singular Value Decomposition (SVD) of previous metrics is used to define higher-level descriptors called \emph{styles}, so $f_{ij}=...$~\cite{chacoma2025data}.
+\item \textbf{rank based ground-truth:} $f_{ij}=i-j$.
+\item \textbf{point based ground-truth:} $f_{ij}=\pi_j-\pi_i$.
 \end{enumerate}
-Additionally, the Singular Value Decomposition of these empirical metrics is considered to define higher-level descriptors called \emph{styles}~\cite{chacoma2025data}.
 

### 2025-12-12 PAPER/main.tex
agrego graficos
progress con paper

@@
-\documentclass[]{article}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage[switch]{lineno}
-\usepackage{amsfonts,amsmath,amssymb}
-\usepackage{graphicx,dcolumn,bm,xcolor,ulem}
-\usepackage{hyperref,url}
-
-\title{Ranking fútbol}
-\author{}
-\date{}
+\documentclass[preprint, amsmath,amssymb, pre]{revtex4-2}
+\usepackage{graphicx,dcolumn,bm,xcolor,lineno,ulem}
+\newcommand{\avg}[1]{\left\langle #1  \right\rangle}
+\newcommand{\agregar}[1]{{\color{blue}{#1}}}
+\newcommand{\sacar}[1]{{\color{red}\sout{#1}}} 
+%\linenumbers
 
@@
 
-\maketitle
+\title{titulo}
+
+\author{Andres Chacoma}
+\email{achacoma@df.uba.ar}
+\affiliation{Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Física. Buenos Aires, Argentina}
+\affiliation{CONICET - Universidad de Buenos Aires, Instituto de Física Interdisciplinaria y Aplicada (INFINA). Buenos Aires, Argentina.}
+
+\author{Juan I. Perotti}
+\affiliation{Universidad Nacional de Córdoba, Facultad de Matemática, Astronomía, Física y Computación. Córdoba, Argentina }
+\affiliation{CONICET- Universidad Nacional de Córdoba, Instituto de Física Enrique Gaviola (IFEG). Córdoba, Argentina.}
+
+\author{Orlando V. Billoni}
+\affiliation{Universidad Nacional de Córdoba, Facultad de Matemática, Astronomía, Física y Computación. Córdoba, Argentina }
+\affiliation{CONICET- Universidad Nacional de Córdoba, Instituto de Física Enrique Gaviola (IFEG). Córdoba, Argentina.}
 
@@
 \end{abstract}
+
+\maketitle
 
@@
 Note que cada BPI contiene información de un solo equipo.
-Recopilamos todos los BPI de todos los equipos de cada liga, y sobre esos datos extrajimos algunas métricas que nos permiten detectar los recursos tácticos que están utilizando los equipos en esa ventana temporal del partido.
-Las métricas recopiladas para nuestro análisis están basadas en las propuestas por J. Fernandez-Navarro en \cite{fernandez2018influence}.
+Recopilamos todos los BPI de todos los equipos de cada liga, y sobre esos datos extrajimos métricas que nos permiten detectar algunos de los recursos tácticos que están utilizando los equipos en esa ventana temporal del partido.
+Las métricas recopiladas para nuestro análisis están basadas en las propuestas por J.
+Fernandez-Navarro en \cite{fernandez2018influence}.
+El estudio de estas metricas fue utilizado tambien en un trabajo anterior para determinar estilos futbolisticos caracterisiticos \cite{chacoma2025data}.
 A continuación describimos en detalle a cada una de estas,
@@
 %
-%Luego de calcular los valores de las métricas en cada uno de estos, estudiamos la distribución de los datos.
+Luego de calcular los valores de las métricas en cada uno de estos, estudiamos la distribución de los datos.
 Observamos que las métricas parecen seguir una distribución tipo log-normal, por lo tanto decidimos transformar los datos como $x \rightarrow log(1+x)$ para trabajar con distribuciones aproximadamente normales.
 %
-Luego, agrupamos la información por partido y por equipo, y sumamos los valores obtenidos en cada métrica.
+Posteriormente, agrupamos la información por partido y por equipo, y sumamos los valores obtenidos en cada métrica.
 De esta manera por ejemplo el feature {\it Shots} cuantifica la cantidad de tiros al arco ejecutados por el equipo en ese partido.
@@
 %
-En consecuencia, DS es una matriz con 744 filas y 10 columnas. 
-Por ultimo, en un dataset aparte recopilamos meta data asociada a cada muestra, útil luego para realizar el análisis: a que equipo pertenece esa muestra, cual es la liga de pertenencia, y el resultado final en la tabla de posiciones.
+En consecuencia, la matriz de datos consta de 744 filas y 10 columnas. 
+Por ultimo, en un dataset aparte recopilamos meta-data asociada a cada muestra, útil luego para realizar el análisis: a que equipo pertenece esa muestra, cual es la liga de pertenencia, y el resultado final en la tabla de posiciones.
 
@@
 %
-En este marco, construimos diez grafos por liga, cada uno asociado a una métrica de rendimiento distinta.
+En este marco, construimos 50 grafos en total, diez grafos por liga. Cada uno asociado a una métrica de rendimiento distinta.
 
@@
 
-\subsection{Caracterización estructural de las redes}
+\subsection{Utilizacion de HodgeRank para la obtencion de los ratings asociados a las metricas}
 
-\subsection{Calculo de rating mediante descomposición de Hodge }
+\subsection{Comparacion entre el rating real y el rating obtenido de las metricas}
 
-\subsection{Comparación con el rating real en cada liga}
+Definimos como true ranking, $r$, de un equipo, al puesto final alcanzado en la tabla de posiciones de su liga.
+Por ejemplo, el true ranking del equipo FC Barcelona en la liga Espanola es 1, por que salio campeon esa temporada.
+Definimos tambien, true rating, $R_T$, de un equipo a la cantidad de puntos obtenidos por ese equipo en su correspondiente liga.
+Por ejemplo, el true rating del FC Barcelona en esa temporada fue 93 puntos.
+%
+Asimismo, definimos como metric rating, $R_M$, de un equipo de una liga, en una metrica dada, al valor de rating calculado a partir del metodo HodgeRank para la red asociada a esa metrica.
 
-\section{Conclusion}
+Por otro lado, como nuestra idea es comparar el true rating con el rating obtenido a partir del metodo de HodgeRank, es necesario llevar los valores de ambas variables a la misma escala.
+Por tal motivo, decidimos estandarizar las variables utilizando la media y la desviacion estandar, de la siguiente manera,
+%
+\begin{align*}
+R_T &\to \frac{R_T - \langle R_T\rangle} { \sigma(R_T)}, \\
+R_M &\to \frac{R_M - \langle R_M\rangle} { \sigma(R_M)},
+\end{align*}
+%
+donde $\langle * \rangle$ indica el valor medio y $\sigma(*)$ la desviacion estandar.
+%
+Por ultimo, definimos como $\Delta_{i,m}$ a la diferencia cuadratica de rating de una metrica $m$, de un equipo $i$, como,
+%
+\begin{equation} \label{eq:delta}
+\Delta_{i,m} = \big(  R_T(i)- R_M(i,m) \big)^2.
+\end{equation}
+%
+Esta cantidad cuantifica que tan distinto es el true rating de un equipo respecto al metric rating calculado a partir de HodgeRank para ese mismo equipo en una metrica dada.
+
+En lo siguiente dsicutimos en orma cualitativa algunas comparaciones intersantes entre true raiting y metric rating. 
+En la Fig. \ref{fi:puntos} (a) mostramos la curva $R_T$ vs. $r$ y $R_M$ vs. $r$ para el caso de la metrica {\it Presure Point} en la liga Inglesa. 
+En primer lugar podemos notar que $R_T$ decrece a medida que crece $r$, esto es trivial por defincion.
+Observamos tambien que $R_M$ tiene tambien una tendencia decreciente, pero no monotona.
+En terminos generales podemos decir que, en este caso, el metric rating se aproxima bastante bien al true rating.
+En la Fig.
+\ref{fi:puntos} (b) mostramos una comparacion analoga a la anterior para el caso de la metrica {\it Direct play} en la liga Espanola.
+En este caso podemos ver que $R_M$ parece seguir una tendencia levemente decreciente, pero difiere bastante de la curva $R_T$.
+En la Fig.
+\ref{fi:puntos} (c) mostramos lo mismo que antes pero para el caso de la metrica {\it Maintenance} en la liga Italiana.
+En este caso vemos que la curva $R_M$ no sigue una tendencia decreciente y se obverva una total descorrelacion con el true rating.
+Note, que la diferencia punto a punto entre la curvas $R_T$ y $R_M$, en cualquiera de estos casos, representa la diferencia entre el true rating y el metric rating para cada equipo de la liga. 
+En estos ejemplos, podemos ver que esta diferencia puede ser muy variada dependiendo del caso.
+Para estudiar en mas detalle estas variaciones, utilizamos la eq.
+1 para calcular la diferencia cuadratica $\Delta_{i,m}$ en todos los equipos de todas las ligas y en todas las metricas.
+La distribucion de esos valores se muestra en la Fig.
+\ref{fi:puntos} (d).
+Se observa una distribucion tipo power law con un cut off alrdedor de $\Delta_{i,m}\approx 1$, esto indica que rara vez las diferencias superan 1 desviacion estandar.
+En el rango $10^{-6} <\Delta_{i,m}<1$ ajustamos el modelo $P(\Delta_{i,m}) \propto \Delta_{i,m}^{-\alpha}$ para obtener el valor del exponente.
+Obtuvimos $\alpha= 0.61 \pm 0.02$, lo que indica una distribucion de valores extremadamente heterogeneos.
+En este caso el valor medio no es un estadisitico informativo, no obstante si lo es la mediana, la cual arrojo un valor de $0.26$.
+Podemos concluir entonces, que si bien la distribucion de diferencias cuadraticas exhibe un alto grado de complejidad los valores observados son en terminos generales relativamente pequenos.
+
+La idea a continuacion es cuantificar la diferencia entre la curva de true rating, $R_T$, y la curva metric rating, $R_M$, asociada a una metrica $m$ en una liga $l$. 
+Para eso, en primer lugar definimos $\{\Delta_{i,m,l} \}$ como al conjunto de las diferencias cuadraticas que se observan punto a punto entre las curvas $R_T$ y $R_M$ asociadas a una metrica $m$ en una liga $l$.
+A partir de esto, definimos la distancia $D(m,l)$ como la raiz cuadrada de la mediana de ese conjunto,
+%
+\begin{equation} \label{eq:D}
+D(m,l) = \sqrt {M_d ( \{\Delta_{i,m,l} \})}.
+\end{equation}
+%
+Notar, que $D(m,l)$ esta definida de manera analoga a la cantidad Root Mean Square Error (RMSE), la cual se define como el mean de las diferencias cuadraticas entre dos curvas.
+En nuestro caso decidimos tomar la mediana, debido a que la heterogeneidad de la distribucion $P(\Delta_{i,m})$ indica que el estadisitico mean podria ser poco informativo en este caso.
+
+En este marco utilizamos la eq.
+\ref{eq:D} para calcular el valor de $D$ para las 10 metricas en las 5 ligas.
+En resultado se muestra en la Fig.
+\ref{fi:ratings}, la cual esta separada en 10 paneles uno por cada metrica, indicada en el parte superior del grafico.
+En cada panel se encuentran 5 barras que indican el valor de $D$ para cada liga.
+Cada panel fue ordenado en orden ascendente, mostrando primero las ligas que muestran una menor distancia.
+En la parte inferior de cada panel mostramos el coeficiente de correlacion de pearson, $\rho$, el cual mide la correlacion entre las curvas $R_T$ y $R_M$ en cada caso.
+Los colores de las barras estan asociados a las ligas para una mejor visualizacion.
+Asimismo, las barras negras indican que en ese caso la correlacion de Person es debil, $|\rho|<0.4$.
+
+\subsection{Comparacion entre el ranking real y el ranking obtenido de las metricas}
+
+\subsection{Propuesta de un raiting multivariado}
+
+\section{Discusion y conclusion}
 
@@
 \centering
-\includegraphics[width=1.\textwidth]{metricas.pdf}
-\caption{}
-\label{}
+\includegraphics[width=1.\textwidth]{puntos.pdf}
+\caption{Comparacion entre el rating verdadero y el rating obtenido de las metricas (algunos casos ejemplares).
+(a) Caso Presure point en la liga inglesa.
+(b) Caso Direct play en la liga espanola.
+(c) Caso Maintenance en la liga italiana.
+(d) Distribucion de diferencias cuadraticas calculada sobre el conjunto de todas las metricas de todas las ligas.}
+\label{fi:puntos}
+\end{figure}
+
+\begin{figure}[t!]
+\centering
+\includegraphics[width=1.\textwidth]{rmse.pdf}
+\caption{Distancia $D$ y coeficiente de person $\rho$, entre el rating verdadero y el rating dado por las metricas.
+Cada panel contiene la informacion de una metrica y las barras muestran el valor de $D$ obtenido en cada liga.
+Las barras en cada panel estan ordenadas en orden creciente respecto del valor de $D$.
+Asimismo, los paneles estan ordenado en orden creciente respecto del valor promedio en cada metrica, $\overline{D}$.
+Las barras negras muestran los casos donde el coeficiente de Pearson indica una correlacion debil, $|\rho|<0.4$.
+}
+\label{fi:ratings}
+\end{figure}
+
+\begin{figure}[t!]
+\centering
+\includegraphics[width=1.\textwidth]{kendalls.pdf}
+\caption{Coeficiente de Kendall, $\tau$, entre los rankings verdaderos y los rankings dados por las metricas. 
+Cada panel contiene la informacion de una metrica y las barras muestran el valor de $\tau$ obtenido en cada liga. 
+Las barras en cada panel estan ordenadas en orden decreciente respecto del valor de $\tau$.
+Asimismo, los paneles estan ordenado en orden decreciente respecto del valor promedio en cada metrica, $\overline{\tau}$.
+Las barras negras muestran los casos donde el p-valor indica que no hay evidencia suficiente para afirmar que existe correlacion significativa.}
+\label{fi:rankings}
+\end{figure}
+
+\begin{figure}[t!]
+\centering
+\includegraphics[width=1.\textwidth]{relevancia.pdf}
+\caption{Relevancia de las metricas en el rating multivariado.
+Cada panel muestra los valores de $\gamma_i$ en cada metrica para una liga dada.
+Las barras negras muestran los casos donde no hay significancia estadistica.}
+\label{fi:relevancia}
 \end{figure}

### 2025-12-12 PAPER/main.tex
progress con paper

@@
 
+La idea de esta seccion es
+
 \section{Discusion y conclusion}
@@
 
+\begin{figure}[t!]
+\centering
+\includegraphics[width=1.\textwidth]{tr_mvr.pdf}
+\caption{Comparacion el true rating y el multivariate rating. 
+(a) Utilizando todas las variables para calcular el raiting multivariado.
+(b) Utilizando solo las variables relevantes de cada liga (ver Fig. \ref{fi:relevancia} ).
+%
+En ambos paneles, los scatters muestran los valores exactos para cada equipo de cada liga, las rectas fueron calculadas a partir un fit lineal con el objetivo de mostrar la tendencia que siguen las relaciones graficadas.
+La linea negra punteada muestra la funcion identidad.
+}
+\label{fi:tr_mvr}
+\end{figure}
+
 \bibliographystyle{unsrt}

### 2025-12-17 PAPER/main.tex
progress in paper

@@
 
+\subsection{Estadística del rating real}
+
+La idea de esta sección es definir un modelo estadistico para el rating real, $R_T$, el cual definimos como la cantidad de puntos totales obtenidos durante la liga.
+%
+Note que esta es una variable estocastica que depende de la cantidad de partidos ganados, empatados y perdidos.
+%
+Ademas, es sabido que en las ligas de futbol, un equipo obtiene 3 puntos cuando gana, 1 cuando empata y 0 cuanto pierde. 
+%
+En este marco, definimos las variables estocasticas $G_i$, $E_i$ y $P_i$, como la cantidad de partidos ganados, empatados y perdidos por el equipo $i$.
+Luego, el rating real de este puede escribirse como,
+
+$$
+{R_T}_i = 3G_i + 1E_i.
+$$
+
+Definiendo ${\bf p}_i = ({p_g}_i, {p_e}_i, {p_p}_i)$ como el vector que contiene las probabilidades de ganar, empatar o perder un partido, para el equipo $i$, podemos modelar la cantidad de partidos ganados, empatados y perdidos por ese equipo con una distribucion multinomial,
+
+$$
+(G_i,E_i,P_i) \sim Multinomial( n, {\bf p}_i ),
+$$
+%
+Donde $n$ es la cantidad total de partidos jugados, lo cual en una liga es igual para todos los equipos. 
+Note que cada equipo tendrá asociado un vector ${\bf p}$ diferente, dependiendo de su nivel competitivo. 
+%
+Podemos modelar este vector a partir de una distribución de Dirichlet,
+
+\begin{align*}
+{\bf p}_i &\sim Dirichlet( \boldsymbol{\alpha} ), \\
+\boldsymbol{\alpha} &= ( \alpha_g, \alpha_e, \alpha_p ).
+\end{align*}
+%
+Esta distribución, en función del conjunto de parámetros $\boldsymbol{\alpha}$, genera vectores de tres componentes, ${\bf p}_i$, que cumplen con las propiedades requeridas para una terna de probabilidades:
+
+\begin{align} \label{simplex}
+{p_g}_i>0,\; {p_e}_i>0,\; {p_p}_i>0, \\ 
+{p_g}_i+{p_e}_i+{p_p}_i=1. \nonumber
+\end{align}
+
+En este marco, la probabilidad de que un equipo durante la liga gane $G=g$ partidos y empate $E=e$, viene dada por,
+
+\begin{equation} \label{integ}
+P( G=g, E=e ) = 
+\int_S f_M( g,e ; n | {\bf p} ) f_D({\bf p},\boldsymbol{\alpha} ) d{\bf p}.
+\end{equation}
+%
+En esta expresion, las densidades de probabilidad son,
+%
+\begin{align*} 
+f_M( g,e ; n | {\bf p} ) &= 
+\frac{n!}{g!\; e!\; (n-g-e)!} p_g^g\; p_e^e\; p_p^{(n-g-e)}, \\
+%
+f_D({\bf p},\boldsymbol{\alpha} ) &= 
+\frac{1}{B(\boldsymbol{\alpha})}\;  
+p_g^{(\alpha_g -1)}\; p_e^{(\alpha_e -1)}\; p_p^{(\alpha_p -1)},
+%
+\end{align*}
+%
+la densidad multinomial condidicionada y la densidad de Dirichlet, respectivamente. Ademas, el espacio de integración, $S$, es el simplex generado por las eqs. \ref{simplex}.
+%
+La integral \ref{integ} admite solucion analitica. Resolviendo, obtenemos la probabilidad de ganar $G=g$ partidos y empatar $E=e$, en funcion de los parametros $\alpha_g$, $\alpha_e$ y $\alpha_p$,
+%
+\begin{equation} \label{proba}
+P( G=g, E=e ) = \frac{n!}{ g!\; e!\; (n-g-e)!} \; 
+\frac{ B(\alpha_g+g, \alpha_e+e, \alpha_p+n-g-e) }
+{ B(\alpha_g, \alpha_e, \alpha_p) }.
+\end{equation}
+
+A partir de la eq.
+\ref{proba}, se puede obtener la distribucion de probabilidad teorica $R_T$.
+Los grados de libertad que otorgan los parametros $\boldsymbol{\alpha}$, permiten ajustando la curva teorica a los datos.
+En la Fig. \ref{fi:stats}
+
 \subsection{Comparacion entre el rating real y el rating obtenido de las metricas}
@@
 
-La idea de esta seccion es
+Para cada liga, definimos la matriz de rating multivariado, $\bm{R_{MV}}$.
+En este objeto, las fila representan los equipos, y las columnas los metric ratings calculados con el metodo de HodgeRank.
+La dimension de esta matriz, en el caso de Espana por ejemplo, esta matriz sera $10 $
 
 \section{Discusion y conclusion}
+
+\begin{figure}[t!]
+\centering
+\includegraphics[width=1.\textwidth]{stats_rating.pdf}
+\caption{Estadística del rating real.
+(a) Comparación de la Cumuative distribution function (CDF) de los valores de rating real con lo obtenido a partir del M-D model y el modelo gaussiano.
+(b) Relacion entre los Quantile asociado a los datos y los obtenidos a partir del M-D model y el modelo gaussiano.}
+\label{fi:stats}
+\end{figure}
 

### 2025-12-17 PAPER/main.tex
progress in paper

--- 2025-12-12 PAPER/main.tex
@@
 
+\subsection{Estadística del rating real}
+
+La idea de esta sección es definir un modelo estadistico para la distribucion de probabilidad del rating real, $R_T$, definido como la cantidad de puntos totales obtenidos por un equipo durante la liga.
+%
+$R_T$ es una variable estocastica que depende de la cantidad de partidos ganados, empatados y perdidos, por los equipos de la liga.
+%
+En las ligas de futbol, un equipo obtiene 3 puntos cuando gana, 1 cuando empata y 0 cuanto pierde. 
+%
+En este marco, $R_T$ puede expresarse en terminos de las variables estocasticas $G$, $E$ y $P$, que representan la cantidad de partidos ganados, empatados y perdidos por un equipo,
+%
+$$
+R_T = 3G + 1E.
+$$
+%
+Definiendo ${\bf p} = (p_g, p_e, p_p)$ como el vector que contiene las probabilidades de ganar, empatar o perder un partido, para el equipo dado, podemos modelar la cantidad de partidos ganados, empatados y perdidos por ese equipo con una distribucion multinomial,
+
+$$
+(G,E,P) \sim Multinomial( n, {\bf p} ),
+$$
+%
+Donde $n$ es la cantidad total de partidos jugados, lo cual en una liga es igual para todos los equipos. 
+Proponemos que cada equipo tiene asociado un vector ${\bf p}$ diferente, dependiendo de su nivel competitivo. Por lo tanto ${\bf p}$ tambien puede ser considerada una variable estocastica.
+%
+Podemos modelarla a partir de una distribución de Dirichlet,
+%
+\begin{align*}
+{\bf p}_i &\sim Dirichlet( \boldsymbol{\alpha} ), \\
+\boldsymbol{\alpha} &= ( \alpha_g, \alpha_e, \alpha_p ).
+\end{align*}
+%
+Esta distribución, en función del conjunto de parámetros $\boldsymbol{\alpha}$, genera vectores de tres componentes, ${\bf p}$, que cumplen con las propiedades requeridas para una terna de probabilidades:
+%
+\begin{align} \label{simplex}
+{p_g}_i>0,\; {p_e}_i>0,\; {p_p}_i>0, \\ 
+{p_g}_i+{p_e}_i+{p_p}_i=1. \nonumber
+\end{align}
+
+Con estos elementos, podemos escribir la probabilidad de que un equipo gane $G=g$ partidos y empate $E=e$, durante la liga,
+%
+\begin{equation} \label{integ}
+P( G=g, E=e ) = 
+\int_S f_M( g,e ; n | {\bf p} ) f_D({\bf p},\boldsymbol{\alpha} ) d{\bf p}.
+\end{equation}
+%
+Las densidades de probabilidad dentro de la integral son la densidad multinomial condidicionada y la densidad de Dirichlet. Estas pueden escribirse como,
+%
+\begin{align*} 
+f_M( g,e ; n | {\bf p} ) &= 
+\frac{n!}{g!\; e!\; (n-g-e)!} p_g^g\; p_e^e\; p_p^{(n-g-e)}, \\
+%
+f_D({\bf p},\boldsymbol{\alpha} ) &= 
+\frac{1}{B(\boldsymbol{\alpha})}\;  
+p_g^{(\alpha_g -1)}\; p_e^{(\alpha_e -1)}\; p_p^{(\alpha_p -1)}.
+%
+\end{align*}
+%
+Ademas, el espacio de integración, $S$, es el simplex generado por las eqs. \ref{simplex}.
+%
+La integral \ref{integ} admite solucion analitica. Resolviendo, obtenemos la probabilidad de ganar $G=g$ partidos y empatar $E=e$, en funcion de los parametros $\alpha_g$, $\alpha_e$ y $\alpha_p$,
+%
+\begin{equation} \label{proba}
+P( G=g, E=e ) = \frac{n!}{ g!\; e!\; (n-g-e)!} \; 
+\frac{ B(\alpha_g+g, \alpha_e+e, \alpha_p+n-g-e) }
+{ B(\alpha_g, \alpha_e, \alpha_p) }.
+\end{equation}
+%
+A partir de la eq. \ref{proba}, escribimos la distribucion de probabilidad teorica para el true rating,
+%
+\begin{align*} 
+&P(R_T = r_T) = \sum_X P( G=g, E=e ) \\
+&X = \{ g,e>0, \; 3g+1e=r_T, \; g+e\ge n \}
+\end{align*}
+
+Los grados de libertad que otorgan los parametros $\boldsymbol{\alpha}$, permiten ajustar la curva teorica a los datos. 
+En la Fig.
+\ref{fi:stats} (a), mostramos la cumulative distribution function (CDF) del true rating, el ajuste realizado a partir del multinomial direchlet model (M-D model) desarrollado arriba, y de una densidad gaussiana de valor medio y desviacion estandar igual al valor medio muestral y la desviacion estandar muestral observados en los datos de true rating.
+
 \subsection{Comparacion entre el rating real y el rating obtenido de las metricas}

### 2025-12-17 PAPER/main.tex
progress in paper

@@
 En la Fig.
-\ref{fi:stats} (a), mostramos la cumulative distribution function (CDF) del true rating, el ajuste realizado a partir del multinomial direchlet model (M-D model) desarrollado arriba, y de una densidad gaussiana de valor medio y desviacion estandar igual al valor medio muestral y la desviacion estandar muestral observados en los datos de true rating.
+\ref{fi:stats} (a), mostramos la cumulative distribution function (CDF) del true rating, el ajuste realizado a partir del multinomial direchlet model (M-D model) desarrollado arriba, y de una densidad gaussiana de valor medio y desviacion estandar igual al valores muestrales $\bar{R}_T=51.31$ $(s_{R_T}=17.89)$.
 

### 2025-12-18 PAPER/main.tex
progress in paper

@@
 %\linenumbers
+\addtolength{\thinmuskip}{4mu}   % Esp entre vars y oper
+\addtolength{\medmuskip}{4mu}    % Esp alred de oper. binarios (+, -)
+\addtolength{\thickmuskip}{2mu}  % Esp alred de relaciones (=, <, >)
 
@@
 
-\subsection{Utilizacion de HodgeRank para la obtencion de los ratings asociados a las metricas}
-
 \subsection{Estadística del rating real}
@@
 %
-Podemos modelarla a partir de una distribución de Dirichlet,
+Podemos modelar este parametro a partir de una distribución de Dirichlet,
 %
@@
 \begin{align} \label{simplex}
-{p_g}_i>0,\; {p_e}_i>0,\; {p_p}_i>0, \\ 
+{p_g}_i>0, {p_e}_i>0, {p_p}_i>0, \\ 
 {p_g}_i+{p_e}_i+{p_p}_i=1. \nonumber
@@
 f_M( g,e ; n | {\bf p} ) &= 
-\frac{n!}{g!\; e!\; (n-g-e)!} p_g^g\; p_e^e\; p_p^{(n-g-e)}, \\
+\frac{n!}{g! e! (n-g-e)!}\; p_g^g \; p_e^e \; p_p^{(n-g-e)}, \\
 %
 f_D({\bf p},\boldsymbol{\alpha} ) &= 
-\frac{1}{B(\boldsymbol{\alpha})}\;  
-p_g^{(\alpha_g -1)}\; p_e^{(\alpha_e -1)}\; p_p^{(\alpha_p -1)}.
+\frac{1}{B(\boldsymbol{\alpha})}  
+p_g^{(\alpha_g -1)}\; p_e^{(\alpha_e -1)} \; p_p^{(\alpha_p -1)}.
 %
@@
 \begin{equation} \label{proba}
-P( G=g, E=e ) = \frac{n!}{ g!\; e!\; (n-g-e)!} \; 
+P( G=g, E=e ) = \frac{n!}{ g! e! (n-g-e)!}  
 \frac{ B(\alpha_g+g, \alpha_e+e, \alpha_p+n-g-e) }
@@
 &P(R_T = r_T) = \sum_X P( G=g, E=e ) \\
-&X = \{ g,e>0, \; 3g+1e=r_T, \; g+e\ge n \}
+&X = \{ g,e>0,  3g+1e=r_T,  g+e\ge n \}
 \end{align*}
@@
 Los grados de libertad que otorgan los parametros $\boldsymbol{\alpha}$, permiten ajustar la curva teorica a los datos. 
+Lo hicimos a partir de un procedimiento tipo boostraping para $1000$ muestreos con remplazo, utilizando los datos de la liga Inglesa, Francesa, Italiana y Epañola.
 En la Fig.
-\ref{fi:stats} (a), mostramos la cumulative distribution function (CDF) del true rating, el ajuste realizado a partir del multinomial direchlet model (M-D model) desarrollado arriba, y de una densidad gaussiana de valor medio y desviacion estandar igual al valores muestrales $\bar{R}_T=51.31$ $(s_{R_T}=17.89)$.
+\ref{fi:stats} (a), mostramos la cumulative distribution function (CDF) del true rating junto con el ajuste.
+El grafico incluye también una densidad gaussiana de valor medio y desviacion estandar igual a los valores muestrales $\bar{R}_T=52.45$ $(s_{R_T}=18.56)$, a modo de referencia.
+Los valores obtenidos del ajuste para los parametros son 
+$\alpha_g= 0.85\pm0.03$, $\alpha_e=2.60\pm0.07$, y $\alpha_p=0.46\pm0.05$.
+Respecto a los resultados obtenidos, en primer lugar, podemos observar que el Multinomial-Dirichlet (M-D) model ajusta bien los datos empiricos en toda la curva, notandose un destacable rendimiento en la zona de raiting bajos.
+Comaprando estos resultados con la curva gaussiana, vemos que los true rating muestran colas mas livianas y un exceso de concentración en la zona de la mediana.
+En la Fig.
+\ref{fi:stats} (b), la realción entre los quantiles resalta el buen renidmiento del modelo y las diferencias claras repecto del modelo gaussiano en la zona de las colas y de la mediana.
 
@@
 \centering
-\includegraphics[width=1.\textwidth]{stats_rating.pdf}
+\includegraphics[width=1.\textwidth]{true_rating_stats.pdf}
 \caption{Estadística del rating real.

### 2025-12-19 PAPER/main.tex
progress in paper

@@
 
-Definimos como true ranking, $r$, de un equipo, al puesto final alcanzado en la tabla de posiciones de su liga.
-Por ejemplo, el true ranking del equipo FC Barcelona en la liga Espanola es 1, por que salio campeon esa temporada.
-Definimos tambien, true rating, $R_T$, de un equipo a la cantidad de puntos obtenidos por ese equipo en su correspondiente liga.
-Por ejemplo, el true rating del FC Barcelona en esa temporada fue 93 puntos.
+Definimos como true ranking de un equipo, al puesto final alcanzado en la tabla de posiciones de su liga.
+Por ejemplo, el true ranking del equipo FC Barcelona en la liga Española es 1, por que salio campeon esa temporada.
 %
@@
 
-Por otro lado, como nuestra idea es comparar el true rating con el rating obtenido a partir del metodo de HodgeRank, es necesario llevar los valores de ambas variables a la misma escala.
-Por tal motivo, decidimos estandarizar las variables utilizando la media y la desviacion estandar, de la siguiente manera,
+Nuestra idea es comparar el true rating con el rating obtenido a partir del metodo de HodgeRank.
+Naturalemente, estas cantidades estan en escalas distintas, por lo tanto para hacer la comparación es necesario llevar los valores de ambas variables a la misma escala.
+Por tal motivo, decidimos estandarizar las variables y hacer la comparacion sobre el {\it z-score} de cada una de estas.
 %
-\begin{align*}
-R_T &\to \frac{R_T - \langle R_T\rangle} { \sigma(R_T)}, \\
-R_M &\to \frac{R_M - \langle R_M\rangle} { \sigma(R_M)},
-\end{align*}
+En lo siguiente discutimos algunas comparaciones intersantes entre true raiting y el metric rating obtenido a partir de alguna de las metricas. 
 %
-donde $\langle * \rangle$ indica el valor medio y $\sigma(*)$ la desviacion estandar.
+En la Fig.
+\ref{fi:rating} (a) comparamos las CDF del true rating asociado a los datos de todos los equipos y todas las ligas, con la del metric rating asociado a los datos de todos los equipos de todas las ligas y de las diez metricas estudiadas.
+A ambas curvas se les estrajeron los valores outliers considerados como aquellos casos que se encuentran por encima del quantile $Q_{99.7}$
 %
-Por ultimo, definimos como $\Delta_{i,m}$ a la diferencia cuadratica de rating de una metrica $m$, de un equipo $i$, como,
+Podemos observar una coincidencia notable en casi todo el rango, observando algunas diferencias en las colas: el metric rating muestra mas valores positivos mas alejados de la media, y el true rating muestra mas valores negativos levemente mas alejados de la media.
 %
-\begin{equation} \label{eq:delta}
-\Delta_{i,m} = \big(  R_T(i)- R_M(i,m) \big)^2.
-\end{equation}
+En la Fig. \ref{fi:rating} (b) comparamos las curva $R_T$ y $R_M$ en función del true rnking, para el caso de la metrica {\it Presure Point} en la liga Inglesa. 
+En primer lugar podemos notar que $R_T$ decrece a medida que crece el ranking, esto es trivial por defincion: el primer equipo tiene un true raiting mayor o igual que el segundo, el segundo tiene un true raiting mayor o igual que el tercero, y asi hasta el ultimo puesto.
+En la curva de $R_M$ observamos una tendencia tambien decreciente, pero no monotona.
+En este caso, podemos decir que en terminos generales el metric rating aproxima relativamente bien la tendencia del true rating.
 %
-Esta cantidad cuantifica que tan distinto es el true rating de un equipo respecto al metric rating calculado a partir de HodgeRank para ese mismo equipo en una metrica dada.
+En la Fig.
+\ref{fi:rating} (c) mostramos una comparacion analoga a la anterior  para el caso de la metrica {\it Maintenance} en la liga Italiana.
+En este caso vemos que la curva $R_M$ no sigue una tendencia decreciente y se obverva una total descorrelacion con el true rating.
+%
+Vemos entonces que el true rating de una dada liga puede diferir poco o mucho del metric rating dependiendo de la metrica que utilizamos para inferir el rating.
+Una pregunta que surge naturalmente es cual es la estadistica asociada a esas diferencias. 
+Para estudiar esto, definimos la cantidad $\Delta R$ como la diferencia entre el true rating de un equipo $i$ de una liga $l$ y el metric rating asociado a una metrica $m$,
+%
+$$
+\Delta R = R_T(i,l) - R_M(i,l,m) 
+$$
+%
+%
+En la Fig.
+\ref{fi:rating} (d) mostramos la CDF de las diferencias $\Delta R$ observadas en cada liga.
+En cada una de estas hemos agrupado los datos provenientes de las 10 metricas.
+A modo de referencia, el grafico incluye tambien la CDF de una densidad gaussiana de valor medio y desvio estandar igual al del conjunto formado por el total de los datos.
+Note que dependiendo de cada liga la CDF se aparta de manera distinto de la referencia gaussiana, mostrando diferentes tipo de comportamiento.
+Podemos ver que las ligas Inglesa e Italiana presentan colas mas livianas que la gaussiana, indicando una densidad de probabilidad aguda alrededor de la media.
+Por el conrario, los datos asociado a liga española muestran colas levemente pesadas, mostrando la presencia de casos mas extremos.
+Por otro lado, los datos asociados a las ligas Francesa y alemana parecen comportarse de manera normal.
+Estas observaciones parecen indicar que hay ligas donde el metric rating tiende a diferenciarse mas del true rating.
+En lo que sigue cuantificamos estas diferencias en funcion de las distintas metricas.
 
-En lo siguiente dsicutimos en orma cualitativa algunas comparaciones intersantes entre true raiting y metric rating. 
-En la Fig. \ref{fi:puntos} (a) mostramos la curva $R_T$ vs. $r$ y $R_M$ vs. $r$ para el caso de la metrica {\it Presure Point} en la liga Inglesa. 
-En primer lugar podemos notar que $R_T$ decrece a medida que crece $r$, esto es trivial por defincion.
-Observamos tambien que $R_M$ tiene tambien una tendencia decreciente, pero no monotona.
-En terminos generales podemos decir que, en este caso, el metric rating se aproxima bastante bien al true rating.
-En la Fig.
-\ref{fi:puntos} (b) mostramos una comparacion analoga a la anterior para el caso de la metrica {\it Direct play} en la liga Espanola.
-En este caso podemos ver que $R_M$ parece seguir una tendencia levemente decreciente, pero difiere bastante de la curva $R_T$.
-En la Fig.
-\ref{fi:puntos} (c) mostramos lo mismo que antes pero para el caso de la metrica {\it Maintenance} en la liga Italiana.
-En este caso vemos que la curva $R_M$ no sigue una tendencia decreciente y se obverva una total descorrelacion con el true rating.
-Note, que la diferencia punto a punto entre la curvas $R_T$ y $R_M$, en cualquiera de estos casos, representa la diferencia entre el true rating y el metric rating para cada equipo de la liga. 
-En estos ejemplos, podemos ver que esta diferencia puede ser muy variada dependiendo del caso.
-Para estudiar en mas detalle estas variaciones, utilizamos la eq.
-1 para calcular la diferencia cuadratica $\Delta_{i,m}$ en todos los equipos de todas las ligas y en todas las metricas.
-La distribucion de esos valores se muestra en la Fig.
-\ref{fi:puntos} (d).
-Se observa una distribucion tipo power law con un cut off alrdedor de $\Delta_{i,m}\approx 1$, esto indica que rara vez las diferencias superan 1 desviacion estandar.
-En el rango $10^{-6} <\Delta_{i,m}<1$ ajustamos el modelo $P(\Delta_{i,m}) \propto \Delta_{i,m}^{-\alpha}$ para obtener el valor del exponente.
-Obtuvimos $\alpha= 0.61 \pm 0.02$, lo que indica una distribucion de valores extremadamente heterogeneos.
-En este caso el valor medio no es un estadisitico informativo, no obstante si lo es la mediana, la cual arrojo un valor de $0.26$.
-Podemos concluir entonces, que si bien la distribucion de diferencias cuadraticas exhibe un alto grado de complejidad los valores observados son en terminos generales relativamente pequenos.
-
-La idea a continuacion es cuantificar la diferencia entre la curva de true rating, $R_T$, y la curva metric rating, $R_M$, asociada a una metrica $m$ en una liga $l$. 
-Para eso, en primer lugar definimos $\{\Delta_{i,m,l} \}$ como al conjunto de las diferencias cuadraticas que se observan punto a punto entre las curvas $R_T$ y $R_M$ asociadas a una metrica $m$ en una liga $l$.
-A partir de esto, definimos la distancia $D(m,l)$ como la raiz cuadrada de la mediana de ese conjunto,
-%
-\begin{equation} \label{eq:D}
-D(m,l) = \sqrt {M_d ( \{\Delta_{i,m,l} \})}.
-\end{equation}
-%
-Notar, que $D(m,l)$ esta definida de manera analoga a la cantidad Root Mean Square Error (RMSE), la cual se define como el mean de las diferencias cuadraticas entre dos curvas.
-En nuestro caso decidimos tomar la mediana, debido a que la heterogeneidad de la distribucion $P(\Delta_{i,m})$ indica que el estadisitico mean podria ser poco informativo en este caso.
-
-En este marco utilizamos la eq.
-\ref{eq:D} para calcular el valor de $D$ para las 10 metricas en las 5 ligas.
-En resultado se muestra en la Fig.
-\ref{fi:ratings}, la cual esta separada en 10 paneles uno por cada metrica, indicada en el parte superior del grafico.
-En cada panel se encuentran 5 barras que indican el valor de $D$ para cada liga.
-Cada panel fue ordenado en orden ascendente, mostrando primero las ligas que muestran una menor distancia.
-En la parte inferior de cada panel mostramos el coeficiente de correlacion de pearson, $\rho$, el cual mide la correlacion entre las curvas $R_T$ y $R_M$ en cada caso.
-Los colores de las barras estan asociados a las ligas para una mejor visualizacion.
-Asimismo, las barras negras indican que en ese caso la correlacion de Person es debil, $|\rho|<0.4$.
+Nuestro objetivo es cuantificar la diferencia entre la curva de true rating, $R_T$, y las curvas de metric rating, $R_M$, asociada a una metrica $m$ en una liga $l$.
+Para esto, definimos la distancia $D$, como  el Root Mean Square Error (RMSE), el valor medio de las diferencias cuadraticas entre dos curvas.
+Ademas, calculamos el coeficiente de Person entre ambas curvas, para estudiar el comportamiento de la tendencia.
 
@@
 \centering
-\includegraphics[width=1.\textwidth]{puntos.pdf}
-\caption{Comparacion entre el rating verdadero y el rating obtenido de las metricas (algunos casos ejemplares).
-(a) Caso Presure point en la liga inglesa.
-(b) Caso Direct play en la liga espanola.
-(c) Caso Maintenance en la liga italiana.
-(d) Distribucion de diferencias cuadraticas calculada sobre el conjunto de todas las metricas de todas las ligas.}
-\label{fi:puntos}
+\includegraphics[width=1.\textwidth]{rating.pdf}
+\caption{Comparacion entre true rating y metric rating. Debido a que estas cantidades estan en unidades distintas, para comparar se usan los datos estandarizados. 
+(a) Comparacion de las cumulative distribution functions asociadas al true ranking y al metric rating.
+(b) Comparación en el caso Presure point en la liga inglesa. 
+(c) Comparación en el caso Maintenance en la liga italiana.. 
+(d) Cumulative distribution function asociadas a las diferencias punto a punto entre el true rating y el metric raiting.
+A modeo de referencia, en lineas punteadas, se muestra la CDF de una densidad gaussiana de media y desvio igual al conjunto de todos los datos.}
+\label{fi:rating}
 \end{figure}
@@
 Asimismo, los paneles estan ordenado en orden creciente respecto del valor promedio en cada metrica, $\overline{D}$.
-Las barras negras muestran los casos donde el coeficiente de Pearson indica una correlacion debil, $|\rho|<0.4$.
+Las barras con hatch muestran los casos donde el coeficiente de Pearson indica una correlacion debil, $|\rho|<0.4$.
 }
@@
 Asimismo, los paneles estan ordenado en orden decreciente respecto del valor promedio en cada metrica, $\overline{\tau}$.
-Las barras negras muestran los casos donde el p-valor indica que no hay evidencia suficiente para afirmar que existe correlacion significativa.}
+Las barras con hatch indica los casos donde el $p-value<0.05$.
+En esos casos no hay evidencia suficiente para afirmar que existe correlacion significativa.}
 \label{fi:rankings}

### 2025-12-26 PAPER/main.tex
progress in paper

@@
 \affiliation{Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Física. Buenos Aires, Argentina}
-\affiliation{CONICET - Universidad de Buenos Aires, Instituto de Física Interdisciplinaria y Aplicada (INFINA). Buenos Aires, Argentina.}
+\affiliation{CONICET - Universidad de Buenos Aires, Instituto de Física Interdisciplinaria y Aplicada. Buenos Aires, Argentina.}
 
@@
 \affiliation{Universidad Nacional de Córdoba, Facultad de Matemática, Astronomía, Física y Computación. Córdoba, Argentina }
-\affiliation{CONICET- Universidad Nacional de Córdoba, Instituto de Física Enrique Gaviola (IFEG). Córdoba, Argentina.}
+\affiliation{CONICET- Universidad Nacional de Córdoba, Instituto de Física Enrique Gaviola. Córdoba, Argentina.}
 
@@
 
+\section{Theory}
+
 \section{Resultados}
@@
 %
-En este marco, $R_T$ puede expresarse en terminos de las variables estocasticas $G$, $E$ y $P$, que representan la cantidad de partidos ganados, empatados y perdidos por un equipo,
+En este marco, $R_T$ puede expresarse en terminos de las variables estocasticas $W$ y $D$ , que representan el numero de partidos ganados y empatados, respectivamente,
 %
 $$
-R_T = 3G + 1E.
+R_T = 3W + D.
 $$
 %
-Definiendo ${\bf p} = (p_g, p_e, p_p)$ como el vector que contiene las probabilidades de ganar, empatar o perder un partido, para el equipo dado, podemos modelar la cantidad de partidos ganados, empatados y perdidos por ese equipo con una distribucion multinomial,
+Sea $n$ el número total de partidos en la temporada. Modelamos $W$ con una distribución binomial,
+%
+$$
+W \sim Bin( n |p_w )
+$$
+%
+donde $p_w$ es la probabilidad de ganar un partido. Condicional al número de victorias $W=w$, los empates siguen,
+%
+$$
+D|W=w \sim Bin( n-w | \frac{p_d}{1-p_w}  ),
+$$
+%
+con $p_d$ probabilidad de empatar.
+%
+Dado que los equipos tienen distinto nivel competitivo, cada equipo posee probabilidades intrínsecas $p_w$ y $p_d$ de ganar y empatar, respectivamente. 
+%
+Modelamos estas probabilidades mediante un enfoque jerárquico: cada equipo tiene un par de habilidades latentes $\boldsymbol{\eta} = (\eta_w, \eta_d)^T$, de ganar o empatar, que siguen una distribución normal bivariada,
 
 $$
-(G,E,P) \sim Multinomial( n, {\bf p} ),
+\boldsymbol{\eta} \sim \mathcal{N}\left(\boldsymbol{\mu},\boldsymbol{\Sigma} \right),
 $$
 %
-Donde $n$ es la cantidad total de partidos jugados, lo cual en una liga es igual para todos los equipos. 
-Proponemos que cada equipo tiene asociado un vector ${\bf p}$ diferente, dependiendo de su nivel competitivo. Por lo tanto ${\bf p}$ tambien puede ser considerada una variable estocastica.
+$$
+\boldsymbol{\mu} = ( \mu_w, \mu_d)^T,
+\quad
+\boldsymbol{\Sigma} = \begin{pmatrix} \sigma_w^2 & \rho\sigma_w\sigma_d \\ \rho\sigma_w\sigma_d & \sigma_d^2 \end{pmatrix},
+$$
 %
-Podemos modelar este parametro a partir de una distribución de Dirichlet,
+donde $\boldsymbol{\mu}$ es el centro de la distribucion, $\boldsymbol{\Sigma}$ la matriz de covarianza. En este marco, $\rho$ representa la correlación entre las habilidades para victoria y empate. 
+Las probabilidades se obtienen mediante la transformación softmax,
+%
+$$
+p_w = \frac{e^{\eta_w}}{1 + e^{\eta_w} + e^{\eta_d}}, \quad
+p_d = \frac{e^{\eta_d}}{1 + e^{\eta_w} + e^{\eta_d}}.
+$$
+%
+Esta transformación garantiza que $p_w + p_d \leq 1$ y que el modelo sea identificable al fijar implícitamente $\eta_l = 0$ para las derrotas.
+%
+Con estos elementos, podemos escribir la distribucion conjunta de victorias y empates marginalizando sobre los efectos aleatorios $\boldsymbol{\eta}$,
+%
+\begin{equation} \label{integ}
+P( W=w, D=d ) = 
+\int_{\mathbb{R}^2}  
+Bin( n |p_w(\boldsymbol{\eta}) )\;
+Bin( n-w | \frac{p_d(\boldsymbol{\eta})}{1-p_w(\boldsymbol{\eta})})\;
+\phi( \boldsymbol{\eta}| \boldsymbol{\mu}, \boldsymbol{\Sigma} )
+d \boldsymbol{\eta},
+\end{equation}
+%
+Notar que la integral en \ref{integ} no tiene forma cerrada, para resolverla utilizamos el metodo de monte carlo.
+%
+Con estos elementos, podemos finalmente escribir la distribucion de probabilidad teorica para el true rating como,
+%
+\begin{equation} \label{theo}
+P(R_T = r_T) = \sum_X P( W=w, D=d ), \\
+\end{equation}
+%
+$$
+X = \{ w,d>0,  3w+d=r_T,  w+d \le n \}.
+$$
+%
+Los grados de libertad que otorgan los parametros $\mu_w$, $\mu_d$, $\sigma_w$, $\sigma_d$ y $\rho$, permiten ajustar la curva teorica \ref{theo} a la curva empirica. 
+%
+El ajuste fue realizado sobre el conjunto de datos asociado a las ligas Inglesa, Francesa,  Italiana y Española.
+Para este analisis, decidimos no utilizar los datos de la liga Alemana, ya que su torneo se disputa entre un numero menor de equipos (18), y por lo tanto se genera una distribucion de valores de true rating menor respecto de las otras ligas, en las cuales participa un total de 20 equipos por liga.
+%
+Para hacer el ajuste, utilizamos el algoritmo Nelder-Mead minimizando el RMSE entre las CDF empírica y teórica.
+%
+Con los parámetros óptimos, realizamos bootstrap paramétrico (1000 réplicas) para estimar la incertidumbre.
+%
+Para la probabilidad de victoria obtuvimos $\bar{p}_w=0.366$ ($SD=0.010$) con un intervalo de confianza del $95\%$ de $[0.3527, 0.3798]$.
+Asimismo, para la probabilidad de empate obtuvimos $\bar{p}_d=0.280$ ($SD=0.020$) con un intervalo de confianza del $95\%$ de $[0.2548, 0.3074]$.
+%
+Para contrastar estos resultados con los datos empiricos, utilizando la informacion de los resultados de los partidos, calculamos para cada equipo $i$ las probabilidades de victoria y empate empiricas, $q_w$ y $q_d$,
 %
 \begin{align*} 
-{\bf p}_i &\sim Dirichlet( \boldsymbol{\alpha} ), \\
-\boldsymbol{\alpha} &= ( \alpha_g, \alpha_e, \alpha_p ).
+q_w^{(i)} = \frac{\# wins^{(i)}}{\#matchs^{(i)}},\quad
+q_d^{(i)} = \frac{\# draws^{(i)}}{\#matchs^{(i)}}.
 \end{align*} 
 %
-Esta distribución, en función del conjunto de parámetros $\boldsymbol{\alpha}$, genera vectores de tres componentes, ${\bf p}$, que cumplen con las propiedades requeridas para una terna de probabilidades:
+Calculando el valor medio y el desvio estandar sobre el conjunto de todos los equipos, obtuvimos para la probabilidad de victoria $\bar{q}_w=0.379$ ($SD=0.168$) con un intervalo de confianza del $95\%$ de $[0.1572, 0.7638]$, y para la probabilidad de empate $\bar{q}_d=0.243$ ($SD=0.081$) con un intervalo de confianza del $95\%$ de $[0.1046, 0.3954]$.
 %
-\begin{align} \label{simplex}
-{p_g}_i>0, {p_e}_i>0, {p_p}_i>0, \\ 
-{p_g}_i+{p_e}_i+{p_p}_i=1. \nonumber
-\end{align}
+Podemos observar, (1) Las medias son similares ($0.366$ vs $0.379$ para victorias, $0.280$ vs $0.243$ para empates), lo que valida el centro de la distribución.
+(2) las desviaciones estándar empíricas $(0.168, 0.081)$ son mucho mayores que las implícitas en el modelo $(0.01, 0.02)$, indicando que el modelo subestima la heterogeneidad entre equipos.
 
-Con estos elementos, podemos escribir la probabilidad de que un equipo gane $G=g$ partidos y empate $E=e$, durante la liga,
+Con respecto al parametro de correlacion, obtuvimos $\bar{\rho}=-0.130$ ($SD=0.07$) con un intervalo de confianza del $95\%$ de $[-0.3040, -0.0371]$.
+Esto indica la presencia de una correlacion estructural negativa debil entre las habilidades de victoria, $\eta_w$, y empate, $\eta_d$, de los equipos.
+Esto sugiere que equipos con mayor habilidad para ganar tienden ligeramente a tener menor habilidad para empatar.
+
+En la Fig.
+\ref{fi:stats} (a), mostramos la CDF del true rating junto con el ajuste.
+Podemos ver que el modelo propuesto captura muy bien el comportamiento de la curva en todo el soporte.
+En Fig.
+\ref{fi:stats} (b) mostramos un grafico con la relacion entre los quantiles teoricos y de los datos.
+A modo de referencia, mostramos tambien la relacion entre los quantiles de los datos y los asociados a una distribucion gausiana con media y desvio estandar igual a los valores muestrales obtenidos de los datos, $\bar{R}_T = 52.45$ ($SD=18.56$).
+En primer lugar, podemos ver que el modelo propuesto captura bien el comportamiento en la zona central izquierda y de manera aceptable el comportamiento de la cola de la derecha. 
+La comparación con el modelo gaussiano nos permite ver que la distribucion real presenta colas mas livianas a la izquierda y mas pesadas a la derecha, respecto del comportamiento gaussiano, develando así la presencia de una asimetria en la distribuicion.
+La concabidad hacia abajo en el centro del grafico, indica tambien que la mediana de los datos esta corrida hacia la izquierda.
+
 %
-\begin{equation} \label{integ}
-P( G=g, E=e ) = 
-\int_S f_M( g,e ; n | {\bf p} ) f_D({\bf p},\boldsymbol{\alpha} ) d{\bf p}.
-\end{equation}
 %
-Las densidades de probabilidad dentro de la integral son la densidad multinomial condidicionada y la densidad de Dirichlet. Estas pueden escribirse como,
-%
-\begin{align*} 
-f_M( g,e ; n | {\bf p} ) &= 
-\frac{n!}{g! e! (n-g-e)!}\; p_g^g \; p_e^e \; p_p^{(n-g-e)}, \\
-%
-f_D({\bf p},\boldsymbol{\alpha} ) &= 
-\frac{1}{B(\boldsymbol{\alpha})}  
-p_g^{(\alpha_g -1)}\; p_e^{(\alpha_e -1)} \; p_p^{(\alpha_p -1)}.
-%
-\end{align*}
-%
-Ademas, el espacio de integración, $S$, es el simplex generado por las eqs. \ref{simplex}.
-%
-La integral \ref{integ} admite solucion analitica. Resolviendo, obtenemos la probabilidad de ganar $G=g$ partidos y empatar $E=e$, en funcion de los parametros $\alpha_g$, $\alpha_e$ y $\alpha_p$,
-%
-\begin{equation} \label{proba}
-P( G=g, E=e ) = \frac{n!}{ g! e! (n-g-e)!}  
-\frac{ B(\alpha_g+g, \alpha_e+e, \alpha_p+n-g-e) }
-{ B(\alpha_g, \alpha_e, \alpha_p) }.
-\end{equation}
-%
-A partir de la eq. \ref{proba}, escribimos la distribucion de probabilidad teorica para el true rating,
-%
-\begin{align*} 
-&P(R_T = r_T) = \sum_X P( G=g, E=e ) \\
-&X = \{ g,e>0,  3g+1e=r_T,  g+e\ge n \}
-\end{align*}
-
-Los grados de libertad que otorgan los parametros $\boldsymbol{\alpha}$, permiten ajustar la curva teorica a los datos. 
-Lo hicimos a partir de un procedimiento tipo boostraping para $1000$ muestreos con remplazo, utilizando los datos de la liga Inglesa, Francesa, Italiana y Epañola.
-En la Fig.
-\ref{fi:stats} (a), mostramos la cumulative distribution function (CDF) del true rating junto con el ajuste.
-El grafico incluye también una densidad gaussiana de valor medio y desviacion estandar igual a los valores muestrales $\bar{R}_T=52.45$ $(s_{R_T}=18.56)$, a modo de referencia.
-Los valores obtenidos del ajuste para los parametros son 
-$\alpha_g= 0.85\pm0.03$, $\alpha_e=2.60\pm0.07$, y $\alpha_p=0.46\pm0.05$.
-Respecto a los resultados obtenidos, en primer lugar, podemos observar que el Multinomial-Dirichlet (M-D) model ajusta bien los datos empiricos en toda la curva, notandose un destacable rendimiento en la zona de raiting bajos.
-Comaprando estos resultados con la curva gaussiana, vemos que los true rating muestran colas mas livianas y un exceso de concentración en la zona de la mediana.
-En la Fig.
-\ref{fi:stats} (b), la realción entre los quantiles resalta el buen renidmiento del modelo y las diferencias claras repecto del modelo gaussiano en la zona de las colas y de la mediana.
 
@@
 Nuestra idea es comparar el true rating con el rating obtenido a partir del metodo de HodgeRank.
-Naturalemente, estas cantidades estan en escalas distintas, por lo tanto para hacer la comparación es necesario llevar los valores de ambas variables a la misma escala.
+Naturalmente, estas cantidades estan en escalas distintas, por lo tanto para hacer la comparación es necesario llevar los valores de ambas variables a la misma escala.
 Por tal motivo, decidimos estandarizar las variables y hacer la comparacion sobre el {\it z-score} de cada una de estas.
@@
 %
+Note que en las ligas Inglesa, Francesa, Italiana y Española, liga obtuvimos $20 \times 10$ muestras de $\Delta R$, ya que tenemos $20$ equipos y $10$ metricas distintas.
+No obstante, en la liga Alemana obtuvimos $18 \times 10$ muestras.
 %
-En la Fig.
-\ref{fi:rating} (d) mostramos la CDF de las diferencias $\Delta R$ observadas en cada liga.
-En cada una de estas hemos agrupado los datos provenientes de las 10 metricas.
-A modo de referencia, el grafico incluye tambien la CDF de una densidad gaussiana de valor medio y desvio estandar igual al del conjunto formado por el total de los datos.
-Note que dependiendo de cada liga la CDF se aparta de manera distinto de la referencia gaussiana, mostrando diferentes tipo de comportamiento.
-Podemos ver que las ligas Inglesa e Italiana presentan colas mas livianas que la gaussiana, indicando una densidad de probabilidad aguda alrededor de la media.
-Por el conrario, los datos asociado a liga española muestran colas levemente pesadas, mostrando la presencia de casos mas extremos.
-Por otro lado, los datos asociados a las ligas Francesa y alemana parecen comportarse de manera normal.
-Estas observaciones parecen indicar que hay ligas donde el metric rating tiende a diferenciarse mas del true rating.
-En lo que sigue cuantificamos estas diferencias en funcion de las distintas metricas.
+Nos interesa veriricar si las distribuciones de esas diferencias tienen comportamiento normal y si difieren entre una liga y otra.
+Para esto, utilizamos el grafico quantil quantile que mostramos en la Fig.
+\ref{fi:rating} (d).
+Para este grafico, tomamos los valores de $\Delta R$ de cada liga y los estandarizados al valor medio y la desviacion estandar.
+Luego calculamos los quantiles y los comparamos contra los quantiles de una distribucion normal.
+En general se observa un comportamiento bastante cercano al normal.
+La liga inglesa, podemos ver que presenta colas ligeramente pesadas tanto a izquierda como a derecha.
+Por el contrario, en la liga alemana podemos ver una asimetria, mostrando colas levente livianas a la iquierda y levemente pesadas a la derecha.
 
-Nuestro objetivo es cuantificar la diferencia entre la curva de true rating, $R_T$, y las curvas de metric rating, $R_M$, asociada a una metrica $m$ en una liga $l$.
-Para esto, definimos la distancia $D$, como  el Root Mean Square Error (RMSE), el valor medio de las diferencias cuadraticas entre dos curvas.
+Nuestro objetivo ahora es cuantificar la diferencia entre la curva de true rating, $R_T$, y las curvas de metric rating, $R_M$, asociadas a cada metrica de cada liga.
+Nuestra idea es verificar de metrica se obtiene el rating mas similar al real, y si eso depende de la liga.
+%
+Para esto, definimos la distancia $D$, como el valor medio de las diferencias cuadraticas entre dos curvas (RMSE), esto cuantifica que tan diference es $R_T$ de $R_M$ en cada caso. 
 Ademas, calculamos el coeficiente de Person entre ambas curvas, para estudiar el comportamiento de la tendencia.
@@
 \caption{Estadística del rating real.
-(a) Comparación de la Cumuative distribution function (CDF) de los valores de rating real con lo obtenido a partir del M-D model y el modelo gaussiano.
-(b) Relacion entre los Quantile asociado a los datos y los obtenidos a partir del M-D model y el modelo gaussiano.}
+(a) Comparación de la Cumuative distribution function (CDF) de los valores de rating real con lo obtenido a partir del modelo.
+(b) Relacion entre los Quantile asociado a los datos y los obtenidos a partir del modelo.
+A modo de referencia, se agrega tambien una comparación con una distribucion gaussiana de media y desvio estandar igual al de los datos.}
 \label{fi:stats}
